### 2016-08-10 robust-software.tex

\documentclass[10pt]{article}
\bibliographystyle{plos2009}
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

\date{}

\pagestyle{myheadings}

\newcommand{\fixme}[2]{\textsc{\textbf{{#1}: {#2}}}}
\newcommand{\recommend}[1]{\textit{#1}}
\newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}

\begin{document}

\begin{flushleft}
{\Large
\textbf{Ten Simple Rules for Making Research Software More Robust}
}

{Morgan~Taschuk}$^{1,\ast}$,
{Greg~Wilson}$^{2}$
\\
1) Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
\\
2) Software Carpentry Foundation / gvwilson@software-carpentry.org
\\
$\ast$ Corresponding author.
\end{flushleft}

\begin{quote}
\begin{center}\textbf{Abstract}\end{center}
\fixme{GW}{Write abstract}
\end{quote}

\section*{Introduction}

As a relatively young field, bioinformatics is full of newly developed software.
\fixme{MT}{examples here}
Efforts such as
\withurl{the ELIXIR tools and data registry}{http://dx.doi.org/10.1093/nar/gkv1116}
and the
\withurl{Bioinformatics Links Directory}{https://bioinformatics.ca/links\_directory/}
\cite{brazas2012}
have made efforts towards cataloguing it: as of May 2016, the former
has 2500 entries and the latter 1700, and those numbers are constantly
growing as both trainees and experienced practitioners produce new
software to support their research.

Typically, this software is initially developed by one person, and may
produce excellent results in their hands. But what happens after that
student leaves that lab and someone else wants to use it? Everyone with
a few years of experience feels a tremor of fear when told, ``Use
{\textless}graduated student{\textgreater}'s code to analyze your
data''. Often, that software will be undocumented and work in unexpected
ways (if it works at all without substantial modification). Equally
often, the potential new user winds up shaking their fist and cursing
the author's name. She then has two choices: hack the existing code to
make it work for her, or start over.

The root cause of this problem is that most of the software researchers
produce isn't \emph{robust}. The difference between running and being
robust is the difference between ``works for me on my machine'' and
``works for other people on a cluster I've never used''. In particular,
robust software:

\begin{itemize}
\item
  Is kept under version control.
\item
  Can be installed on systems outside the original institution
\item
  Works for users other than the original author
\item
  Has well-defined input and output formats
\item
  Has documentation that describes what its dependencies are, how to
  install it, and what the options are.
\item
  Comes with enough tests to show that it actually runs.
\end{itemize}

These are all necessary steps toward creating a reusable library that
can be shared through a site like CPAN or CRAN, and apply to both
closed-source and open-source software. They do not depend on specific
languages, libraries, packages, documentation styles, or operating
systems. Whether the aim is as simple as sharing the code with
collaborators or as complex as using the software in a production
analysis environment, increasing the robustness of your software
decreases headaches all around.

Note: we do not recommend that these rules be applied to \emph{every}
coding effort. The vast majority of code produced in the marathon of a
graduate thesis is ``throw-away'' code that is used once to answer a
specific question related to a specific dataset. However, once that
little script is dragged out three or four times for slightly different
purposes, it may be time to apply ten simple rules for robust software.
As the saying goes, not everything worth doing is worth doing right
away.

\section{Have a README that explains in a few lines what the software does and what its dependencies are.}

The README is the first stop for any potential users interested in your
software. At a minimum, it needs to provide or point to everything a new
user needs to get started, where they can turn to for help, and which
licenses apply to the software package. Exhaustive details regarding
parameters and usage are not usually necessary in a README if they are
present in usage (\#2), although a working example using test data (per
\#9) is always appreciated.

\textbf{Explain what the software does:} At the beginning of the README,
explain what the software does in one or two sentences. The description
does not need to be long or detailed. There's nothing more frustrating
than spending the time to download and install some software only to
find out that it doesn't do what you thought it did.

\begin{verbatim}
Debarcer (De-Barcoding and Error Correction) is a package for
working with next-gen sequencing data that contains molecular
barcodes.
\end{verbatim}

\textbf{List required dependencies:} Often, software depends on very
specific versions of libraries, modules, or operating systems. This is
entirely reasonable as long as it is properly documented. Often,
multiple libraries exist with the same or very similar names, so
either provide the commands necessary to download the dependencies or
link to the software homepage. Include the version number for each
dependency.  Especially if you use older versions, include links where
the packages can be downloaded. Package managers like apt, pip and
homebrew stop offering older packages after a few years.
\fixme{MT}{Is this true of pip and Homebrew?}

If your dependency is to an internal package that is not available on
the internet, you have several options depending on the sensitivity of
the code in question. If it is plain text, you can add it directly to
you repository with appropriate attribution. If the dependency is a
binary, we recommend using a
\withurl{binary repository manager}{https://en.wikipedia.org/wiki/Binary\_repository\_manager}
such as Artifactory or ProGet. These managers keep versioned copies of
software at constant URLs so they can be downloaded as long as the
manager continues to run. As a last resort, you can place it at an
internal location on shared disk, remove all write permissions, and
link to it from your README, although this method is heavily
discouraged because of the potential for the directory to go missing
because of factors outside the developer's control.

\textbf{Installation instructions:} If the software needs to be compiled
or installed, list those instructions in the readme. New users may not
be familiar with your build system, even if it is \texttt{make}. Also
mention here if you recommend they use a pre-compiled binary instead
through a system such as pip or apt.

\textbf{Input and output files}: All possible input and output files
should be listed in this section. Do the files conform to a particular
industry standard, an extension of an existing format, or is it your own
format? If using a standard format, link to the specification and
version. If you extend the standard or have your own format, define it
here explicitly, listing all the required fields and acceptable values.
(You get bonus points if you include a script to convert between
standard format and your file format). If there is no rigorous format
(such as with log files), show an example file, or the first few lines,
and explain what the sections mean.

Input files and their formats are included in most documentation.
However, the definitions of the output files are often missing. In
addition to the expected output, software will often produce
intermediate files, auxilliary files, and log files. We believe
\emph{all} output files should be listed in the README. Log and
auxilliary files are often full of valuable information that can be
mined for the user's specific purpose. Even if the files are considered
self-explanatory. Sometimes your users will need to answer a question of
the format, ``Does X tell you the percentage of reads trimmed to remove
adapter sequences?'' and you can check the documentation and confidently
say ``yes, it is in the log file''.

\textbf{Attributions and licensing:} Attributions are how you credit
your main contributors; licenses are how you want others to use and
credit your software. Both are important in your README. Leave no
question in anyone's mind about whether your software can be used
commercially, how much modification is permitted, and how other software
needs to attribute to you. If your software is not open source, include
a statement here. Attributions can also contain a list of `expert' users
that can be contacted if new users have problems with the software. (for
better or for worse)

\begin{itemize}
\item
  This is readable \emph{before} the software is installed (or even
  downloaded).
\item
  Should also include (or better yet, point to) the license for the
  software, so that (potential) users will know what they're allowed to
  do.
\item
  \fixme{GW}{example from real software:\\ https://github.com/dib-lab/khmer/blob/master/README.rst}
\end{itemize}

\section{Print usage information when launching from the command line that explains the software's features.}

Users who run your software after installation may not have access your
well-crafted README (or may not have bothered to read it). Usage
information provides their first line of help.

Ideally, usage is a terse, informative command-line help message that
guides the user in the correct use of your software. Terseness is
important: usage that extends for multiple screens, especially when
printed to standard error instead of standard output (where it can
easily be paged), is a nuisance, and is as unlikely to be read as the
README file.

Usage should provide all of the information necessary to run the
software. It is usually invoked either by running the software without
any arguments; running the software with incorrect arguments; or by
explicitly choosing a help or usage option.

An example of good usage is the Unix \texttt{mkdir} command, which makes
new directories:

\begin{small}
\begin{verbatim}
$ mkdir --help
Usage: mkdir [OPTION]... DIRECTORY...
Create the DIRECTORY(ies), if they do not already exist.

Mandatory arguments to long options are mandatory for short options too.
  -m, --mode=MODE   set file mode (as in chmod), not a=rwx - umask
  -p, --parents     no error if existing, make parent directories as needed
  -v, --verbose     print a message for each created directory
  -Z, --context=CTX  set the SELinux security context of each created
                      directory to CTX
      --help     display this help and exit
      --version  output version information and exit

Report mkdir bugs to bug-coreutils@gnu.org
GNU coreutils home page: <http://www.gnu.org/software/coreutils/>
General help using GNU software: <http://www.gnu.org/gethelp/>
For complete documentation, run: info coreutils 'mkdir invocation'
\end{verbatim}
\end{small}

There is no standard format for usage statements, but good ones share
several features:

\textbf{The syntax for running the program}: This defines the relative
location of optional and required flags and arguments for execution, and
includes the name of the program. Arguments in {[}square brackets{]}
tend to be optional. Multiple periods (e.g. ``{[}OPTION{]}\ldots{}'')
indicate that more than one can be provided.

\textbf{A text description of its purpose}: Similar to the README, the
description reminds users of the software's primary function.

\textbf{Most commonly used flags, a description of each flag, and the
default value}: Not all flags need to appear in the usage, but the most
commonly used ones should be listed here. Users will rely on this for
quick reference when working with your software.

\textbf{Where to find more information}: Whether an email address, web
site or manual page, there should be an indication where the user can go
to find more information about the software.

\textbf{Printed to standard output} : So that it can be piped into
\texttt{less}, searched with \texttt{grep}, or compared to the previous
version's help with \texttt{diff}.

\textbf{Exit with an appropriate exit code}: When usage is invoked by
providing incorrect parameters, the program should exit with a non-zero
code to indicate an error. However, when help is explicitly requested,
the software should not exit with an error, because requesting help is
sometimes used to verify that a dependency is available.

\section{Do not require root or other special privileges.}

Root (also known as ``superuser'' or ``admin'') is a special account on
a computer that has (among other things) the power to modify or delete
system files and user accounts. Conversely, files and directories owned
by root usually \emph{cannot} be modifed by normal users.

Installing or running a program with root privileges is often
convenient, since doing so automatically bypasses all those pesky safety
checks that might otherwise get in the user's way. However, those checks
are there for a reason: scientific software packages may not
intentionally be malware, but one small bug or over-eager file-matching
expression can certainly make them behave as if they were. Outside of
very unusual circumstances, packages should therefore not require root
privileges to set up or use.

Another reason for this rule is that users may want to try out a new
package before installing it system-wide on a cluster. Requiring root
privileges will frustrate such efforts, and thereby reduce uptake of the
package. Requiring that software be installed under its own user account
(e.g., that \texttt{packagename} be made a user, and all of the
package's software be installed in that ``user's'' space) is similarly
limiting, and makes side-by-side installation of multiple versions of
the package more difficult.

Developers should therefore allow packages to be installed in an
arbitrary location, e.g., under a user's home directory in
\texttt{\textasciitilde{}/packagename}, or in directories with standard
names like \texttt{bin}, \texttt{lib}, and \texttt{man} under a chosen
directory. If the first option is chosen, the user may need to modify
her search path to include the package's executables and libraries, but
this can (more or less) be automated, and is much less risky than
setting things up as root.

\section{Allow configuration of all useful parameters from the command line.}

You know what they say about assumptions*. As soon as you make one
assumption about how a software should work, a use case will come along
that will require you to change the assumption.

Every useful parameter should be configurable on the command line.
Useful parameters are those that a user will need to modify to suit
their computer, dataset or application. Providing parameters on the
command line increases the flexibility and usability of the program. You
may have determined early on that 0.58 is an optimal seed for your
original dataset, but that doesn't mean that is the best seed for every
case. Being able to change parameters on the fly to determine if and how
they change the results is important as your software gains more users,
facilitating exploratory analysis and parameter sweeping. Keep in mind
that if a parameter can be adjusted, users will want to be able to turn
off the feature entirely to have a baseline comparison.

The list of useful parameters is software-specific and so cannot be
provided here, but here is a short list of common useful parameters.

\begin{itemize}
\item
  Input and reference files/directories
\item
  Output files/directories
\item
  Filtering
\item
  Tuning (e.g.~alphas and gammas)
\item
  Seeds
\item
  Any alternatives that you've built-in, e.g.~compress results, use a
  different calculation, verbose output, etc.
\end{itemize}

When the software starts, it should echo all parameters and software
versions to standard out or a log file alongside the results. This
feature supports greater reproducibility because any result can be
replicated with only the previous output files as reference.

You can set reasonable default values to reduce the length of the
execution command as long as any parameters given on the command line
override those values. Configuration files should be used in preference
to hardcoding the defaults directly. Only values that are unlikely to
change between runs belong in the config file, such as dependencies,
servers, version numbers, network drives, and any other defaults for
your lab or institutions. Configuration files can be in a standard
location, e.g. \texttt{.packagerc} in the user's home directory or
provided on the command line as an additional argument.

We caution against overusing configuration files. When a user needs to
locate, open, change and save a file in order to change a parameter, the
import of the change seems larger and discourages experimentation. Since
all parameters should be echoed in the results, config files must be
cleaned up after execution completes or they occupy valuable disk space.
Use a default configuration file and specify all other parameters on the
command line.

* They make fools out of you and me.

\section{Eliminate hard-coded paths.}

It's easy to write software that reads input from a file called
\texttt{mydata.csv}, but also very limiting. If a colleague asks you to
process her data, you must either overwrite your data file (which is
risky) or edit your code to read \texttt{otherdata.csv} (which is also
risky, because there's every likelihood you'll forget to change the
filename back, or will change three uses of the filename but not a
fourth).

Hard-coding filenames in a program also makes the software harder to run
in other environments. If your package is installed on a cluster, for
example, the user's data will almost certainly \emph{not} be in the same
directory as the software, and the folder
\texttt{C:\textbackslash{}users\textbackslash{}yourname\textbackslash{}}
will probably not even exist.

For these reasons, users should be able to set the names and locations
of input and output files as command-line parameters. This rule applies
to reference data sets as well as the user's own data: if a user wants
to try a new gene identification algorithm using a different set of
genes as a training set, she should not have to edit the software to do
so.

A corollary to this rule is that a package should not require users to
navigate to a particular directory to do their work. ``Where I have to
be'' is just another hard-coded path.

In order to save typing, it is often convenient to allow users to
specify an input or output \emph{directory}, and then require that there
be files with particular names in that directory. This practice, which
is sometimes called ``convention over configuration'', is used by many
software frameworks, such as WordPress and Ruby on Rails, and often
strikes a good balance between adaptability and consistency.

\section{Reuse software (without tears).}

In the spirit of code reuse and interoperability, developers often want
to use software written by others. The tool could be standard for the
task at hand or do exactly what is necessary. With a few lines, a call
is made out to the other program, the results are incorporated into the
primary script. Using popular projects reduces the amount of code that
needs to be maintained and adds the strength of vetted software to the
final program.

Unfortunately, the interface between two software packages can be a
source of considerable frustration. Support requests descend into
debugging errors produced by the other project.

The way that the second program is invoked can throw up errors. For
example, the program may not be in the user's path, or it may be an
older or newer version and produce results differen than expected.
Windows users will be frustrated if you invoke bash or sh explictly or
use shell-specific conventions like `*' expansion. It is unlikely that
all of your users will all be on the operating system and version as
you. Even Linux-standard functions available vary slightly between
installs. For example: GNU \texttt{sort} is available on almost every
*nix distribution, but sorts differently depending on locale.

We fully support the reuse of software between projects and have some
suggestions to reduce the aforementioned pains. First, ensure the
appropriate software and version is available. Either allow the user to
configure the exact path to the package, distribute the program with the
dependent software, or download it during installation using a
dependency management system. Regardless, check whether the program is
executable and what version is running. Be sure to remind users in the
documentation what the compatible versions are.

Second, to ensure support on as many different operating systems as
possible, use native functions for starting other processes, such as
Java's Runtime.exec call, Python's subprocess module, and Perl's system
command, and be sure to capture and report the standard error output of
the subprocess to facilitate debugging.

Finally, make sure that you really need the call-out. If you are
executing GNU sort instead of figuring out how to sort lists in Python,
it may not be worth the tears of reuse.

\section{Do not rely on the pre-installation of non-standard packages or libraries unless clearly stated in the documentaton.}

Every package someone has to install before being able to use yours is a
possible (some would say ``likely'') source of frustration for some
potential user. On the other hand, research software developers should
re-use existing software wherever possible. To strike a balance between
these two, developers should document \emph{all} of the packages that
theirs depends on, preferably in a machine-readable form. For example,
it is common for Python projects to include a file called
\texttt{requirements.txt} that lists the names of required libraries,
along with version ranges:

\begin{verbatim}
requests>=2.0
pygithub>=1.26,<=1.27
python-social-auth>=0.2.19,<0.3
\end{verbatim}

This file can be read by a package manager, which can check that the
required software is available, and install it if it is not. Similar
mechanisms exist for Perl, R, and other languages.

A common way to break this rule is to depend on scripts and tools that
are installed on the computers the original developer is using, but
which aren't otherwise packaged and available. In many cases, the author
of a package may not realize that some tool was built locally, and
doesn't exist elsewhere. At present, the only sure way to discover such
unknown dependencies is to install on a system administered by someone
else and see what breaks. In future, as use of lightweight
virtualization containers like Docker becomes more widespread, it may
become common to test installation on a virtual machine.

\section{Produce identical results when given identical inputs.}

Given a set of parameters and a dataset, the package should produce the
same results every time it is run.

Many bioinformatics applications rely on randomized algorithms to
improve performance or runtimes. As a consequence, results can change
between runs, even when provided with the same data and parameters. By
its nature, this randomness renders strict reproducibility impossible.
Debugging is more difficult. If even the small test set (\#9) produces
different results, new users may be able to tell whether the software is
working properly, eroding confidence in the application. When comparing
results between versions or after changing parameters, even small
differences can confuse or muddy the comparison. And especially when
producing results for publications, grants or diagnoses, any analysis
should be absolutely reproducible.

Given the size of biological data, it is unreasonable to suggest that
random algorithms be removed. However, most programs use a pseudo-random
number generator, which uses a starting seed and an equation to
approximate random numbers. Setting the seed to a consistent value
removes randomness between runs. Allow the user to optionally provide
the seed as an input parameter, thus rendering the program deterministic
for those cases where it matters. If the seed is set internally (e.g.,
using clock time), echo it to the output for re-use later.

\section{Include a small test set that can be run to ensure the software is actually working.}

Every package should come with a small test script for users to run
after installation. Its purpose is \emph{not} to check that the software
is working correctly (although that is extremely helpful), but rather to
ensure that it will work at all. This test script can also serve as a
working example of how to run the software, which is valuable in case
its documentation has fallen out of sync with recent changes to the code
itself.

In order to be useful, this test script must be easy to find and run. A
single file in the project's root directory named \texttt{runtests.sh}
or something equally obvious is a much better solution than documenting
test cases and requiring people to copy and paste them into the shell.

Equally, the test script's output must be easy to interpret. Screens
full of correlation coefficients do not qualify: instead, the script's
output should be something like one line per test, with the test's name
and its pass/fail status, followed by a single summary line saying how
many tests were run and how many passed or failed. If many or all tests
fail because of missing dependencies, that fact should be displayed
once, clearly, rather than once per test, so that users have a clear
idea of what they need to fix and how much work it's likely to take.

\section{Give the software a meaningful version number.}

Most software has a version number composed of a decimal number that
increments as new versions are released. There are many different ways
to construct and interpret the version, but most importantly for us, a
particular software version run with the same parameters should give
identical results no matter when it's run. Results include correct
output as well as any errors, whether they arise from incorrect input or
are bugs. Every time you release your software, i.e.~distribute it to
someone other than yourself and/or the development team, you should
increment your version number.

\withurl{Semantic versioning}{http://semver.org/} is one of the most common
types of versioning for open-source software. Version numbers take the
form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6-RC1. The major and
minor numbers versions are almost always provided. Changes in the major
version number herald significant changes in the software that are not
backwards compatible, such as changing or removing features or altering
the primary functions of the software. Increasing the minor version
represents incremental improvements in the software like adding new
features. Following the minor version number can be an arbitrary number
of project-specific versions, including patches, builds and qualifiers.
Common qualifiers include \texttt{-SNAPSHOT}, for applications that are
not yet stable or released, and \texttt{-RC} for release candidate prior
to official release.

The version of your software should be easily available, both when
supplying \texttt{-\/-version} or \texttt{-v} on the command line as
well as in the results. The software version should be printed to the
same location as all of the other parameters (see Rule 4).

Old versions of your software should be available to ensure that results
are reproducible into the far future. A number of mechanisms exist for
controlled release that range from as simple as adding an appropriate
commit message or tag to version control, to official releases alongside
code on Sourceforge, Bitbutcket or Github, to depositing into a
repository like apt, yum, homebrew, CPAN, etc. Choose the method that
best suits the number and expertise of users you anticipate.

\section*{Conclusion}

\fixme{GW}{How to tell if the software is working:}

\begin{itemize}
\item
  Test in a `vanilla' environment, such as another user's computer or a
  dummy account with none of the settings of the original developer.
\item
  Test with different sizes of data: ridiculously small, small, medium,
  and large. The software should run on all sizes given some parameter
  tweaking, or fail with a sensible error message if the data is too
  big.
\item
  Compare results from multiple iterations that have the same parameters
  and inputs.(See rule \#8.)
\item
  Consider a container - but most of these rules still apply
\end{itemize}

\bibliography{robust-software}

\end{document}

### 2016-08-11 robust-software.tex

@@
 have made efforts towards cataloguing it: as of May 2016, the former
-has 2500 entries and the latter 1700, and those numbers are constantly
-growing as both trainees and experienced practitioners produce new
+has 2500 entries and the latter 1700. Those numbers constantly
+grow as both trainees and experienced practitioners produce new
 software to support their research.
 
-Typically, this software is initially developed by one person, and may
-produce excellent results in their hands. But what happens after that
-student leaves that lab and someone else wants to use it? Everyone with
+Typically, this software is developed by one person and 
+produces excellent results in their hands. But what happens after the
+student leaves someone else from their group wants to use the program? Everyone with
 a few years of experience feels a tremor of fear when told, ``Use
-{\textless}graduated student{\textgreater}'s code to analyze your
+Graduated Student's code to analyze your
 data''. Often, that software will be undocumented and work in unexpected
-ways (if it works at all without substantial modification). Equally
-often, the potential new user winds up shaking their fist and cursing
-the author's name. She then has two choices: hack the existing code to
-make it work for her, or start over.
+ways, if it works at all, relying on nonexistent paths or resources, tuned for
+the dataset, or simply an older version than was used in the paper.
+The poor student or staff member then has two choices: hack the existing code
+to make it work for her, or start over.
 
@@
 \item
+  Works for users other than the original author
+\item
   Is kept under version control.
 \item
-  Can be installed on systems outside the original institution
-\item
-  Works for users other than the original author
+  Can be installed on more than one computer or account with relative ease
 \item
@@
 
-These are all necessary steps toward creating a reusable library that
-can be shared through a site like CPAN or CRAN, and apply to both
-closed-source and open-source software. They do not depend on specific
-languages, libraries, packages, documentation styles, or operating
-systems. Whether the aim is as simple as sharing the code with
-collaborators or as complex as using the software in a production
+These rules need not be applied to \emph{every} coding effort. The vast majority of
+code produced in the marathon of a graduate thesis is used once to answer a
+specific question related to a specific dataset. Scientific endeavours often
+require speed and rapid iteration over a question, not thoughtful software
+engineering and thorough documentation. A README is of no use if you are the
+sole user of your code. As the saying goes, not everything worth doing is worth
+doing right away. However, once your script is dusted off and run three or four
+times for slightly different purposes, when it comes time to publish your
+methods, to leave a project, or to pass it on to someone else, it may be time
+to apply ten simple rules for robust software.
+
+These simple rules are also necessary steps toward creating a reusable library that
+can be shared through a site like CPAN or CRAN. They can apply to both
+closed-source and open-source software.  They are generic and can be applied to all
+languages, libraries, packages, documentation styles, and
+operating systems.  Whether the aim is as simple as sharing the code
+with collaborators or as complex as using the software in a production
 analysis environment, increasing the robustness of your software
@@
 
-Note: we do not recommend that these rules be applied to \emph{every}
-coding effort. The vast majority of code produced in the marathon of a
-graduate thesis is ``throw-away'' code that is used once to answer a
-specific question related to a specific dataset. However, once that
-little script is dragged out three or four times for slightly different
-purposes, it may be time to apply ten simple rules for robust software.
-As the saying goes, not everything worth doing is worth doing right
-away.
-
 \section{Have a README that explains in a few lines what the software does and what its dependencies are.}
 
-The README is the first stop for any potential users interested in your
-software. At a minimum, it needs to provide or point to everything a new
-user needs to get started, where they can turn to for help, and which
-licenses apply to the software package. Exhaustive details regarding
-parameters and usage are not usually necessary in a README if they are
-present in usage (\#2), although a working example using test data (per
+The README is the first stop for most new users.
+At a minimum, it needs to get a new user started and point them towards more
+help, if they need it.
+A working example using test data (per
 \#9) is always appreciated.
@@
 \textbf{Explain what the software does:} At the beginning of the README,
-explain what the software does in one or two sentences. The description
-does not need to be long or detailed. There's nothing more frustrating
+explain what the software does in one or two sentences.
+There's nothing more frustrating
 than spending the time to download and install some software only to
 find out that it doesn't do what you thought it did.
-
-\begin{verbatim}
-Debarcer (De-Barcoding and Error Correction) is a package for
-working with next-gen sequencing data that contains molecular
-barcodes.
-\end{verbatim}
 
@@
 link to the software homepage. Include the version number for each
-dependency.  Especially if you use older versions, include links where
-the packages can be downloaded. Package managers like apt, pip and
-homebrew stop offering older packages after a few years.
-\fixme{MT}{Is this true of pip and Homebrew?}
+dependency.
 
-If your dependency is to an internal package that is not available on
-the internet, you have several options depending on the sensitivity of
-the code in question. If it is plain text, you can add it directly to
-you repository with appropriate attribution. If the dependency is a
-binary, we recommend using a
+If your dependency is not publicly available,
+you have several options to ensure your software remains useable.
+You can add plain-text or code directly to
+your repository with appropriate attribution. If the dependency is a binary, use a
 \withurl{binary repository manager}{https://en.wikipedia.org/wiki/Binary\_repository\_manager}
@@
 software at constant URLs so they can be downloaded as long as the
-manager continues to run. As a last resort, you can place it at an
+manager continues to run. If the dependency must remain closed, place it at an
 internal location on shared disk, remove all write permissions, and
-link to it from your README, although this method is heavily
-discouraged because of the potential for the directory to go missing
-because of factors outside the developer's control.
+link to it from your README, although this method is 
+discouraged because of the potential for sharing and risks accidental removal.
 
 \textbf{Installation instructions:} If the software needs to be compiled
-or installed, list those instructions in the readme. New users may not
-be familiar with your build system, even if it is \texttt{make}. Also
-mention here if you recommend they use a pre-compiled binary instead
-through a system such as pip or apt.
+or installed, list those instructions in the README. Also
+mention if you recommend they use a pre-compiled binary instead
+through a system such as pip, yum or apt.
 
@@
 should be listed in this section. Do the files conform to a particular
-industry standard, an extension of an existing format, or is it your own
+standard, an extension of an existing format, or is it your own
 format? If using a standard format, link to the specification and
 version. If you extend the standard or have your own format, define it
-here explicitly, listing all the required fields and acceptable values.
+explicitly, listing all the required fields and acceptable values.
 (You get bonus points if you include a script to convert between
-standard format and your file format). If there is no rigorous format
+a standard format and your file format). If there is no rigorous format
 (such as with log files), show an example file, or the first few lines,
@@
 
-Input files and their formats are included in most documentation.
-However, the definitions of the output files are often missing. In
-addition to the expected output, software will often produce
-intermediate files, auxilliary files, and log files. We believe
-\emph{all} output files should be listed in the README. Log and
-auxilliary files are often full of valuable information that can be
-mined for the user's specific purpose. Even if the files are considered
-self-explanatory. Sometimes your users will need to answer a question of
-the format, ``Does X tell you the percentage of reads trimmed to remove
-adapter sequences?'' and you can check the documentation and confidently
+Input files and their formats are included in most documentation, but
+intermediate, auxilliary and log files are often missing. 
+\emph{All} files should be listed in the README, even those considered self-explanatory. Log
+files are often full of valuable information that can be
+mined for the user's specific purpose. Sometimes your users will need to answer a question of
+the format, ``Does Software tell you the percentage of reads trimmed to remove
+adapter sequences?'' and they can check the README and confidently
 say ``yes, it is in the log file''.
@@
 your main contributors; licenses are how you want others to use and
-credit your software. Both are important in your README. Leave no
+credit your software. By putting it in the README, it is readable \emph{before}
+the software is installed (or even downloaded).
+Leave no
 question in anyone's mind about whether your software can be used
 commercially, how much modification is permitted, and how other software
-needs to attribute to you. If your software is not open source, include
-a statement here. Attributions can also contain a list of `expert' users
+needs to credit you. If your software is not open source, state that clearly.
+Attributions can also contain a list of `expert' users
 that can be contacted if new users have problems with the software. (for
@@
 \begin{itemize}
-\item
-  This is readable \emph{before} the software is installed (or even
-  downloaded).
-\item
-  Should also include (or better yet, point to) the license for the
-  software, so that (potential) users will know what they're allowed to
-  do.
 \item
@@
 
-Users who run your software after installation may not have access your
-well-crafted README (or may not have bothered to read it). Usage
-information provides their first line of help.
-
+Usage information provides the first line of help for both first-time and
+experienced users of command-line applications.
 Ideally, usage is a terse, informative command-line help message that
 guides the user in the correct use of your software. Terseness is
-important: usage that extends for multiple screens, especially when
+important: usage that extends for multiple screens is a nuisance, especially when
 printed to standard error instead of standard output (where it can
-easily be paged), is a nuisance, and is as unlikely to be read as the
-README file.
+easily be paged).
 
 Usage should provide all of the information necessary to run the
-software. It is usually invoked either by running the software without
+software. It is invoked either by running the software without
 any arguments; running the software with incorrect arguments; or by
@@
 
-\textbf{The syntax for running the program}: This defines the relative
-location of optional and required flags and arguments for execution, and
-includes the name of the program. Arguments in {[}square brackets{]}
-tend to be optional. Multiple periods (e.g. ``{[}OPTION{]}\ldots{}'')
-indicate that more than one can be provided.
+\textbf{The syntax for running the program}: This includes the name of the
+program and defines the relative
+location of optional and required flags, arguments and values for execution.
+Arguments in {[}square brackets{]}
+are usually optional. An ellipsis (\ldots e.g. ``{[}OPTION{]}\ldots{}'')
+indicates that more than one value can be provided.
 
-\textbf{A text description of its purpose}: Similar to the README, the
+\textbf{Description}: Similar to the README, the
 description reminds users of the software's primary function.
 
-\textbf{Most commonly used flags, a description of each flag, and the
-default value}: Not all flags need to appear in the usage, but the most
+\textbf{Most commonly used arguments, a description of each, and the
+default values}: Not all arguments need to appear in the usage, but the most
 commonly used ones should be listed here. Users will rely on this for
@@
 \texttt{less}, searched with \texttt{grep}, or compared to the previous
-version's help with \texttt{diff}.
+version with \texttt{diff}.
 
 \textbf{Exit with an appropriate exit code}: When usage is invoked by
-providing incorrect parameters, the program should exit with a non-zero
+providing incorrect arguments, the program should exit with a non-zero
 code to indicate an error. However, when help is explicitly requested,
-the software should not exit with an error, because requesting help is
-sometimes used to verify that a dependency is available.
+the software should not exit with an error.
 
@@
 system files and user accounts. Conversely, files and directories owned
-by root usually \emph{cannot} be modifed by normal users.
+by root usually cannot be modifed by normal users.
 
@@
 
-You know what they say about assumptions*. As soon as you make one
-assumption about how a software should work, a use case will come along
-that will require you to change the assumption.
-
-Every useful parameter should be configurable on the command line.
-Useful parameters are those that a user will need to modify to suit
-their computer, dataset or application. Providing parameters on the
-command line increases the flexibility and usability of the program. You
-may have determined early on that 0.58 is an optimal seed for your
-original dataset, but that doesn't mean that is the best seed for every
-case. Being able to change parameters on the fly to determine if and how
-they change the results is important as your software gains more users,
-facilitating exploratory analysis and parameter sweeping. Keep in mind
-that if a parameter can be adjusted, users will want to be able to turn
-off the feature entirely to have a baseline comparison.
-
-The list of useful parameters is software-specific and so cannot be
-provided here, but here is a short list of common useful parameters.
+Useful parameters are those values that a user will need to modify to suit
+their computer, dataset or application. 
+The full list of useful parameters is software-specific and so cannot be
+detailed here, but here is a short list:
 
@@
 
+Providing parameters on the
+command line increases the flexibility and usability of the program. You
+may have determined early in development that 0.58 is an optimal seed for your
+original dataset, but that doesn't mean that is the best seed for every
+case. Being able to change parameters on the fly to determine if and how
+they change the results is important as your software gains more users,
+facilitating exploratory analysis and parameter sweeping. Keep in mind
+that if a parameter can be adjusted, users will want to be able to turn
+off the feature entirely to have a baseline comparison.
+
 When the software starts, it should echo all parameters and software
@@
 
-You can set reasonable default values to reduce the length of the
-execution command as long as any parameters given on the command line
+You can set reasonable default values
+as long as any command line arguments
 override those values. Configuration files should be used in preference
-to hardcoding the defaults directly. Only values that are unlikely to
-change between runs belong in the config file, such as dependencies,
-servers, version numbers, network drives, and any other defaults for
-your lab or institutions. Configuration files can be in a standard
+to hardcoding the defaults directly. Configuration files can be in a standard
 location, e.g. \texttt{.packagerc} in the user's home directory or
@@
 locate, open, change and save a file in order to change a parameter, the
-import of the change seems larger and discourages experimentation. Since
-all parameters should be echoed in the results, config files must be
+import of the change seems larger and discourages experimentation. 
+Only values that are unlikely to
+change between runs belong in the config file, such as dependencies,
+servers, version numbers, network drives, and any other defaults for
+your lab or institution. 
+Since
+all parameters should be echoed in the results, config files should be
 cleaned up after execution completes or they occupy valuable disk space.
-Use a default configuration file and specify all other parameters on the
+Use a default configuration file and specify all other parameter values on the
 command line.
-
-* They make fools out of you and me.
 
@@
 
-Hard-coding filenames in a program also makes the software harder to run
+Hard-coding file paths in a program also makes the software harder to run
 in other environments. If your package is installed on a cluster, for
@@
 In the spirit of code reuse and interoperability, developers often want
-to use software written by others. The tool could be standard for the
-task at hand or do exactly what is necessary. With a few lines, a call
+to reuse software written by others. 
+With a few lines, a call
 is made out to the other program, the results are incorporated into the
@@
 
-The way that the second program is invoked can throw up errors. For
-example, the program may not be in the user's path, or it may be an
-older or newer version and produce results differen than expected.
+The way that the second program is invoked can throw up errors.
+The program may not be in the user's path, or it may be an
+older or newer version and produce results different than expected.
 Windows users will be frustrated if you invoke bash or sh explictly or
-use shell-specific conventions like `*' expansion. It is unlikely that
-all of your users will all be on the operating system and version as
-you. Even Linux-standard functions available vary slightly between
+use shell-specific conventions like `*' expansion.
+Even Linux-standard functions available vary slightly between
 installs. For example: GNU \texttt{sort} is available on almost every
@@
 We fully support the reuse of software between projects and have some
-suggestions to reduce the aforementioned pains. First, ensure the
+suggestions to reduce the aforementioned pains. First, 
+make sure that you really need the call-out. If you are
+executing GNU sort instead of figuring out how to sort lists in Python,
+it may not be worth the pain of integration.
+
+Second, ensure the
 appropriate software and version is available. Either allow the user to
@@
 
-Second, to ensure support on as many different operating systems as
+Finally, to ensure support on as many different operating systems as
 possible, use native functions for starting other processes, such as
@@
 the subprocess to facilitate debugging.
-
-Finally, make sure that you really need the call-out. If you are
-executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the tears of reuse.
 
@@
 Debugging is more difficult. If even the small test set (\#9) produces
-different results, new users may be able to tell whether the software is
-working properly, eroding confidence in the application. When comparing
+different results, new users may not be able to tell whether the software is
+working properly. When comparing
 results between versions or after changing parameters, even small
@@
 ensure that it will work at all. This test script can also serve as a
-working example of how to run the software, which is valuable in case
-its documentation has fallen out of sync with recent changes to the code
-itself.
+working example of how to run the software.
 
@@
 full of correlation coefficients do not qualify: instead, the script's
-output should be something like one line per test, with the test's name
+output should be simple to understand for non-experts, such as one line per test, with the test's name
 and its pass/fail status, followed by a single summary line saying how
@@
 particular software version run with the same parameters should give
-identical results no matter when it's run. Results include correct
-output as well as any errors, whether they arise from incorrect input or
-are bugs. Every time you release your software, i.e.~distribute it to
+identical results no matter when it's run. Results include both correct
+output as well as any errors.
+Every time you release your software, i.e.~distribute it to
 someone other than yourself and/or the development team, you should
@@
 types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6-RC1. The major and
+form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6. The major and
 minor numbers versions are almost always provided. Changes in the major
@@
 the primary functions of the software. Increasing the minor version
-represents incremental improvements in the software like adding new
+represents incremental improvements in the software, like adding new
 features. Following the minor version number can be an arbitrary number
-of project-specific versions, including patches, builds and qualifiers.
-Common qualifiers include \texttt{-SNAPSHOT}, for applications that are
-not yet stable or released, and \texttt{-RC} for release candidate prior
+of project-specific identifiers, including patches, builds and qualifiers.
+Common qualifiers include \texttt{alpha}, \texttt{beta}, and \texttt{SNAPSHOT}, for applications that are
+not yet stable or released, and \texttt{-RC} for release candidates prior
 to official release.

### 2016-08-12 robust-software.tex
Reorganizing the rules.
1. Minor editing on some existing rules.
2. Merged two rules about depending on external programs.
3. Reordered: 1, 2, 10, 3, 7/6, 5, 4, 9, 8
4. Added a paragraph for a new rule about keeping old versions available - needs work.

@@
 \begin{center}\textbf{Abstract}\end{center}
-\fixme{GW}{Write abstract}
+Software produced for research---even published software---suffers from a number of common problems
+that make it difficult or impossible to run outside the original institution
+or even the primary developer's computer.
+We present ten simple rules to make software robust enough to run anywhere
+and delight your users and collaborators.
 \end{quote}
@@
 student leaves someone else from their group wants to use the program? Everyone with
-a few years of experience feels a tremor of fear when told, ``Use
-Graduated Student's code to analyze your
-data''. Often, that software will be undocumented and work in unexpected
-ways, if it works at all, relying on nonexistent paths or resources, tuned for
-the dataset, or simply an older version than was used in the paper.
-The poor student or staff member then has two choices: hack the existing code
-to make it work for her, or start over.
+a few years of experience feels a tremor of fear when told to use
+a graduated student's code to analyze their data.
+Often, that software will be undocumented and work in unexpected
+ways (if it works at all).  It will often rely on nonexistent paths or resources,
+be tuned for a single dataset,
+or simply be an older version than was used in published papers.
+Its new user is often faced with two unpalatable options:
+hack the existing code to make it work, or start over.
 
-The root cause of this problem is that most of the software researchers
-produce isn't \emph{robust}. The difference between running and being
+The root cause of this problem is that most research software
+isn't \emph{robust}. The difference between running and being
 robust is the difference between ``works for me on my machine'' and
@@
 \item
-  Works for users other than the original author
+  works for users other than the original author;
 \item
-  Is kept under version control.
+  is kept under version control;
 \item
-  Can be installed on more than one computer or account with relative ease
+  can be installed on more than one computer or account with relative ease;
 \item
-  Has well-defined input and output formats
+  has well-defined input and output formats;
 \item
-  Has documentation that describes what its dependencies are, how to
-  install it, and what the options are.
+  has documentation that describes what its dependencies are, how to
+  install it, and what the options are; and
 \item
-  Comes with enough tests to show that it actually runs.
+  comes with enough tests to show that it actually runs.
 \end{itemize}
 
-These rules need not be applied to \emph{every} coding effort. The vast majority of
-code produced in the marathon of a graduate thesis is used once to answer a
-specific question related to a specific dataset. Scientific endeavours often
-require speed and rapid iteration over a question, not thoughtful software
-engineering and thorough documentation. A README is of no use if you are the
-sole user of your code. As the saying goes, not everything worth doing is worth
-doing right away. However, once your script is dusted off and run three or four
-times for slightly different purposes, when it comes time to publish your
-methods, to leave a project, or to pass it on to someone else, it may be time
-to apply ten simple rules for robust software.
+These rules need not be applied to \emph{every} coding effort:
+as the saying goes, not everything worth doing is worth doing right (or right away).
+Code  that is used once to answer a specific question related to a specific dataset
+doesn't require comprehensive documentation,
+and the only sensible way to test it may well be to run it on the dataset in question.
+However, if a script is dusted off and run three or four
+times for slightly different purposes,
+crucial to a publication,
+or being passed on to someone else,
+it may be time to apply ten simple rules for robust software.
 
-These simple rules are also necessary steps toward creating a reusable library that
-can be shared through a site like CPAN or CRAN. They can apply to both
-closed-source and open-source software.  They are generic and can be applied to all
-languages, libraries, packages, documentation styles, and
-operating systems.  Whether the aim is as simple as sharing the code
-with collaborators or as complex as using the software in a production
-analysis environment, increasing the robustness of your software
-decreases headaches all around.
+These simple rules are also necessary steps toward creating a reusable
+library that can be shared through a site like CPAN or CRAN.  They are
+generic and can be applied to all languages, libraries, packages,
+documentation styles, and operating systems for both closed-source and
+open-source sofwtare.  Whether the aim is as simple as sharing the
+code with collaborators or as complex as using the software in a
+production analysis environment, increasing the robustness of your
+software decreases headaches all around.
 
@@
 files are often full of valuable information that can be
-mined for the user's specific purpose. Sometimes your users will need to answer a question of
-the format, ``Does Software tell you the percentage of reads trimmed to remove
-adapter sequences?'' and they can check the README and confidently
-say ``yes, it is in the log file''.
+mined for the user's specific purpose. If your users might need to know,
+``Does this program report the percentage of reads trimmed to remove
+adapter sequences?'' they should be able to check the README and confidently
+say, ``Yes, it is in the log file''.
 
@@
 your main contributors; licenses are how you want others to use and
-credit your software. By putting it in the README, it is readable \emph{before}
+credit your software. By putting them in the README, it is readable \emph{before}
 the software is installed (or even downloaded).
@@
 needs to credit you. If your software is not open source, state that clearly.
-Attributions can also contain a list of `expert' users
-that can be contacted if new users have problems with the software. (for
-better or for worse)
+Attributions can also contain a list of ``expert'' users
+that can be contacted if new users have problems with the software.
 
-\begin{itemize}
-\item
-  \fixme{GW}{example from real software:\\ https://github.com/dib-lab/khmer/blob/master/README.rst}
-\end{itemize}
+The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
+is a good model:
+it explains the software's purpose,
+tells readers where to get help,
+and includes links to a CITATION file (explaining how to cite the project)
+and the license.
 
@@
 
+\section{Give the software a meaningful version number.}
+
+Most software has a version number composed of a decimal number that
+increments as new versions are released. There are many different ways
+to construct and interpret the version, but most importantly for us, a
+particular software version run with the same parameters should give
+identical results no matter when it's run. Results include both correct
+output as well as any errors.
+Every time you release your software, i.e.~distribute it to
+someone other than yourself and/or the development team, you should
+increment your version number.
+
+\withurl{Semantic versioning}{http://semver.org/} is one of the most common
+types of versioning for open-source software. Version numbers take the
+form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6. The major and
+minor numbers versions are almost always provided. Changes in the major
+version number herald significant changes in the software that are not
+backwards compatible, such as changing or removing features or altering
+the primary functions of the software. Increasing the minor version
+represents incremental improvements in the software, like adding new
+features. Following the minor version number can be an arbitrary number
+of project-specific identifiers, including patches, builds and qualifiers.
+Common qualifiers include \texttt{alpha}, \texttt{beta}, and \texttt{SNAPSHOT},
+for applications that are
+not yet stable or released, and \texttt{-RC} for release candidates prior
+to official release.
+
+The version of your software should be easily available, both when
+supplying \texttt{-\/-version} or \texttt{-v} on the command line as
+well as in the results. The software version should be printed to the
+same location as all of the other parameters (see Rule 4).
+
+Old versions of your software should be available to ensure that results
+are reproducible into the far future. A number of mechanisms exist for
+controlled release that range from as simple as adding an appropriate
+commit message or tag to version control, to official releases alongside
+code on Sourceforge, Bitbutcket or Github, to depositing into a
+repository like apt, yum, homebrew, CPAN, etc. Choose the method that
+best suits the number and expertise of users you anticipate.
+
+\section{Make older versions available.}
+
+Software evolves, but not all software evolves at the same pace.
+While more consistent command-line arguments
+or more flexible handling of output directories
+may make a program better in general,
+it can simultaneously make work for someone
+who integrated the old version into their own workflow a year or two ago,
+and won't see any benefits from upgrading.
+Programs' authors should therefore ensure that every version they have released
+continues to be available.
+
+\fixme{MT}{Can you please flesh this out?}
+
 \section{Do not require root or other special privileges.}
@@
 setting things up as root.
+
+\section{Reuse software (within reason).}
+
+In the spirit of code reuse and interoperability, developers often want
+to reuse software written by others. 
+With a few lines, a call
+is made out to the other program, the results are incorporated into the
+primary script. Using popular projects reduces the amount of code that
+needs to be maintained and adds the strength of vetted software to the
+final program.
+
+Unfortunately, the interface between two software packages can be a
+source of considerable frustration. Support requests descend into
+debugging errors produced by the other project.
+
+In addition, every package someone has to install
+before being able to use yours is a
+possible (some would say ``likely'') source of frustration for some
+potential user. On the other hand, research software developers should
+re-use existing software wherever possible.
+
+To strike a balance between
+these two, developers should
+\textbf{document \emph{all} of the packages that theirs depends on, preferably in a machine-readable form}.
+For example, it is common for Python projects to include a file called
+\texttt{requirements.txt} that lists the names of required libraries,
+along with version ranges:
+
+\begin{verbatim}
+requests>=2.0
+pygithub>=1.26,<=1.27
+python-social-auth>=0.2.19,<0.3
+\end{verbatim}
+
+This file can be read by a package manager, which can check that the
+required software is available, and install it if it is not. Similar
+mechanisms exist for Perl, R, and other languages.
+
+Conversely, developers should
+\textbf{avoid depending on scripts and tools which are not available as packages},
+even (or especially) if they are installed on the computers the original developer is using.
+In many cases, a program's author may not realize that some tool was built locally, and
+doesn't exist elsewhere. At present, the only sure way to discover such
+unknown dependencies is to install on a system administered by someone
+else and see what breaks. As use of lightweight
+virtualization containers like Docker becomes more widespread, it may
+become common to test installation on a virtual machine.
+
+The way that the second program is invoked can throw up errors.
+The program may not be in the user's path, or it may be an
+older or newer version and produce results different than expected.
+Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
+use shell-specific conventions like `*' expansion.
+Even Linux-standard functions available vary slightly between
+installs. For example, GNU \texttt{sort} is available on almost every
+*nix distribution, but sorts differently depending on locale.
+
+Despite these caveats,
+we fully support the reuse of software between projects provided
+three additional guidelines are adhered to. First, 
+\textbf{make sure that you really need the auxiliary program}. If you are
+executing GNU sort instead of figuring out how to sort lists in Python,
+it may not be worth the pain of integration.
+
+Second, \textbf{ensure the appropriate software and version is available}.
+Either allow the user to
+configure the exact path to the package, distribute the program with the
+dependent software, or download it during installation using a
+dependency management system. Regardless, check whether the program is
+executable and what version is running. Be sure to remind users in the
+documentation what the compatible versions are.
+
+Finally, to ensure support on as many different operating systems as
+possible, \textbf{use native functions for starting other processes}, such as
+Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
+command, and be sure to capture and report the standard error output of
+the subprocess to facilitate debugging.
+
+\section{Eliminate hard-coded paths.}
+
+It's easy to write software that reads input from a file called
+\texttt{mydata.csv}, but also very limiting. If a colleague asks you to
+process her data, you must either overwrite your data file (which is
+risky) or edit your code to read \texttt{otherdata.csv} (which is also
+risky, because there's every likelihood you'll forget to change the
+filename back, or will change three uses of the filename but not a
+fourth).
+
+Hard-coding file paths in a program also makes the software harder to run
+in other environments. If your package is installed on a cluster, for
+example, the user's data will almost certainly \emph{not} be in the same
+directory as the software, and the folder
+\texttt{C:\textbackslash{}users\textbackslash{}yourname\textbackslash{}}
+will probably not even exist.
+
+For these reasons, users should be able to set the names and locations
+of input and output files as command-line parameters. This rule applies
+to reference data sets as well as the user's own data: if a user wants
+to try a new gene identification algorithm using a different set of
+genes as a training set, she should not have to edit the software to do
+so.
+
+A corollary to this rule is that a package should not require users to
+navigate to a particular directory to do their work. ``Where I have to
+be'' is just another hard-coded path.
+
+In order to save typing, it is often convenient to allow users to
+specify an input or output \emph{directory}, and then require that there
+be files with particular names in that directory. This practice, which
+is sometimes called ``convention over configuration'', is used by many
+software frameworks, such as WordPress and Ruby on Rails, and often
+strikes a good balance between adaptability and consistency.
 
@@
 
-\section{Eliminate hard-coded paths.}
+\section{Include a small test set that can be run to ensure the software is actually working.}
 
-It's easy to write software that reads input from a file called
-\texttt{mydata.csv}, but also very limiting. If a colleague asks you to
-process her data, you must either overwrite your data file (which is
-risky) or edit your code to read \texttt{otherdata.csv} (which is also
-risky, because there's every likelihood you'll forget to change the
-filename back, or will change three uses of the filename but not a
-fourth).
+Every package should come with a small test script for users to run
+after installation. Its purpose is \emph{not} to check that the software
+is working correctly (although that is extremely helpful), but rather to
+ensure that it will work at all. This test script can also serve as a
+working example of how to run the software.
 
-Hard-coding file paths in a program also makes the software harder to run
-in other environments. If your package is installed on a cluster, for
-example, the user's data will almost certainly \emph{not} be in the same
-directory as the software, and the folder
-\texttt{C:\textbackslash{}users\textbackslash{}yourname\textbackslash{}}
-will probably not even exist.
+In order to be useful, this test script must be easy to find and run. A
+single file in the project's root directory named \texttt{runtests.sh}
+or something equally obvious is a much better solution than documenting
+test cases and requiring people to copy and paste them into the shell.
 
-For these reasons, users should be able to set the names and locations
-of input and output files as command-line parameters. This rule applies
-to reference data sets as well as the user's own data: if a user wants
-to try a new gene identification algorithm using a different set of
-genes as a training set, she should not have to edit the software to do
-so.
-
-A corollary to this rule is that a package should not require users to
-navigate to a particular directory to do their work. ``Where I have to
-be'' is just another hard-coded path.
-
-In order to save typing, it is often convenient to allow users to
-specify an input or output \emph{directory}, and then require that there
-be files with particular names in that directory. This practice, which
-is sometimes called ``convention over configuration'', is used by many
-software frameworks, such as WordPress and Ruby on Rails, and often
-strikes a good balance between adaptability and consistency.
-
-\section{Reuse software (without tears).}
-
-In the spirit of code reuse and interoperability, developers often want
-to reuse software written by others. 
-With a few lines, a call
-is made out to the other program, the results are incorporated into the
-primary script. Using popular projects reduces the amount of code that
-needs to be maintained and adds the strength of vetted software to the
-final program.
-
-Unfortunately, the interface between two software packages can be a
-source of considerable frustration. Support requests descend into
-debugging errors produced by the other project.
-
-The way that the second program is invoked can throw up errors.
-The program may not be in the user's path, or it may be an
-older or newer version and produce results different than expected.
-Windows users will be frustrated if you invoke bash or sh explictly or
-use shell-specific conventions like `*' expansion.
-Even Linux-standard functions available vary slightly between
-installs. For example: GNU \texttt{sort} is available on almost every
-*nix distribution, but sorts differently depending on locale.
-
-We fully support the reuse of software between projects and have some
-suggestions to reduce the aforementioned pains. First, 
-make sure that you really need the call-out. If you are
-executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration.
-
-Second, ensure the
-appropriate software and version is available. Either allow the user to
-configure the exact path to the package, distribute the program with the
-dependent software, or download it during installation using a
-dependency management system. Regardless, check whether the program is
-executable and what version is running. Be sure to remind users in the
-documentation what the compatible versions are.
-
-Finally, to ensure support on as many different operating systems as
-possible, use native functions for starting other processes, such as
-Java's Runtime.exec call, Python's subprocess module, and Perl's system
-command, and be sure to capture and report the standard error output of
-the subprocess to facilitate debugging.
-
-\section{Do not rely on the pre-installation of non-standard packages or libraries unless clearly stated in the documentaton.}
-
-Every package someone has to install before being able to use yours is a
-possible (some would say ``likely'') source of frustration for some
-potential user. On the other hand, research software developers should
-re-use existing software wherever possible. To strike a balance between
-these two, developers should document \emph{all} of the packages that
-theirs depends on, preferably in a machine-readable form. For example,
-it is common for Python projects to include a file called
-\texttt{requirements.txt} that lists the names of required libraries,
-along with version ranges:
-
-\begin{verbatim}
-requests>=2.0
-pygithub>=1.26,<=1.27
-python-social-auth>=0.2.19,<0.3
-\end{verbatim}
-
-This file can be read by a package manager, which can check that the
-required software is available, and install it if it is not. Similar
-mechanisms exist for Perl, R, and other languages.
-
-A common way to break this rule is to depend on scripts and tools that
-are installed on the computers the original developer is using, but
-which aren't otherwise packaged and available. In many cases, the author
-of a package may not realize that some tool was built locally, and
-doesn't exist elsewhere. At present, the only sure way to discover such
-unknown dependencies is to install on a system administered by someone
-else and see what breaks. In future, as use of lightweight
-virtualization containers like Docker becomes more widespread, it may
-become common to test installation on a virtual machine.
+Equally, the test script's output must be easy to interpret. Screens
+full of correlation coefficients do not qualify: instead, the script's
+output should be simple to understand for non-experts,
+such as one line per test, with the test's name
+and its pass/fail status, followed by a single summary line saying how
+many tests were run and how many passed or failed. If many or all tests
+fail because of missing dependencies, that fact should be displayed
+once, clearly, rather than once per test, so that users have a clear
+idea of what they need to fix and how much work it's likely to take.
 
@@
 
-\section{Include a small test set that can be run to ensure the software is actually working.}
-
-Every package should come with a small test script for users to run
-after installation. Its purpose is \emph{not} to check that the software
-is working correctly (although that is extremely helpful), but rather to
-ensure that it will work at all. This test script can also serve as a
-working example of how to run the software.
-
-In order to be useful, this test script must be easy to find and run. A
-single file in the project's root directory named \texttt{runtests.sh}
-or something equally obvious is a much better solution than documenting
-test cases and requiring people to copy and paste them into the shell.
-
-Equally, the test script's output must be easy to interpret. Screens
-full of correlation coefficients do not qualify: instead, the script's
-output should be simple to understand for non-experts, such as one line per test, with the test's name
-and its pass/fail status, followed by a single summary line saying how
-many tests were run and how many passed or failed. If many or all tests
-fail because of missing dependencies, that fact should be displayed
-once, clearly, rather than once per test, so that users have a clear
-idea of what they need to fix and how much work it's likely to take.
-
-\section{Give the software a meaningful version number.}
-
-Most software has a version number composed of a decimal number that
-increments as new versions are released. There are many different ways
-to construct and interpret the version, but most importantly for us, a
-particular software version run with the same parameters should give
-identical results no matter when it's run. Results include both correct
-output as well as any errors.
-Every time you release your software, i.e.~distribute it to
-someone other than yourself and/or the development team, you should
-increment your version number.
-
-\withurl{Semantic versioning}{http://semver.org/} is one of the most common
-types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6. The major and
-minor numbers versions are almost always provided. Changes in the major
-version number herald significant changes in the software that are not
-backwards compatible, such as changing or removing features or altering
-the primary functions of the software. Increasing the minor version
-represents incremental improvements in the software, like adding new
-features. Following the minor version number can be an arbitrary number
-of project-specific identifiers, including patches, builds and qualifiers.
-Common qualifiers include \texttt{alpha}, \texttt{beta}, and \texttt{SNAPSHOT}, for applications that are
-not yet stable or released, and \texttt{-RC} for release candidates prior
-to official release.
-
-The version of your software should be easily available, both when
-supplying \texttt{-\/-version} or \texttt{-v} on the command line as
-well as in the results. The software version should be printed to the
-same location as all of the other parameters (see Rule 4).
-
-Old versions of your software should be available to ensure that results
-are reproducible into the far future. A number of mechanisms exist for
-controlled release that range from as simple as adding an appropriate
-commit message or tag to version control, to official releases alongside
-code on Sourceforge, Bitbutcket or Github, to depositing into a
-repository like apt, yum, homebrew, CPAN, etc. Choose the method that
-best suits the number and expertise of users you anticipate.
-
 \section*{Conclusion}

### 2016-08-22 robust-software.tex
Filling in prior to asking for feedback.
1. Filled in section on making old versions available.
2. Moved section on dependencies (since it seems to fit most naturally right after the old version section).
3. Rewrote conclusion to emphasize changing the ratio between effort available and effort required.

@@
 \begin{flushleft}
-{\Large
-\textbf{Ten Simple Rules for Making Research Software More Robust}
-}
-
+{\large\textbf{Ten Simple Rules for Making Research Software More Robust}}
+\\
 {Morgan~Taschuk}$^{1,\ast}$,
@@
 \begin{center}\textbf{Abstract}\end{center}
-Software produced for research---even published software---suffers from a number of common problems
-that make it difficult or impossible to run outside the original institution
-or even the primary developer's computer.
-We present ten simple rules to make software robust enough to run anywhere
-and delight your users and collaborators.
+Software produced for research,
+published and otherwise,
+suffers from a number of common problems
+that make it difficult or impossible to run outside the original institution,
+or even off the primary developer's computer.
+We present ten simple rules to make such software robust enough to run anywhere
+and thereby delight your users and collaborators.
 \end{quote}
@@
 output as well as any errors.
-Every time you release your software, i.e.~distribute it to
+Every time you release your software, i.e., distribute it to
 someone other than yourself and/or the development team, you should
@@
 types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g.~0.2.6. The major and
+form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6. The major and
 minor numbers versions are almost always provided. Changes in the major
@@
 
-\fixme{MT}{Can you please flesh this out?}
-
-\section{Do not require root or other special privileges.}
-
-Root (also known as ``superuser'' or ``admin'') is a special account on
-a computer that has (among other things) the power to modify or delete
-system files and user accounts. Conversely, files and directories owned
-by root usually cannot be modifed by normal users.
-
-Installing or running a program with root privileges is often
-convenient, since doing so automatically bypasses all those pesky safety
-checks that might otherwise get in the user's way. However, those checks
-are there for a reason: scientific software packages may not
-intentionally be malware, but one small bug or over-eager file-matching
-expression can certainly make them behave as if they were. Outside of
-very unusual circumstances, packages should therefore not require root
-privileges to set up or use.
-
-Another reason for this rule is that users may want to try out a new
-package before installing it system-wide on a cluster. Requiring root
-privileges will frustrate such efforts, and thereby reduce uptake of the
-package. Requiring that software be installed under its own user account
-(e.g., that \texttt{packagename} be made a user, and all of the
-package's software be installed in that ``user's'' space) is similarly
-limiting, and makes side-by-side installation of multiple versions of
-the package more difficult.
-
-Developers should therefore allow packages to be installed in an
-arbitrary location, e.g., under a user's home directory in
-\texttt{\textasciitilde{}/packagename}, or in directories with standard
-names like \texttt{bin}, \texttt{lib}, and \texttt{man} under a chosen
-directory. If the first option is chosen, the user may need to modify
-her search path to include the package's executables and libraries, but
-this can (more or less) be automated, and is much less risky than
-setting things up as root.
+The most common way to do this is to include the version number in the
+name of the released package, and to link to those old versions from
+the project's home page or downloads page.
+For example,
+as of the time of writing,
+the downloads page for the Python programming language includes links to
+Versions 3.5.2, 3.5.1, 3.5.0, 3.4.5, 3.4.4, 2.7.12, and 2.7.11,
+along with a link to an archive page of even older versions.
+Such archives are most useful if these packages document their dependencies,
+which we discuss in the next section.
 
@@
 
+\section{Do not require root or other special privileges.}
+
+Root (also known as ``superuser'' or ``admin'') is a special account on
+a computer that has (among other things) the power to modify or delete
+system files and user accounts. Conversely, files and directories owned
+by root usually cannot be modifed by normal users.
+
+Installing or running a program with root privileges is often
+convenient, since doing so automatically bypasses all those pesky safety
+checks that might otherwise get in the user's way. However, those checks
+are there for a reason: scientific software packages may not
+intentionally be malware, but one small bug or over-eager file-matching
+expression can certainly make them behave as if they were. Outside of
+very unusual circumstances, packages should therefore not require root
+privileges to set up or use.
+
+Another reason for this rule is that users may want to try out a new
+package before installing it system-wide on a cluster. Requiring root
+privileges will frustrate such efforts, and thereby reduce uptake of the
+package. Requiring that software be installed under its own user account
+(e.g., that \texttt{packagename} be made a user, and all of the
+package's software be installed in that ``user's'' space) is similarly
+limiting, and makes side-by-side installation of multiple versions of
+the package more difficult.
+
+Developers should therefore allow packages to be installed in an
+arbitrary location, e.g., under a user's home directory in
+\texttt{\textasciitilde{}/packagename}, or in directories with standard
+names like \texttt{bin}, \texttt{lib}, and \texttt{man} under a chosen
+directory. If the first option is chosen, the user may need to modify
+her search path to include the package's executables and libraries, but
+this can (more or less) be automated, and is much less risky than
+setting things up as root.
+
 \section{Eliminate hard-coded paths.}
@@
 The full list of useful parameters is software-specific and so cannot be
-detailed here, but here is a short list:
-
-\begin{itemize}
-\item
-  Input and reference files/directories
-\item
-  Output files/directories
-\item
-  Filtering
-\item
-  Tuning (e.g.~alphas and gammas)
-\item
-  Seeds
-\item
-  Any alternatives that you've built-in, e.g.~compress results, use a
-  different calculation, verbose output, etc.
-\end{itemize}
+detailed here, but a short list includes
+input and reference files and directories,
+output files and directories,
+filtering,
+tuning,
+random number generation seeds,
+and
+any alternatives that you've built-in,
+e.g.,
+compress results, use a different calculation, verbose output, etc.
 
@@
 
-\fixme{GW}{How to tell if the software is working:}
-
-\begin{itemize}
-\item
-  Test in a `vanilla' environment, such as another user's computer or a
-  dummy account with none of the settings of the original developer.
-\item
-  Test with different sizes of data: ridiculously small, small, medium,
-  and large. The software should run on all sizes given some parameter
-  tweaking, or fail with a sensible error message if the data is too
-  big.
-\item
-  Compare results from multiple iterations that have the same parameters
-  and inputs.(See rule \#8.)
-\item
-  Consider a container - but most of these rules still apply
-\end{itemize}
+There has been extended discussion over the past few years of the
+sustainability of research software, but this question is meaningless
+in isolation: any piece of software can be sustained if its users are
+willing to put in enough effort.  The real equation is the ratio
+between the skill and effort available, and the ease with which
+software can be installed, understood, used, maintained, and extended.
+Following the ten rules we outline here reduce the denominator, and
+thereby enable researchers to build on each other's work more easily.
 

### 2016-08-25 robust-software.tex
Re-wrote introducation and added references

@@
 or even off the primary developer's computer.
-We present ten simple rules to make such software robust enough to run anywhere
+We present ten simple rules to make such software robust enough to run anywhere,
+and inspire confidence in your reproducibility,
 and thereby delight your users and collaborators.
@@
 
-As a relatively young field, bioinformatics is full of newly developed software.
-\fixme{MT}{examples here}
-Efforts such as
-\withurl{the ELIXIR tools and data registry}{http://dx.doi.org/10.1093/nar/gkv1116}
-and the
-\withurl{Bioinformatics Links Directory}{https://bioinformatics.ca/links\_directory/}
-\cite{brazas2012}
-have made efforts towards cataloguing it: as of May 2016, the former
-has 2500 entries and the latter 1700. Those numbers constantly
-grow as both trainees and experienced practitioners produce new
-software to support their research.
-
-Typically, this software is developed by one person and 
-produces excellent results in their hands. But what happens after the
-student leaves someone else from their group wants to use the program? Everyone with
+Typically, scientific software is developed and used by a single person,
+usually a graduate student or postdoc, and produces excellent results in their
+hands\cite{prins2015}. 
+But what happens
+when someone else wants to use their software? Everyone with
 a few years of experience feels a tremor of fear when told to use
@@
 or simply be an older version than was used in published papers.
-Its new user is often faced with two unpalatable options:
-hack the existing code to make it work, or start over.
+The new user is then faced with two unpalatable options:
+hack the existing code to make it work, or start over. Being unable to
+replicate results is so common that one publication refers to it as "a rite of
+passage"\cite{baker2016}.
 
@@
 robust is the difference between ``works for me on my machine'' and
-``works for other people on a cluster I've never used''. In particular,
-robust software:
+``works for other people on a cluster I've never used''. With a lack of
+robustness also comes a lack of reproducibility and many duplicated efforts,
+slowing the pace of research\cite{prabhu2011,lawlor2015}.
+Bioinformatics is full of such efforts, catalogued by efforts such as
+the ELIXIR tools and data registry\cite{ison2016}
+and the Bioinformatics Links Directory \cite{brazas2012}. In the Bioinformatics
+Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
+to analyze transcript expression, and 182 pathway and interaction resources.
+
+This problem is not unique to bioinformatics, or even to computing\cite{baker2016}. 
+Best practices in software engineering specifically aim to increase software
+robustness. However, most bioinformaticians come from related disciplines and
+are trained on-the-job or otherwise informally\cite{prins2015,atwood2015}. Even
+the existing programs and initiatives rarely have the time to cover engineering
+topics in-depth, especially since the field is so broad and rapidly
+developing\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
+and knowledge are taught, developing robust software is not directly rewarded
+in science and funding is difficult to come by\cite{prins2015}. Some proposed
+solutions to this problem, in addition to restructuring educational programs,
+include hiring dedicated software engineers \cite{lawlor2015,sanders2008},
+partnering with private sector or grassroots organizations\cite{prins2015,ison2016},
+or using specific technical tools like containerization or cloud
+computing\cite{afgan2016,howe2012}. Each of these requires time and, in some
+cases, institutional change.
+
+You don't need to be a full-time programmer to write robust software. In fact,
+some of the best, most reliable pieces of software in many scientific
+communities are written by researchers \cite{prabhu2011,sanders2008}. 
+The most widely-used software packages tend to adopt strong software
+engineering approaches, have high standards of reproducibility, good testing
+practices, and inter-institutional collaborations that grow into strong user
+bases. In the bioinformatics community, Bioconductor and Galaxy follow this
+path\cite{Gentleman2004,afgan2016}.
+
+That said, not \emph{every} coding effort requires such rigor:
+as the saying goes, not everything worth doing is worth doing right (or right away).
+Code that is used once to answer a specific question related to a specific dataset
+doesn't require comprehensive documentation or flexible configuration,
+and the only sensible way to test it may well be to run it on the dataset in
+question. Exploratory analysis is an iterative process that is developed
+quick and revised often\cite{lawlor2015,sanders2008}. 
+However, if a script is dusted off and run three or four
+times for slightly different purposes,
+is crucial to a publication or a lab,
+or being passed on to someone else,
+it may be time to make your software more robust.
+In particular, robust software:
 
@@
 
-These rules need not be applied to \emph{every} coding effort:
-as the saying goes, not everything worth doing is worth doing right (or right away).
-Code  that is used once to answer a specific question related to a specific dataset
-doesn't require comprehensive documentation,
-and the only sensible way to test it may well be to run it on the dataset in question.
-However, if a script is dusted off and run three or four
-times for slightly different purposes,
-crucial to a publication,
-or being passed on to someone else,
-it may be time to apply ten simple rules for robust software.
-
 These simple rules are also necessary steps toward creating a reusable
-library that can be shared through a site like CPAN or CRAN.  They are
+library that can be shared and reused through a site like CPAN or CRAN.  They are
 generic and can be applied to all languages, libraries, packages,
 documentation styles, and operating systems for both closed-source and
-open-source sofwtare.  Whether the aim is as simple as sharing the
+open-source software.  
+Whether the aim is as simple as sharing the
 code with collaborators or as complex as using the software in a

### 2016-08-30 robust-software.tex
Formatting with PLOS template

@@
-\documentclass[10pt]{article}
-\bibliographystyle{plos2009}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% enumerate package lets us use letters instead of numbers
+\usepackage{enumerate}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
 \makeatletter
@@
 
+% Leave date blank
 \date{}
 
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
 \pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
 
+%% Include all macros below
 \newcommand{\fixme}[2]{\textsc{\textbf{{#1}: {#2}}}}
@@
 \begin{document}
+\vspace*{0.2in}
 
 \begin{flushleft}
-{\large\textbf{Ten Simple Rules for Making Research Software More Robust}}
+{\Large
+\textbf\newline{Ten Simple Rules for Making Research Software More Robust}
+}
+\newline
 \\
-{Morgan~Taschuk}$^{1,\ast}$,
-{Greg~Wilson}$^{2}$
+{Morgan~Taschuk}\textsuperscript{1,\ddag *},
+{Greg~Wilson}\textsuperscript{2}
 \\
-1) Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
+\textbf{1} Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
 \\
-2) Software Carpentry Foundation / gvwilson@software-carpentry.org
+\textbf{2} Software Carpentry Foundation / gvwilson@software-carpentry.org
 \\
-$\ast$ Corresponding author.
+\bigskip
+{\ddag} These authors contributed equally to this work.
+\\
+* Corresponding author.
 \end{flushleft}
 
-\begin{quote}
-\begin{center}\textbf{Abstract}\end{center}
+\section*{Abstract}
+
 Software produced for research,
@@
 and thereby delight your users and collaborators.
-\end{quote}
+
+\section*{Author Summary}
+
+\fixme{GW}{Write non-technical paragraph for the general public.}
+
+\linenumbers
 

### 2016-08-31 robust-software.tex
Prettier home page

@@
 usually a graduate student or postdoc, and produces excellent results in their
-hands\cite{prins2015}. 
+hands~\cite{prins2015}. 
 But what happens
@@
 replicate results is so common that one publication refers to it as "a rite of
-passage"\cite{baker2016}.
+passage"~\cite{baker2016}.
 
@@
 robustness also comes a lack of reproducibility and many duplicated efforts,
-slowing the pace of research\cite{prabhu2011,lawlor2015}.
+slowing the pace of research~\cite{prabhu2011,lawlor2015}.
 Bioinformatics is full of such efforts, catalogued by efforts such as
-the ELIXIR tools and data registry\cite{ison2016}
-and the Bioinformatics Links Directory \cite{brazas2012}. In the Bioinformatics
+the ELIXIR tools and data registry~\cite{ison2016}
+and the Bioinformatics Links Directory~\cite{brazas2012}. In the Bioinformatics
 Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
@@
 
-This problem is not unique to bioinformatics, or even to computing\cite{baker2016}. 
+This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}. 
 Best practices in software engineering specifically aim to increase software
 robustness. However, most bioinformaticians come from related disciplines and
-are trained on-the-job or otherwise informally\cite{prins2015,atwood2015}. Even
+are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
 the existing programs and initiatives rarely have the time to cover engineering
 topics in-depth, especially since the field is so broad and rapidly
-developing\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
+developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
 and knowledge are taught, developing robust software is not directly rewarded
-in science and funding is difficult to come by\cite{prins2015}. Some proposed
+in science and funding is difficult to come by~\cite{prins2015}. Some proposed
 solutions to this problem, in addition to restructuring educational programs,
-include hiring dedicated software engineers \cite{lawlor2015,sanders2008},
-partnering with private sector or grassroots organizations\cite{prins2015,ison2016},
+include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
+partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
 or using specific technical tools like containerization or cloud
-computing\cite{afgan2016,howe2012}. Each of these requires time and, in some
+computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
 cases, institutional change.
@@
 some of the best, most reliable pieces of software in many scientific
-communities are written by researchers \cite{prabhu2011,sanders2008}. 
+communities are written by researchers~\cite{prabhu2011,sanders2008}. 
 The most widely-used software packages tend to adopt strong software
@@
 bases. In the bioinformatics community, Bioconductor and Galaxy follow this
-path\cite{Gentleman2004,afgan2016}.
+path~\cite{gentleman2004,afgan2016}.
 
@@
 question. Exploratory analysis is an iterative process that is developed
-quick and revised often\cite{lawlor2015,sanders2008}. 
+quick and revised often~\cite{lawlor2015,sanders2008}. 
 However, if a script is dusted off and run three or four

### 2016-09-29 robust-software.tex
Itemizing feedback

@@
 
+\section*{Notes}
+
+\begin{itemize}
+
+\item
+Mention that these rules are necessary but not sufficient for software that needs to interact with services
+such as database and web servers.
+Such software is often stopped and started under automatic control,
+may need to authenticate,
+etc.,
+all of which are out of scope of this paper.
+
+\item
+Mention that reusing software introduces dependencies,
+which can bring its own special pain.
+
+\item
+``Write documentation'' is not one of our ten rules.
+Explain in the conclusion that we think this is really important,
+but it's not the software per se.
+We also don't think people are unlikely do it during normal development,
+and we don't want to recommend practices that people won't actually adopt.
+
+\item
+Find somewhere to say, ``Check that all input values are in a reasonable range,
+choose reasonable defaults where possible,
+and no defaults at all otherwise.''
+
+\item
+Explain somewhere that there's an 11th rule for compiled languages:
+use standard build tools like Make
+rather than home-brewed scripts.
+
+\item
+Remind people that a program should print its version number in all of its output,
+particularly debugging traces.
+
+\item
+Recommend that people \emph{always} install dependencies, even on their own machine,
+using their equivalent of \texttt{requirements.txt}
+so that they're sure it works.
+
+\item
+Point out that even minor changes to code can cause minor changes in output
+because of floating-point issues,
+which means that getting exactly the same output for the same input and parameters
+probably won't work during development
+(but should still be a goal for people who've deployed).
+
+\item
+Point out that having tests of any kind makes it simpler for outsiders to contribute.
+
+\item
+When we tell people not to require root privileges,
+we should point out that testing in a virtual machine is a good way to check for this.
+
+\item
+References:
+  \begin{itemize}
+  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-2-15
+  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-3-31
+  \item https://tspace.library.utoronto.ca/handle/1807/73111
+  \end{itemize}
+
+\item
+Cite http://hmmer.org/ as an example of scientific software done right.
+
+\item
+Suggest that people put all the checking and validation up front,
+so that if the program is going to crash it does so right away
+(rather than in the middle of the night after 18 hours of runtime).
+
+\item
+Should we tell people to create several single-purpose tools
+rather than One Script to Rule Them All?
+
+\item
+Should we tone down the ``tremor of fear'' remark?
+
+\item
+``Keep in mind that if a parameter can be adjusted,
+users will want to be able to turn off the feature entirely to have a baseline comparison.''
+What's the relationship here between parameter and feature?
+
+\item
+Honor local conventions,
+e.g.,
+use \texttt{TMPDIR} for temporary files.
+(The problem is,
+where are all of these written down?)
+
+\item
+Combine versioning and making previous versions available into a single point
+in order to free up space for something else to be rule 10?
+
+\end{itemize}
+
 \section*{Author Summary}
@@
 Typically, scientific software is developed and used by a single person,
-usually a graduate student or postdoc, and produces excellent results in their
+usually a graduate student or postdoc, and produces intended results in their
 hands~\cite{prins2015}. 
@@
 are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
-the existing programs and initiatives rarely have the time to cover engineering
-topics in-depth, especially since the field is so broad and rapidly
+the existing programs and initiatives rarely have the time to cover software engineering
+in-depth, especially since the field is so broad and rapidly
 developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
@@
 
-An example of good usage is the Unix \texttt{mkdir} command, which makes
+An example of good usage is GNU's \texttt{mkdir} command, which makes
 new directories:
@@
 
-Old versions of your software should be available to ensure that results
-are reproducible into the far future. A number of mechanisms exist for
-controlled release that range from as simple as adding an appropriate
-commit message or tag to version control, to official releases alongside
-code on Sourceforge, Bitbutcket or Github, to depositing into a
-repository like apt, yum, homebrew, CPAN, etc. Choose the method that
-best suits the number and expertise of users you anticipate.
-
 \section{Make older versions available.}
@@
 and won't see any benefits from upgrading.
-Programs' authors should therefore ensure that every version they have released
+A program's authors should therefore ensure that every version they have released
 continues to be available.
+
+A number of mechanisms exist for
+controlled release that range from as simple as adding an appropriate
+commit message or tag to version control, to official releases alongside
+code on Sourceforge, Bitbucket or GitHub, to depositing into a
+repository like apt, yum, homebrew, CPAN, etc. Choose the method that
+best suits the number and expertise of users you anticipate.
 
@@
 With a few lines, a call
-is made out to the other program, the results are incorporated into the
+is made out to the other program and the results are incorporated into the
 primary script. Using popular projects reduces the amount of code that
-needs to be maintained and adds the strength of vetted software to the
-final program.
+needs to be maintained and leverages the work done to check that software.
 
@@
 
-The way that the second program is invoked can throw up errors.
+The way the second program is invoked can throw up errors.
 The program may not be in the user's path, or it may be an
@@
 Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
-command, and be sure to capture and report the standard error output of
-the subprocess to facilitate debugging.
+command, and be sure to capture and report the output of the subprocess's standard error
+to facilitate debugging.
 
@@
 
-Many bioinformatics applications rely on randomized algorithms to
+Many applications rely on randomized algorithms to
 improve performance or runtimes. As a consequence, results can change

### 2016-10-05 robust-software.tex
Rules are necessary but not sufficient for services

@@
 \begin{itemize}
-
-\item
-Mention that these rules are necessary but not sufficient for software that needs to interact with services
-such as database and web servers.
-Such software is often stopped and started under automatic control,
-may need to authenticate,
-etc.,
-all of which are out of scope of this paper.
 
@@
 
+Any short list of rules for developing robust software will
+necessarily be incomplete.  For example, the rules we present above
+are necessary, but not sufficient, for software that needs to interact
+with services such as database and web servers.  Such software is
+often stopped and started under automatic control, may need to
+authenticate, etc., all of which are out of scope of this paper.
+
 There has been extended discussion over the past few years of the

### 2016-10-05 robust-software.tex
Reuse introduces dependencies

@@
 \begin{itemize}
-
-\item
-Mention that reusing software introduces dependencies,
-which can bring its own special pain.
 
@@
 
-Unfortunately, the interface between two software packages can be a
-source of considerable frustration. Support requests descend into
-debugging errors produced by the other project.
+Unfortunately, reusing software introduces dependencies, which can
+bring their own special pain.  The interface between two software
+packages can also be a source of considerable frustration: all too
+often, support requests descend into debugging errors produced by the
+other project.
 

### 2016-10-05 robust-software.tex
Don't recommend writing docs because...

@@
 \begin{itemize}
-
-\item
-``Write documentation'' is not one of our ten rules.
-Explain in the conclusion that we think this is really important,
-but it's not the software per se.
-We also don't think people are unlikely do it during normal development,
-and we don't want to recommend practices that people won't actually adopt.
 
@@
 
+Another notable omission is documentation.
+We think it is very important for developers to do this,
+but our experience is that people are unlikely do it during normal development,
+and we don't want to recommend practices that people won't actually adopt.
+
 There has been extended discussion over the past few years of the

### 2016-10-05 robust-software.tex
Setting defaults

@@
 \begin{itemize}
-
-\item
-Find somewhere to say, ``Check that all input values are in a reasonable range,
-choose reasonable defaults where possible,
-and no defaults at all otherwise.''
 
@@
 
+Finally,
+if the software is configurable in any way,
+it should \textbf{check that all input values are in a reasonable range},
+\textbf{choose reasonable defaults where they exist},
+and \textbf{set no defaults at all when there aren't any reasonable ones}.
+Software should also \textbf{check that configuration values are reasonable at startup}:
+few things are as annoying as having a program announce after running for two hours
+that it isn't going to save its results
+because the requested directory doesn't exist.
+
 \section{Include a small test set that can be run to ensure the software is actually working.}

### 2016-10-05 robust-software.tex
Bolding a few subpoints

@@
 expression can certainly make them behave as if they were. Outside of
-very unusual circumstances, packages should therefore not require root
-privileges to set up or use.
+very unusual circumstances,
+\textbf{packages should not require root privileges to set up or use}.
 
@@
 
-Developers should therefore allow packages to be installed in an
-arbitrary location, e.g., under a user's home directory in
+Developers should therefore
+\textbf{allow packages to be installed in an arbitrary location},
+e.g., under a user's home directory in
 \texttt{\textasciitilde{}/packagename}, or in directories with standard
@@
 
-For these reasons, users should be able to set the names and locations
-of input and output files as command-line parameters. This rule applies
-to reference data sets as well as the user's own data: if a user wants
-to try a new gene identification algorithm using a different set of
-genes as a training set, she should not have to edit the software to do
-so.
+For these reasons, users should be able to
+\textbf{set the names and locations of input and output files as command-line parameters}.
+This rule applies to reference data sets as well as the user's own
+data: if a user wants to try a new gene identification algorithm using
+a different set of genes as a training set, she should not have to
+edit the software to do so.
 
-A corollary to this rule is that a package should not require users to
-navigate to a particular directory to do their work. ``Where I have to
-be'' is just another hard-coded path.
+A corollary to this rule is
+\textbf{do not require users to navigate to a particular directory to do their work}.
+``Where I have to be'' is just another hard-coded path.
 

### 2016-10-05 robust-software.tex
creating section for what we left out

@@
 \begin{itemize}
-
-\item
-Explain somewhere that there's an 11th rule for compiled languages:
-use standard build tools like Make
-rather than home-brewed scripts.
 
@@
 
-\section*{Conclusion}
+\section*{What We Left Out}
 
 Any short list of rules for developing robust software will
-necessarily be incomplete.  For example, the rules we present above
-are necessary, but not sufficient, for software that needs to interact
-with services such as database and web servers.  Such software is
-often stopped and started under automatic control, may need to
-authenticate, etc., all of which are out of scope of this paper.
+necessarily be incomplete.  A few of our more significant omissions
+include:
 
-Another notable omission is documentation.
-We think it is very important for developers to do this,
-but our experience is that people are unlikely do it during normal development,
-and we don't want to recommend practices that people won't actually adopt.
+\begin{description}
+
+\item[\textbf{Services}] The rules we present above are necessary, but
+  not sufficient, for software that needs to interact with services
+  such as database and web servers.  Such software is often stopped
+  and started under automatic control, may need to authenticate, etc.,
+  all of which are out of scope of this paper.
+
+\item[\textbf{Documentation}] We think it is very important for
+  developers to document their work, but our experience is that people
+  are unlikely do it during normal development, and we don't want to
+  recommend practices that people won't actually adopt.
+
+\item[\textbf{Build Systems}] If the software depends on a build tool,
+  either because it is written in a compiled language or to manage
+  workflow, it should use a standard tool such as Make rather than
+  home-brewed scripts.
+
+\end{description}
+
+\section*{Conclusion}
 

### 2016-10-05 robust-software.tex
Print the version number

@@
 \begin{itemize}
-
-\item
-Remind people that a program should print its version number in all of its output,
-particularly debugging traces.
 
@@
 well as in the results. The software version should be printed to the
-same location as all of the other parameters (see Rule 4).
+same location as all of the other parameters (see Rule 4).  It should
+also be included in all of the program's output, particularly debugging
+traces: if someone needs help, it's important that they be able to tell
+whoever's helping them which version of the software they're using.
 

### 2016-10-05 robust-software.tex
Installing dependencies

@@
 \begin{itemize}
-
-\item
-Recommend that people \emph{always} install dependencies, even on their own machine,
-using their equivalent of \texttt{requirements.txt}
-so that they're sure it works.
 
@@
 required software is available, and install it if it is not. Similar
-mechanisms exist for Perl, R, and other languages.
+mechanisms exist for Perl, R, and other languages.  Whatever is used,
+we recommend that developers should \emph{always} install dependencies
+using this mechanism, especially on their personal machines, so that
+they're sure it works.
 

### 2016-10-05 robust-software.tex
Floating point issues

@@
 \begin{itemize}
-
-\item
-Point out that even minor changes to code can cause minor changes in output
-because of floating-point issues,
-which means that getting exactly the same output for the same input and parameters
-probably won't work during development
-(but should still be a goal for people who've deployed).
 
@@
 
-Given a set of parameters and a dataset, the package should produce the
-same results every time it is run.
+Given a set of parameters and a dataset, a particular version of a program
+should produce the same results every time it is run
+to aid testing, debugging, and reproducibility.
+Even minor changes to code can cause minor changes in output because of floating-point issues,
+which means that getting exactly the same output for the same input and parameters
+probably won't work during development,
+but it should still be a goal for people who have deployed a specific version.
 

### 2016-10-05 robust-software.tex
Adding ref to Steinmacher

@@
 \begin{itemize}
-
-\item
-Point out that having tests of any kind makes it simpler for outsiders to contribute.
 
@@
 
+Finally,
+studies have found that the ease with which people can start making
+contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
+By making it simpler for outsiders to contribute,
+a test suite of any kind also makes it more likely that they will.
+
 \section{Produce identical results when given identical inputs.}

### 2016-10-05 robust-software.tex
Testing installation using Docker

@@
 \begin{itemize}
-
-\item
-When we tell people not to require root privileges,
-we should point out that testing in a virtual machine is a good way to check for this.
 
@@
 
+Note:
+testing software installation has traditionally been regarded as difficult,
+since it necessarily alters the machine on which the test is conducted.
+Lightweight virtualization containers like Docker make this much easier as well.
+
 \section{Eliminate hard-coded paths.}

### 2016-10-06 robust-software.tex

@@
 \item
-Cite http://hmmer.org/ as an example of scientific software done right.
-
-\item
-Suggest that people put all the checking and validation up front,
-so that if the program is going to crash it does so right away
-(rather than in the middle of the night after 18 hours of runtime).
-
-\item
-Should we tell people to create several single-purpose tools
-rather than One Script to Rule Them All?
-
-\item
-Should we tone down the ``tremor of fear'' remark?
-
-\item
-``Keep in mind that if a parameter can be adjusted,
-users will want to be able to turn off the feature entirely to have a baseline comparison.''
-What's the relationship here between parameter and feature?
-
-\item
 Honor local conventions,
@@
 
-\fixme{GW}{Write non-technical paragraph for the general public.}
+Many researchers have found out the hard way that there's a world of difference
+between ``works for me on my machine'' and ``works for other people on theirs''.
+Many common challenges can be avoided by following a few simple rules; doing so
+not only accelerates research, but also improves reproducibility.
 
@@
 when someone else wants to use their software? Everyone with
-a few years of experience feels a tremor of fear when told to use
+a few years of experience feels a bit nervous when told to use
 a graduated student's code to analyze their data.

### 2016-10-07 robust-software.tex
Hopefully finalizing changes

@@
 {Morgan~Taschuk}\textsuperscript{1,\ddag *},
-{Greg~Wilson}\textsuperscript{2}
+{Greg~Wilson}\textsuperscript{2,\ddag}
 \\
@@
 Typically, scientific software is developed and used by a single person,
-usually a graduate student or postdoc, and produces intended results in their
+usually a graduate student or postdoc, and produces the intended results in their
 hands~\cite{prins2015}. 
@@
 a few years of experience feels a bit nervous when told to use
-a graduated student's code to analyze their data.
+another person's code to analyze their data.
 Often, that software will be undocumented and work in unexpected
@@
 slowing the pace of research~\cite{prabhu2011,lawlor2015}.
-Bioinformatics is full of such efforts, catalogued by efforts such as
+Bioinformatics is full of such efforts, catalogued by repositories such as
 the ELIXIR tools and data registry~\cite{ison2016}
@@
 to analyze transcript expression, and 182 pathway and interaction resources.
+Some of these tools are legitimate efforts to improve the state-of-the-art, but
+often they are difficult to install and run~\cite{stajich2002}, and are effectively abandoned
+after publication~\cite{nekrutenko2012}.
 
@@
 are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
-the existing programs and initiatives rarely have the time to cover software engineering
+the existing training programs and initiatives rarely have the time to cover software engineering
 in-depth, especially since the field is so broad and rapidly
@@
 
-That said, not \emph{every} coding effort requires such rigor:
-as the saying goes, not everything worth doing is worth doing right (or right away).
+That said, not \emph{every} coding effort needs to be engineered to last.
 Code that is used once to answer a specific question related to a specific dataset
@@
 it may be time to make your software more robust.
+
 In particular, robust software:
@@
 At a minimum, it needs to get a new user started and point them towards more
-help, if they need it.
-A working example using test data (per
-\#9) is always appreciated.
+help, if they need it. Numerous guidelines exist on how to write good
+READMEs\cite{Johnson1997,gnustandards}. We summarize the crucial points here.
 
@@
 
-\textbf{List required dependencies:} Often, software depends on very
-specific versions of libraries, modules, or operating systems. This is
-entirely reasonable as long as it is properly documented. Often,
-multiple libraries exist with the same or very similar names, so
-either provide the commands necessary to download the dependencies or
-link to the software homepage. Include the version number for each
-dependency.
-
-If your dependency is not publicly available,
-you have several options to ensure your software remains useable.
-You can add plain-text or code directly to
-your repository with appropriate attribution. If the dependency is a binary, use a
-\withurl{binary repository manager}{https://en.wikipedia.org/wiki/Binary\_repository\_manager}
-such as Artifactory or ProGet. These managers keep versioned copies of
-software at constant URLs so they can be downloaded as long as the
-manager continues to run. If the dependency must remain closed, place it at an
-internal location on shared disk, remove all write permissions, and
-link to it from your README, although this method is 
-discouraged because of the potential for sharing and risks accidental removal.
+\textbf{List required dependencies:} Often, software depends on
+specific versions of libraries, modules, or operating systems. 
+This is
+entirely reasonable as long as it is properly documented. Include the name and
+version number for each dependency.  Several methods exist for acquiring
+required packages: either provide the commands necessary to download the
+dependencies; link to the software homepage; or (our preference) use a
+dependency manager. We address dependencies in more detail in rule
+\textcolor{red}{5}.
 
@@
 your main contributors; licenses are how you want others to use and
-credit your software. By putting them in the README, it is readable \emph{before}
+credit your software. READMEs are usualy available \emph{before}
 the software is installed (or even downloaded).
@@
 and includes links to a CITATION file (explaining how to cite the project)
-and the license.
+and the license file.
 
@@
 
+Almost all command-line applications use a combination of
+POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for usage. 
 An example of good usage is GNU's \texttt{mkdir} command, which makes
@@
 \end{small}
-
-There is no standard format for usage statements, but good ones share
-several features:
 
@@
 
-\section{Give the software a meaningful version number.}
+\section{Version your releases}
 
+Software evolves over time, with developers adding or removing features as need
+dictates. Making official releases stamps a particular set of features with a
+project-specific identifier so that version can be retrieved for later use. For
+example, if a paper is published, the software should be released at the same
+time so that the results can be reproduced. 
 Most software has a version number composed of a decimal number that
-increments as new versions are released. There are many different ways
+increments as new versions are released.
+
+There are many different ways
 to construct and interpret the version, but most importantly for us, a
@@
 output as well as any errors.
-Every time you release your software, i.e., distribute it to
-someone other than yourself and/or the development team, you should
-increment your version number.
+\textbf{Increment your version number every time you release your software to
+other people}.
 
@@
 types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6. The major and
-minor numbers versions are almost always provided. Changes in the major
+form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6.
+Changes in the major
 version number herald significant changes in the software that are not
@@
 not yet stable or released, and \texttt{-RC} for release candidates prior
-to official release.
+to an official release.
 
-The version of your software should be easily available, both when
-supplying \texttt{-\/-version} or \texttt{-v} on the command line as
-well as in the results. The software version should be printed to the
-same location as all of the other parameters (see Rule 4).  It should
+The version of your software should be easily available.
+Supplying \texttt{-\/-version} or \texttt{-v} on the command line should print
+the software name and version number, and it should
 also be included in all of the program's output, particularly debugging
@@
 
-\section{Make older versions available.}
-
-Software evolves, but not all software evolves at the same pace.
-While more consistent command-line arguments
-or more flexible handling of output directories
-may make a program better in general,
-it can simultaneously make work for someone
+While new releases may make a program better in general,
+they can simultaneously create work for someone
 who integrated the old version into their own workflow a year or two ago,
@@
 A program's authors should therefore ensure that every version they have released
-continues to be available.
-
-A number of mechanisms exist for
+continues to be available. A number of mechanisms exist for
 controlled release that range from as simple as adding an appropriate
@@
 best suits the number and expertise of users you anticipate.
-
-The most common way to do this is to include the version number in the
-name of the released package, and to link to those old versions from
-the project's home page or downloads page.
-For example,
-as of the time of writing,
-the downloads page for the Python programming language includes links to
-Versions 3.5.2, 3.5.1, 3.5.0, 3.4.5, 3.4.4, 2.7.12, and 2.7.11,
-along with a link to an archive page of even older versions.
-Such archives are most useful if these packages document their dependencies,
-which we discuss in the next section.
 
@@
 primary script. Using popular projects reduces the amount of code that
-needs to be maintained and leverages the work done to check that software.
+needs to be maintained and leverages the work done by the other software.
 
@@
 often, support requests descend into debugging errors produced by the
-other project.
-
-In addition, every package someone has to install
+other project. In addition, every package someone has to install
 before being able to use yours is a
 possible (some would say ``likely'') source of frustration for some
-potential user. On the other hand, research software developers should
-re-use existing software wherever possible.
+potential user. The way the second program is invoked can throw up errors.
+The program may not be in the user's path, or it may be an
+older or newer version and produce results different than expected.
+Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
+use shell-specific conventions like `*' expansion.
+Even Linux-standard functions available vary slightly between
+installs. For example, GNU \texttt{sort} is available on almost every
+*nix distribution, but sorts differently depending on locale.
+
+Despite these problems, software developers in research should
+re-use existing software wherever possible and reasonable to avoid
+``reinventing the wheel''. We fully support the reuse of software between
+projects provided three additional guidelines are adhered to.
+
+First, 
+\textbf{make sure that you really need the auxiliary program}. If you are
+executing GNU sort instead of figuring out how to sort lists in Python,
+it may not be worth the pain of integration.
+
+Second, \textbf{ensure the appropriate software and version is available}.
+Early in execution, check whether the dependency is executable and whether the
+version is compatible.
+Either allow the user to
+configure the exact path to the package, distribute the program with the
+dependent software, or download it during installation using a
+dependency management system (see Rule \textcolor{red}{RULENO}).
+
+Finally, to ensure support on as many different operating systems as
+possible, \textbf{use native functions for starting other processes}, such as
+Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
+command, and be sure to capture and report the output of the subprocess's standard error
+to facilitate debugging.
+
+\section{Use a build system and dependency manager.}
+
+If the software depends on a build tool, either because it is written in a
+compiled language or to manage a workflow, it should use a standard tool such
+as Make rather than home-brewed scripts.
 
@@
 
-The way the second program is invoked can throw up errors.
-The program may not be in the user's path, or it may be an
-older or newer version and produce results different than expected.
-Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
-use shell-specific conventions like `*' expansion.
-Even Linux-standard functions available vary slightly between
-installs. For example, GNU \texttt{sort} is available on almost every
-*nix distribution, but sorts differently depending on locale.
-
-Despite these caveats,
-we fully support the reuse of software between projects provided
-three additional guidelines are adhered to. First, 
-\textbf{make sure that you really need the auxiliary program}. If you are
-executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration.
-
-Second, \textbf{ensure the appropriate software and version is available}.
-Either allow the user to
-configure the exact path to the package, distribute the program with the
-dependent software, or download it during installation using a
-dependency management system. Regardless, check whether the program is
-executable and what version is running. Be sure to remind users in the
-documentation what the compatible versions are.
-
-Finally, to ensure support on as many different operating systems as
-possible, \textbf{use native functions for starting other processes}, such as
-Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
-command, and be sure to capture and report the output of the subprocess's standard error
-to facilitate debugging.
-
 \section{Do not require root or other special privileges.}
@@
 (e.g., that \texttt{packagename} be made a user, and all of the
-package's software be installed in that ``user's'' space) is similarly
+package's software be installed in that ``user's'' space (see: Apache Tomcat))
+is similarly
 limiting, and makes side-by-side installation of multiple versions of
@@
 
-Note:
-testing software installation has traditionally been regarded as difficult,
+Testing the ability to install software has traditionally been regarded as difficult,
 since it necessarily alters the machine on which the test is conducted.
-Lightweight virtualization containers like Docker make this much easier as well.
+Lightweight virtualization containers like Docker make this much easier as
+well, or simply asking another person to try and build your software.
 
@@
 The full list of useful parameters is software-specific and so cannot be
-detailed here, but a short list includes
+detailed here, but a short list includes providing
 input and reference files and directories,
@@
 
+\textbf{Allow the user to configure useful parameters on the command line.}
 Providing parameters on the
@@
 they change the results is important as your software gains more users,
-facilitating exploratory analysis and parameter sweeping. Keep in mind
-that if a parameter can be adjusted, users will want to be able to turn
-off the feature entirely to have a baseline comparison.
+facilitating exploratory analysis and parameter sweeping.
 
-When the software starts, it should echo all parameters and software
-versions to standard out or a log file alongside the results. This
+When the software starts, it should \textbf{echo all parameters and software
+versions to standard out or a log file alongside the results}. This
 feature supports greater reproducibility because any result can be
@@
 provided on the command line as an additional argument.
-
-We caution against overusing configuration files. When a user needs to
-locate, open, change and save a file in order to change a parameter, the
-import of the change seems larger and discourages experimentation. 
 Only values that are unlikely to
@@
 cleaned up after execution completes or they occupy valuable disk space.
-Use a default configuration file and specify all other parameter values on the
+Use a default configuration file and specify all other parameters on the
 command line.
@@
 Every package should come with a small test script for users to run
-after installation. Its purpose is \emph{not} to check that the software
+after installation. Its purpose is not only to check that the software
 is working correctly (although that is extremely helpful), but rather to
@@
 
-In order to be useful, this test script must be easy to find and run. A
+In order to be useful, this test script must be easy to find and run. 
+Many build systems will also run unit tests if provided them at compile time.
+If the build system is not amenable to testing, a 
 single file in the project's root directory named \texttt{runtests.sh}
@@
 By making it simpler for outsiders to contribute,
-a test suite of any kind also makes it more likely that they will.
+a test suite of any kind also makes it more likely that they will, and software
+with collaborators stands a better chance of surviving in the busy field of
+scientific software.
 
@@
 between runs, even when provided with the same data and parameters. By
-its nature, this randomness renders strict reproducibility impossible.
-Debugging is more difficult. If even the small test set (\#9) produces
+its nature, this randomness renders strict reproducibility and therefore
+debugging more difficult. If even the small test set (\#9) produces
 different results, new users may not be able to tell whether the software is
@@
 approximate random numbers. Setting the seed to a consistent value
-removes randomness between runs. Allow the user to optionally provide
+can remove randomness between runs. Allow the user to optionally provide
 the seed as an input parameter, thus rendering the program deterministic
@@
 using clock time), echo it to the output for re-use later. 
+If setting the seed is not possible, make sure the acceptable tolerance is
+known and detailed in documentation and in the tests.
 
@@
   are unlikely do it during normal development, and we don't want to
-  recommend practices that people won't actually adopt.
+  recommend practices that people won't actually adopt. However, it is worth
+  noting that software that is widely used and contributed to has and enforces
+  the need for good documentation~\cite{gentleman2004}.
 
-\item[\textbf{Build Systems}] If the software depends on a build tool,
-  either because it is written in a compiled language or to manage
-  workflow, it should use a standard tool such as Make rather than
-  home-brewed scripts.
-
+\item[\textbf{Build Systems}] 
 \end{description}

### 2016-10-11 robust-software.tex
fixed some spacing and leftover elements
Smoothed out words, added build utility intro, added two more suggested references, spell-checked

@@
 
-\begin{itemize}
-
-\item
-References:
-  \begin{itemize}
-  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-2-15
-  \item https://gigascience.biomedcentral.com/articles/10.1186/2047-217X-3-31
-  \item https://tspace.library.utoronto.ca/handle/1807/73111
-  \end{itemize}
-
-\item
-Honor local conventions,
-e.g.,
-use \texttt{TMPDIR} for temporary files.
-(The problem is,
-where are all of these written down?)
-
-\item
-Combine versioning and making previous versions available into a single point
-in order to free up space for something else to be rule 10?
-
-\end{itemize}
-
 \section*{Author Summary}
@@
 robust is the difference between ``works for me on my machine'' and
-``works for other people on a cluster I've never used''. With a lack of
-robustness also comes a lack of reproducibility and many duplicated efforts,
+``works for other people on a cluster I've never used''. The lack of
+robustness also leads to many duplicated efforts,
 slowing the pace of research~\cite{prabhu2011,lawlor2015}.
-Bioinformatics is full of such efforts, catalogued by repositories such as
-the ELIXIR tools and data registry~\cite{ison2016}
-and the Bioinformatics Links Directory~\cite{brazas2012}. In the Bioinformatics
+Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
+hundreds of tools to perform the same task. In the Bioinformatics
 Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
@@
 Some of these tools are legitimate efforts to improve the state-of-the-art, but
-often they are difficult to install and run~\cite{stajich2002}, and are effectively abandoned
+often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
 after publication~\cite{nekrutenko2012}.
@@
 At a minimum, it needs to get a new user started and point them towards more
-help, if they need it. Numerous guidelines exist on how to write good
+help, if they need it. 
+Numerous guidelines exist on how to write good
 READMEs\cite{Johnson1997,gnustandards}. We summarize the crucial points here.
@@
 dependencies; link to the software homepage; or (our preference) use a
-dependency manager. We address dependencies in more detail in rule
-\textcolor{red}{5}.
+dependency manager. We address dependencies in more detail in rule 5.
 
-\textbf{Installation instructions:} If the software needs to be compiled
+\textbf{Provide compile/installation instructions:} If the software needs to be compiled
 or installed, list those instructions in the README. Also
@@
 
-\textbf{Input and output files}: All possible input and output files
+\textbf{List input and output files}: All possible input and output files
 should be listed in this section. Do the files conform to a particular
@@
 Input files and their formats are included in most documentation, but
-intermediate, auxilliary and log files are often missing. 
+intermediate, auxiliary and log files are often missing.
 \emph{All} files should be listed in the README, even those considered self-explanatory. Log
@@
 
-\textbf{Attributions and licensing:} Attributions are how you credit
+\textbf{State attributions and licensing:} Attributions are how you credit
 your main contributors; licenses are how you want others to use and
-credit your software. READMEs are usualy available \emph{before}
+credit your software. READMEs are usually available \emph{before}
 the software is installed (or even downloaded).
@@
 explicitly choosing a help or usage option.
+More standard command-line behaviours are detailed in \cite{Seemann2013}.
 
@@
 
-The version of your software should be easily available.
-Supplying \texttt{-\/-version} or \texttt{-v} on the command line should print
+\textbf{The version of your software should be easily available by 
+supplying \texttt{-\/-version} or \texttt{-v} on the command line}. This command should print
 the software name and version number, and it should
-also be included in all of the program's output, particularly debugging
+also be \textbf{included in all of the program's output}, particularly debugging
 traces: if someone needs help, it's important that they be able to tell
@@
 and won't see any benefits from upgrading.
-A program's authors should therefore ensure that every version they have released
-continues to be available. A number of mechanisms exist for
+A program's authors should therefore \textbf{ensure that every released version
+continues to be available.}
+A number of mechanisms exist for
 controlled release that range from as simple as adding an appropriate
 commit message or tag to version control, to official releases alongside
-code on Sourceforge, Bitbucket or GitHub, to depositing into a
+code on Bitbucket or GitHub, to depositing into a
 repository like apt, yum, homebrew, CPAN, etc. Choose the method that
@@
 often, support requests descend into debugging errors produced by the
-other project. In addition, every package someone has to install
+other project. 
+In addition, every package someone has to install
 before being able to use yours is a
@@
 older or newer version and produce results different than expected.
-Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explictly or
+Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explicitly or
 use shell-specific conventions like `*' expansion.
@@
 dependent software, or download it during installation using a
-dependency management system (see Rule \textcolor{red}{RULENO}).
+dependency management system (see Rule 5).
 
@@
 
-\section{Use a build system and dependency manager.}
+\section{Use a build utility and package manager.}
 
-If the software depends on a build tool, either because it is written in a
-compiled language or to manage a workflow, it should use a standard tool such
-as Make rather than home-brewed scripts.
+In all but the simplest of scripts, \textbf{we highly recommend using a build
+automation utility to compile code and/or manage your software workflow}. The benefits of
+using a build utility like Make, Rake, Maven, Ant or MS Build vastly outweighs the
+drawbacks associated with the learning curve. Most importantly, a build system
+lets someone else build your software the same way you would. Most build
+utilities can also be configured to run tests during every build and
+re-generate documentation, and can be hooked into continuous integration,
+cutting releases, and deploying the software to other environments.
 
-To strike a balance between
-these two, developers should
-\textbf{document \emph{all} of the packages that theirs depends on, preferably in a machine-readable form}.
+Build utilities rely on any dependencies being present in the environment for a
+successful build, and so developers should
+\textbf{document \emph{all} dependencies, preferably in a machine-readable form}.
+Package managers like apt and yum are available on most *nix systems, and
+application package managers exist for specific languages like Python (pip),
+Java (Maven/Gradle), and Ruby (RubyGems). These package managers can be used
+together with the build utility to ensure that dependencies are available at
+compile/run time.
 For example, it is common for Python projects to include a file called
@@
 
-This file can be read by a package manager, which can check that the
-required software is available, and install it if it is not. Similar
-mechanisms exist for Perl, R, and other languages.  Whatever is used,
+This file can be read by pip package manager, which can check that the
+required software is available, and install it if it is not. 
+Whatever is used,
 we recommend that developers should \emph{always} install dependencies
@@
 Conversely, developers should
-\textbf{avoid depending on scripts and tools which are not available as packages},
-even (or especially) if they are installed on the computers the original developer is using.
+\textbf{avoid depending on scripts and tools which are not available as packages}.
 In many cases, a program's author may not realize that some tool was built locally, and
@@
 unknown dependencies is to install on a system administered by someone
-else and see what breaks. As use of lightweight
-virtualization containers like Docker becomes more widespread, it may
-become common to test installation on a virtual machine.
+else and see what breaks. As use of virtualization containers becomes more
+widespread, software installation can also be tested on a virtual machine or
+container system like Docker.
 
@@
 (e.g., that \texttt{packagename} be made a user, and all of the
-package's software be installed in that ``user's'' space (see: Apache Tomcat))
+package's software be installed in that ``user's'' space 
+(see: Apache Tomcat))
 is similarly
@@
 since it necessarily alters the machine on which the test is conducted.
-Lightweight virtualization containers like Docker make this much easier as
-well, or simply asking another person to try and build your software.
+Lightweight virtualization containers like Docker make this much easier as well, 
+or simply \textbf{ask another person to try and build your software}.
 
@@
 
-\textbf{Allow the user to configure useful parameters on the command line.}
 Providing parameters on the
@@
 facilitating exploratory analysis and parameter sweeping.
-
-When the software starts, it should \textbf{echo all parameters and software
-versions to standard out or a log file alongside the results}. This
-feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
+\textbf{Choose reasonable defaults where they exist},
+and \textbf{set no defaults at all when there aren't any reasonable ones}.
 
@@
 override those values. Configuration files should be used in preference
-to hardcoding the defaults directly. Configuration files can be in a standard
+to hard-coding the defaults directly. Configuration files can be in a standard
 location, e.g. \texttt{.packagerc} in the user's home directory or
@@
 your lab or institution. 
-Since
-all parameters should be echoed in the results, config files should be
-cleaned up after execution completes or they occupy valuable disk space.
 Use a default configuration file and specify all other parameters on the
@@
 
+When the software starts, it should \textbf{echo all parameters and software
+versions to standard out or a log file alongside the results}. This
+feature supports greater reproducibility because any result can be
+replicated with only the previous output files as reference.
 Finally,
 if the software is configurable in any way,
-it should \textbf{check that all input values are in a reasonable range},
-\textbf{choose reasonable defaults where they exist},
-and \textbf{set no defaults at all when there aren't any reasonable ones}.
-Software should also \textbf{check that configuration values are reasonable at startup}:
-few things are as annoying as having a program announce after running for two hours
+it should \textbf{check that all input values are in a reasonable range at startup}. 
+Few things are as annoying as having a program announce after running for two hours
 that it isn't going to save its results
@@
 
-In order to be useful, this test script must be easy to find and run. 
+In order to be useful, \textbf{this test script must be easy to find and run}. 
 Many build systems will also run unit tests if provided them at compile time.
@@
 
-Equally, the test script's output must be easy to interpret. Screens
+Equally, \textbf{the test script's output must be easy to interpret}. Screens
 full of correlation coefficients do not qualify: instead, the script's
@@
 
-Finally,
-studies have found that the ease with which people can start making
+Studies have found that the ease with which people can start making
 contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
@@
 
-Given a set of parameters and a dataset, a particular version of a program
-should produce the same results every time it is run
+Given a set of parameters and a dataset, \textbf{a particular version of a program
+should produce the same results every time it is run}
 to aid testing, debugging, and reproducibility.
@@
 approximate random numbers. Setting the seed to a consistent value
-can remove randomness between runs. Allow the user to optionally provide
-the seed as an input parameter, thus rendering the program deterministic
+can remove randomness between runs. \textbf{Allow the user to optionally provide
+the random seed as an input parameter}, thus rendering the program deterministic
 for those cases where it matters. If the seed is set internally (e.g.,
 using clock time), echo it to the output for re-use later. 
-If setting the seed is not possible, make sure the acceptable tolerance is
-known and detailed in documentation and in the tests.
+If setting the seed is not possible, \textbf{make sure the acceptable tolerance is
+known and detailed in documentation and in the tests}.
 
@@
   the need for good documentation~\cite{gentleman2004}.
+  Numerous resources exist for writing high quality
+  documentation~\cite{karimzadeh2016} and so we do not cover it here.
 
-\item[\textbf{Build Systems}] 
 \end{description}

### 2016-10-11 robust-checks.tex
Added cheeky checklist

--- 2016-10-11 robust-software.tex
@@
 
+\usepackage[ampersand]{easylist}
+
 \begin{document}
@@
 {\Large
-\textbf\newline{Ten Simple Rules for Making Research Software More Robust}
+\textbf\newline{Ten Simple Rules for Making Research Software More Robust: The Checklist}
 }
@@
 
-\section*{Abstract}
+\begin{easylist}[checklist]
+& Have a README
+&& Explain what the software does
+&& List required dependencies
+&& Provide compile/installation instructions
+&& List input and output files
+&& State attributions and licensing
+& Print usage information from the command line
+&& Include:
+&&& the syntax for running the program in GNU/POSIX format
+&&& a one line description
+&&& the most commonly used arguments, a description of each, and the default values
+&&&& Where to find more information
+&& Print to standard output
+&& Exit with an appropriate exit code
+& Version your releases
+&& Increment your version number every time you release your software to other people
+&& Print when supplying \texttt{-\/-version} or \texttt{-v} on the command line
+&& Include version number in output
+&& Deposit releases in a stable location so they are available in perpetuity
+& Reuse software (within reason)
+&& Make sure that you really need the auxiliary program
+&& Check for dependent software and version early in execution
+&& Use native functions for starting other processes
+& Use a build utilty and package manager
+&& Document \emph{all} dependencies, preferably in a machine-readable form
+&& Avoid depending on scripts and tools which are not available as packages
+& Do not require root or other special privileges
+&& Allow packages to be installed in an arbitrary location
+&& Ask another person to try and build your software
+& Eliminate hard-coded paths
+&& Set the names and locations of input and output files as command-line parameters
+&& Do not require users to navigate to a particular directory to do their work
+& Allow configuration of all useful parameters from the command line.
+&& Choose reasonable defaults where they exist
+&& Set no defaults at all when there aren't any reasonable ones
+&& Echo all parameters and software versions to standard out or a log file alongside the results
+&& Check that all input values are in a reasonable range near startup
+& Include a small test set that can be run to ensure the software is actually working.
+&& Tests are easy to find and run
+&& Test results are easy to interpret
+& Produce identical results when given identical inputs
+&& For randomized algorithms: allow the user to optionally provide the seed as an input parameter; or
+&& Make sure the acceptable tolerance is known and detailed in documentation and in the tests
+\end{easylist}
 
-Software produced for research,
-published and otherwise,
-suffers from a number of common problems
-that make it difficult or impossible to run outside the original institution,
-or even off the primary developer's computer.
-We present ten simple rules to make such software robust enough to run anywhere,
-and inspire confidence in your reproducibility,
-and thereby delight your users and collaborators.
+\section*{Bonus points}
 
-\section*{Notes}
+\begin{easylist}[checklist]
+& Conform to command-line conventions~\cite{Seemann2013}
+& Write high quality documentation~\cite{karimzadeh2016}
 
-\section*{Author Summary}
+\end{easylist}
 
-Many researchers have found out the hard way that there's a world of difference
-between ``works for me on my machine'' and ``works for other people on theirs''.
-Many common challenges can be avoided by following a few simple rules; doing so
-not only accelerates research, but also improves reproducibility.
-
-\linenumbers
-
-\section*{Introduction}
-
-Typically, scientific software is developed and used by a single person,
-usually a graduate student or postdoc, and produces the intended results in their
-hands~\cite{prins2015}. 
-But what happens
-when someone else wants to use their software? Everyone with
-a few years of experience feels a bit nervous when told to use
-another person's code to analyze their data.
-Often, that software will be undocumented and work in unexpected
-ways (if it works at all).  It will often rely on nonexistent paths or resources,
-be tuned for a single dataset,
-or simply be an older version than was used in published papers.
-The new user is then faced with two unpalatable options:
-hack the existing code to make it work, or start over. Being unable to
-replicate results is so common that one publication refers to it as "a rite of
-passage"~\cite{baker2016}.
-
-The root cause of this problem is that most research software
-isn't \emph{robust}. The difference between running and being
-robust is the difference between ``works for me on my machine'' and
-``works for other people on a cluster I've never used''. The lack of
-robustness also leads to many duplicated efforts,
-slowing the pace of research~\cite{prabhu2011,lawlor2015}.
-Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
-hundreds of tools to perform the same task. In the Bioinformatics
-Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
-to analyze transcript expression, and 182 pathway and interaction resources.
-Some of these tools are legitimate efforts to improve the state-of-the-art, but
-often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
-after publication~\cite{nekrutenko2012}.
-
-This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}. 
-Best practices in software engineering specifically aim to increase software
-robustness. However, most bioinformaticians come from related disciplines and
-are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
-the existing training programs and initiatives rarely have the time to cover software engineering
-in-depth, especially since the field is so broad and rapidly
-developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
-and knowledge are taught, developing robust software is not directly rewarded
-in science and funding is difficult to come by~\cite{prins2015}. Some proposed
-solutions to this problem, in addition to restructuring educational programs,
-include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
-partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
-or using specific technical tools like containerization or cloud
-computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
-cases, institutional change.
-
-You don't need to be a full-time programmer to write robust software. In fact,
-some of the best, most reliable pieces of software in many scientific
-communities are written by researchers~\cite{prabhu2011,sanders2008}. 
-The most widely-used software packages tend to adopt strong software
-engineering approaches, have high standards of reproducibility, good testing
-practices, and inter-institutional collaborations that grow into strong user
-bases. In the bioinformatics community, Bioconductor and Galaxy follow this
-path~\cite{gentleman2004,afgan2016}.
-
-That said, not \emph{every} coding effort needs to be engineered to last.
-Code that is used once to answer a specific question related to a specific dataset
-doesn't require comprehensive documentation or flexible configuration,
-and the only sensible way to test it may well be to run it on the dataset in
-question. Exploratory analysis is an iterative process that is developed
-quick and revised often~\cite{lawlor2015,sanders2008}. 
-However, if a script is dusted off and run three or four
-times for slightly different purposes,
-is crucial to a publication or a lab,
-or being passed on to someone else,
-it may be time to make your software more robust.
-
-In particular, robust software:
-
-\begin{itemize}
-\item
-  works for users other than the original author;
-\item
-  is kept under version control;
-\item
-  can be installed on more than one computer or account with relative ease;
-\item
-  has well-defined input and output formats;
-\item
-  has documentation that describes what its dependencies are, how to
-  install it, and what the options are; and
-\item
-  comes with enough tests to show that it actually runs.
-\end{itemize}
-
-These simple rules are also necessary steps toward creating a reusable
-library that can be shared and reused through a site like CPAN or CRAN.  They are
-generic and can be applied to all languages, libraries, packages,
-documentation styles, and operating systems for both closed-source and
-open-source software.  
-Whether the aim is as simple as sharing the
-code with collaborators or as complex as using the software in a
-production analysis environment, increasing the robustness of your
-software decreases headaches all around.
-
-\section{Have a README that explains in a few lines what the software does and what its dependencies are.}
-
-The README is the first stop for most new users.
-At a minimum, it needs to get a new user started and point them towards more
-help, if they need it. 
-Numerous guidelines exist on how to write good
-READMEs\cite{Johnson1997,gnustandards}. We summarize the crucial points here.
-
-\textbf{Explain what the software does:} At the beginning of the README,
-explain what the software does in one or two sentences.
-There's nothing more frustrating
-than spending the time to download and install some software only to
-find out that it doesn't do what you thought it did.
-
-\textbf{List required dependencies:} Often, software depends on
-specific versions of libraries, modules, or operating systems. 
-This is
-entirely reasonable as long as it is properly documented. Include the name and
-version number for each dependency.  Several methods exist for acquiring
-required packages: either provide the commands necessary to download the
-dependencies; link to the software homepage; or (our preference) use a
-dependency manager. We address dependencies in more detail in rule 5.
-
-\textbf{Provide compile/installation instructions:} If the software needs to be compiled
-or installed, list those instructions in the README. Also
-mention if you recommend they use a pre-compiled binary instead
-through a system such as pip, yum or apt.
-
-\textbf{List input and output files}: All possible input and output files
-should be listed in this section. Do the files conform to a particular
-standard, an extension of an existing format, or is it your own
-format? If using a standard format, link to the specification and
-version. If you extend the standard or have your own format, define it
-explicitly, listing all the required fields and acceptable values.
-(You get bonus points if you include a script to convert between
-a standard format and your file format). If there is no rigorous format
-(such as with log files), show an example file, or the first few lines,
-and explain what the sections mean.
-
-Input files and their formats are included in most documentation, but
-intermediate, auxiliary and log files are often missing.
-\emph{All} files should be listed in the README, even those considered self-explanatory. Log
-files are often full of valuable information that can be
-mined for the user's specific purpose. If your users might need to know,
-``Does this program report the percentage of reads trimmed to remove
-adapter sequences?'' they should be able to check the README and confidently
-say, ``Yes, it is in the log file''.
-
-\textbf{State attributions and licensing:} Attributions are how you credit
-your main contributors; licenses are how you want others to use and
-credit your software. READMEs are usually available \emph{before}
-the software is installed (or even downloaded).
-Leave no
-question in anyone's mind about whether your software can be used
-commercially, how much modification is permitted, and how other software
-needs to credit you. If your software is not open source, state that clearly.
-Attributions can also contain a list of ``expert'' users
-that can be contacted if new users have problems with the software.
-
-The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
-is a good model:
-it explains the software's purpose,
-tells readers where to get help,
-and includes links to a CITATION file (explaining how to cite the project)
-and the license file.
-
-\section{Print usage information when launching from the command line that explains the software's features.}
-
-Usage information provides the first line of help for both first-time and
-experienced users of command-line applications.
-Ideally, usage is a terse, informative command-line help message that
-guides the user in the correct use of your software. Terseness is
-important: usage that extends for multiple screens is a nuisance, especially when
-printed to standard error instead of standard output (where it can
-easily be paged).
-
-Usage should provide all of the information necessary to run the
-software. It is invoked either by running the software without
-any arguments; running the software with incorrect arguments; or by
-explicitly choosing a help or usage option.
-More standard command-line behaviours are detailed in \cite{Seemann2013}.
-
-Almost all command-line applications use a combination of
-POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for usage. 
-An example of good usage is GNU's \texttt{mkdir} command, which makes
-new directories:
-
-\begin{small}
-\begin{verbatim}
-$ mkdir --help
-Usage: mkdir [OPTION]... DIRECTORY...
-Create the DIRECTORY(ies), if they do not already exist.
-
-Mandatory arguments to long options are mandatory for short options too.
-  -m, --mode=MODE   set file mode (as in chmod), not a=rwx - umask
-  -p, --parents     no error if existing, make parent directories as needed
-  -v, --verbose     print a message for each created directory
-  -Z, --context=CTX  set the SELinux security context of each created
-                      directory to CTX
-      --help     display this help and exit
-      --version  output version information and exit
-
-Report mkdir bugs to bug-coreutils@gnu.org
-GNU coreutils home page: <http://www.gnu.org/software/coreutils/>
-General help using GNU software: <http://www.gnu.org/gethelp/>
-For complete documentation, run: info coreutils 'mkdir invocation'
-\end{verbatim}
-\end{small}
-
-\textbf{The syntax for running the program}: This includes the name of the
-program and defines the relative
-location of optional and required flags, arguments and values for execution.
-Arguments in {[}square brackets{]}
-are usually optional. An ellipsis (\ldots e.g. ``{[}OPTION{]}\ldots{}'')
-indicates that more than one value can be provided.
-
-\textbf{Description}: Similar to the README, the
-description reminds users of the software's primary function.
-
-\textbf{Most commonly used arguments, a description of each, and the
-default values}: Not all arguments need to appear in the usage, but the most
-commonly used ones should be listed here. Users will rely on this for
-quick reference when working with your software.
-
-\textbf{Where to find more information}: Whether an email address, web
-site or manual page, there should be an indication where the user can go
-to find more information about the software.
-
-\textbf{Printed to standard output} : So that it can be piped into
-\texttt{less}, searched with \texttt{grep}, or compared to the previous
-version with \texttt{diff}.
-
-\textbf{Exit with an appropriate exit code}: When usage is invoked by
-providing incorrect arguments, the program should exit with a non-zero
-code to indicate an error. However, when help is explicitly requested,
-the software should not exit with an error.
-
-\section{Version your releases}
-
-Software evolves over time, with developers adding or removing features as need
-dictates. Making official releases stamps a particular set of features with a
-project-specific identifier so that version can be retrieved for later use. For
-example, if a paper is published, the software should be released at the same
-time so that the results can be reproduced. 
-Most software has a version number composed of a decimal number that
-increments as new versions are released.
-
-There are many different ways
-to construct and interpret the version, but most importantly for us, a
-particular software version run with the same parameters should give
-identical results no matter when it's run. Results include both correct
-output as well as any errors.
-\textbf{Increment your version number every time you release your software to
-other people}.
-
-\withurl{Semantic versioning}{http://semver.org/} is one of the most common
-types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6.
-Changes in the major
-version number herald significant changes in the software that are not
-backwards compatible, such as changing or removing features or altering
-the primary functions of the software. Increasing the minor version
-represents incremental improvements in the software, like adding new
-features. Following the minor version number can be an arbitrary number
-of project-specific identifiers, including patches, builds and qualifiers.
-Common qualifiers include \texttt{alpha}, \texttt{beta}, and \texttt{SNAPSHOT},
-for applications that are
-not yet stable or released, and \texttt{-RC} for release candidates prior
-to an official release.
-
-\textbf{The version of your software should be easily available by 
-supplying \texttt{-\/-version} or \texttt{-v} on the command line}. This command should print
-the software name and version number, and it should
-also be \textbf{included in all of the program's output}, particularly debugging
-traces: if someone needs help, it's important that they be able to tell
-whoever's helping them which version of the software they're using.
-
-While new releases may make a program better in general,
-they can simultaneously create work for someone
-who integrated the old version into their own workflow a year or two ago,
-and won't see any benefits from upgrading.
-A program's authors should therefore \textbf{ensure that every released version
-continues to be available.}
-A number of mechanisms exist for
-controlled release that range from as simple as adding an appropriate
-commit message or tag to version control, to official releases alongside
-code on Bitbucket or GitHub, to depositing into a
-repository like apt, yum, homebrew, CPAN, etc. Choose the method that
-best suits the number and expertise of users you anticipate.
-
-\section{Reuse software (within reason).}
-
-In the spirit of code reuse and interoperability, developers often want
-to reuse software written by others. 
-With a few lines, a call
-is made out to the other program and the results are incorporated into the
-primary script. Using popular projects reduces the amount of code that
-needs to be maintained and leverages the work done by the other software.
-
-Unfortunately, reusing software introduces dependencies, which can
-bring their own special pain.  The interface between two software
-packages can also be a source of considerable frustration: all too
-often, support requests descend into debugging errors produced by the
-other project. 
-In addition, every package someone has to install
-before being able to use yours is a
-possible (some would say ``likely'') source of frustration for some
-potential user. The way the second program is invoked can throw up errors.
-The program may not be in the user's path, or it may be an
-older or newer version and produce results different than expected.
-Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explicitly or
-use shell-specific conventions like `*' expansion.
-Even Linux-standard functions available vary slightly between
-installs. For example, GNU \texttt{sort} is available on almost every
-*nix distribution, but sorts differently depending on locale.
-
-Despite these problems, software developers in research should
-re-use existing software wherever possible and reasonable to avoid
-``reinventing the wheel''. We fully support the reuse of software between
-projects provided three additional guidelines are adhered to.
-
-First, 
-\textbf{make sure that you really need the auxiliary program}. If you are
-executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration.
-
-Second, \textbf{ensure the appropriate software and version is available}.
-Early in execution, check whether the dependency is executable and whether the
-version is compatible.
-Either allow the user to
-configure the exact path to the package, distribute the program with the
-dependent software, or download it during installation using a
-dependency management system (see Rule 5).
-
-Finally, to ensure support on as many different operating systems as
-possible, \textbf{use native functions for starting other processes}, such as
-Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
-command, and be sure to capture and report the output of the subprocess's standard error
-to facilitate debugging.
-
-\section{Use a build utility and package manager.}
-
-In all but the simplest of scripts, \textbf{we highly recommend using a build
-automation utility to compile code and/or manage your software workflow}. The benefits of
-using a build utility like Make, Rake, Maven, Ant or MS Build vastly outweighs the
-drawbacks associated with the learning curve. Most importantly, a build system
-lets someone else build your software the same way you would. Most build
-utilities can also be configured to run tests during every build and
-re-generate documentation, and can be hooked into continuous integration,
-cutting releases, and deploying the software to other environments.
-
-Build utilities rely on any dependencies being present in the environment for a
-successful build, and so developers should
-\textbf{document \emph{all} dependencies, preferably in a machine-readable form}.
-Package managers like apt and yum are available on most *nix systems, and
-application package managers exist for specific languages like Python (pip),
-Java (Maven/Gradle), and Ruby (RubyGems). These package managers can be used
-together with the build utility to ensure that dependencies are available at
-compile/run time.
-For example, it is common for Python projects to include a file called
-\texttt{requirements.txt} that lists the names of required libraries,
-along with version ranges:
-
-\begin{verbatim}
-requests>=2.0
-pygithub>=1.26,<=1.27
-python-social-auth>=0.2.19,<0.3
-\end{verbatim}
-
-This file can be read by pip package manager, which can check that the
-required software is available, and install it if it is not. 
-Whatever is used,
-we recommend that developers should \emph{always} install dependencies
-using this mechanism, especially on their personal machines, so that
-they're sure it works.
-
-Conversely, developers should
-\textbf{avoid depending on scripts and tools which are not available as packages}.
-In many cases, a program's author may not realize that some tool was built locally, and
-doesn't exist elsewhere. At present, the only sure way to discover such
-unknown dependencies is to install on a system administered by someone
-else and see what breaks. As use of virtualization containers becomes more
-widespread, software installation can also be tested on a virtual machine or
-container system like Docker.
-
-\section{Do not require root or other special privileges.}
-
-Root (also known as ``superuser'' or ``admin'') is a special account on
-a computer that has (among other things) the power to modify or delete
-system files and user accounts. Conversely, files and directories owned
-by root usually cannot be modifed by normal users.
-
-Installing or running a program with root privileges is often
-convenient, since doing so automatically bypasses all those pesky safety
-checks that might otherwise get in the user's way. However, those checks
-are there for a reason: scientific software packages may not
-intentionally be malware, but one small bug or over-eager file-matching
-expression can certainly make them behave as if they were. Outside of
-very unusual circumstances,
-\textbf{packages should not require root privileges to set up or use}.
-
-Another reason for this rule is that users may want to try out a new
-package before installing it system-wide on a cluster. Requiring root
-privileges will frustrate such efforts, and thereby reduce uptake of the
-package. Requiring that software be installed under its own user account
-(e.g., that \texttt{packagename} be made a user, and all of the
-package's software be installed in that ``user's'' space 
-(see: Apache Tomcat))
-is similarly
-limiting, and makes side-by-side installation of multiple versions of
-the package more difficult.
-
-Developers should therefore
-\textbf{allow packages to be installed in an arbitrary location},
-e.g., under a user's home directory in
-\texttt{\textasciitilde{}/packagename}, or in directories with standard
-names like \texttt{bin}, \texttt{lib}, and \texttt{man} under a chosen
-directory. If the first option is chosen, the user may need to modify
-her search path to include the package's executables and libraries, but
-this can (more or less) be automated, and is much less risky than
-setting things up as root.
-
-Testing the ability to install software has traditionally been regarded as difficult,
-since it necessarily alters the machine on which the test is conducted.
-Lightweight virtualization containers like Docker make this much easier as well, 
-or simply \textbf{ask another person to try and build your software}.
-
-\section{Eliminate hard-coded paths.}
-
-It's easy to write software that reads input from a file called
-\texttt{mydata.csv}, but also very limiting. If a colleague asks you to
-process her data, you must either overwrite your data file (which is
-risky) or edit your code to read \texttt{otherdata.csv} (which is also
-risky, because there's every likelihood you'll forget to change the
-filename back, or will change three uses of the filename but not a
-fourth).
-
-Hard-coding file paths in a program also makes the software harder to run
-in other environments. If your package is installed on a cluster, for
-example, the user's data will almost certainly \emph{not} be in the same
-directory as the software, and the folder
-\texttt{C:\textbackslash{}users\textbackslash{}yourname\textbackslash{}}
-will probably not even exist.
-
-For these reasons, users should be able to
-\textbf{set the names and locations of input and output files as command-line parameters}.
-This rule applies to reference data sets as well as the user's own
-data: if a user wants to try a new gene identification algorithm using
-a different set of genes as a training set, she should not have to
-edit the software to do so.
-
-A corollary to this rule is
-\textbf{do not require users to navigate to a particular directory to do their work}.
-``Where I have to be'' is just another hard-coded path.
-
-In order to save typing, it is often convenient to allow users to
-specify an input or output \emph{directory}, and then require that there
-be files with particular names in that directory. This practice, which
-is sometimes called ``convention over configuration'', is used by many
-software frameworks, such as WordPress and Ruby on Rails, and often
-strikes a good balance between adaptability and consistency.
-
-\section{Allow configuration of all useful parameters from the command line.}
-
-Useful parameters are those values that a user will need to modify to suit
-their computer, dataset or application. 
-The full list of useful parameters is software-specific and so cannot be
-detailed here, but a short list includes providing
-input and reference files and directories,
-output files and directories,
-filtering,
-tuning,
-random number generation seeds,
-and
-any alternatives that you've built-in,
-e.g.,
-compress results, use a different calculation, verbose output, etc.
-
-Providing parameters on the
-command line increases the flexibility and usability of the program. You
-may have determined early in development that 0.58 is an optimal seed for your
-original dataset, but that doesn't mean that is the best seed for every
-case. Being able to change parameters on the fly to determine if and how
-they change the results is important as your software gains more users,
-facilitating exploratory analysis and parameter sweeping.
-\textbf{Choose reasonable defaults where they exist},
-and \textbf{set no defaults at all when there aren't any reasonable ones}.
-
-You can set reasonable default values
-as long as any command line arguments
-override those values. Configuration files should be used in preference
-to hard-coding the defaults directly. Configuration files can be in a standard
-location, e.g. \texttt{.packagerc} in the user's home directory or
-provided on the command line as an additional argument.
-Only values that are unlikely to
-change between runs belong in the config file, such as dependencies,
-servers, version numbers, network drives, and any other defaults for
-your lab or institution. 
-Use a default configuration file and specify all other parameters on the
-command line.
-
-When the software starts, it should \textbf{echo all parameters and software
-versions to standard out or a log file alongside the results}. This
-feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
-Finally,
-if the software is configurable in any way,
-it should \textbf{check that all input values are in a reasonable range at startup}. 
-Few things are as annoying as having a program announce after running for two hours
-that it isn't going to save its results
-because the requested directory doesn't exist.
-
-\section{Include a small test set that can be run to ensure the software is actually working.}
-
-Every package should come with a small test script for users to run
-after installation. Its purpose is not only to check that the software
-is working correctly (although that is extremely helpful), but rather to
-ensure that it will work at all. This test script can also serve as a
-working example of how to run the software.
-
-In order to be useful, \textbf{this test script must be easy to find and run}. 
-Many build systems will also run unit tests if provided them at compile time.
-If the build system is not amenable to testing, a 
-single file in the project's root directory named \texttt{runtests.sh}
-or something equally obvious is a much better solution than documenting
-test cases and requiring people to copy and paste them into the shell.
-
-Equally, \textbf{the test script's output must be easy to interpret}. Screens
-full of correlation coefficients do not qualify: instead, the script's
-output should be simple to understand for non-experts,
-such as one line per test, with the test's name
-and its pass/fail status, followed by a single summary line saying how
-many tests were run and how many passed or failed. If many or all tests
-fail because of missing dependencies, that fact should be displayed
-once, clearly, rather than once per test, so that users have a clear
-idea of what they need to fix and how much work it's likely to take.
-
-Studies have found that the ease with which people can start making
-contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
-By making it simpler for outsiders to contribute,
-a test suite of any kind also makes it more likely that they will, and software
-with collaborators stands a better chance of surviving in the busy field of
-scientific software.
-
-\section{Produce identical results when given identical inputs.}
-
-Given a set of parameters and a dataset, \textbf{a particular version of a program
-should produce the same results every time it is run}
-to aid testing, debugging, and reproducibility.
-Even minor changes to code can cause minor changes in output because of floating-point issues,
-which means that getting exactly the same output for the same input and parameters
-probably won't work during development,
-but it should still be a goal for people who have deployed a specific version.
-
-Many applications rely on randomized algorithms to
-improve performance or runtimes. As a consequence, results can change
-between runs, even when provided with the same data and parameters. By
-its nature, this randomness renders strict reproducibility and therefore
-debugging more difficult. If even the small test set (\#9) produces
-different results, new users may not be able to tell whether the software is
-working properly. When comparing
-results between versions or after changing parameters, even small
-differences can confuse or muddy the comparison. And especially when
-producing results for publications, grants or diagnoses, any analysis
-should be absolutely reproducible.
-
-Given the size of biological data, it is unreasonable to suggest that
-random algorithms be removed. However, most programs use a pseudo-random
-number generator, which uses a starting seed and an equation to
-approximate random numbers. Setting the seed to a consistent value
-can remove randomness between runs. \textbf{Allow the user to optionally provide
-the random seed as an input parameter}, thus rendering the program deterministic
-for those cases where it matters. If the seed is set internally (e.g.,
-using clock time), echo it to the output for re-use later. 
-If setting the seed is not possible, \textbf{make sure the acceptable tolerance is
-known and detailed in documentation and in the tests}.
-
-\section*{What We Left Out}
-
-Any short list of rules for developing robust software will
-necessarily be incomplete.  A few of our more significant omissions
-include:
-
-\begin{description}
-
-\item[\textbf{Services}] The rules we present above are necessary, but
-  not sufficient, for software that needs to interact with services
-  such as database and web servers.  Such software is often stopped
-  and started under automatic control, may need to authenticate, etc.,
-  all of which are out of scope of this paper.
-
-\item[\textbf{Documentation}] We think it is very important for
-  developers to document their work, but our experience is that people
-  are unlikely do it during normal development, and we don't want to
-  recommend practices that people won't actually adopt. However, it is worth
-  noting that software that is widely used and contributed to has and enforces
-  the need for good documentation~\cite{gentleman2004}.
-  Numerous resources exist for writing high quality
-  documentation~\cite{karimzadeh2016} and so we do not cover it here.
-
-\end{description}
-
-\section*{Conclusion}
-
-There has been extended discussion over the past few years of the
-sustainability of research software, but this question is meaningless
-in isolation: any piece of software can be sustained if its users are
-willing to put in enough effort.  The real equation is the ratio
-between the skill and effort available, and the ease with which
-software can be installed, understood, used, maintained, and extended.
-Following the ten rules we outline here reduce the denominator, and
-thereby enable researchers to build on each other's work more easily.
-
-\bibliography{robust-software}
+\bibliography{robust-checks}
 

### 2016-10-14 robust-software.tex
Submission to arXiv

@@
 
-\section*{Notes}
-
 \section*{Author Summary}
@@
 
-Typically, scientific software is developed and used by a single person,
-usually a graduate student or postdoc, and produces the intended results in their
-hands~\cite{prins2015}. 
-But what happens
-when someone else wants to use their software? Everyone with
+Scientific software is typically developed and used by a single person,
+usually a graduate student or postdoc~\cite{prins2015}.
+It may produce the intended results in their hands,
+but what happens when someone else wants to run it? Everyone with
 a few years of experience feels a bit nervous when told to use
-another person's code to analyze their data.
-Often, that software will be undocumented and work in unexpected
-ways (if it works at all).  It will often rely on nonexistent paths or resources,
+another person's code to analyze their data:
+it will often be undocumented,
+work in unexpected ways (if it works at all),
+rely on nonexistent paths or resources,
 be tuned for a single dataset,
 or simply be an older version than was used in published papers.
-The new user is then faced with two unpalatable options:
-hack the existing code to make it work, or start over. Being unable to
+The potential new user is then faced with two unpalatable options:
+hack the existing code to make it work, or start over.
+
+Being unable to
 replicate results is so common that one publication refers to it as "a rite of
 passage"~\cite{baker2016}. 
-
 The root cause of this problem is that most research software
@@
 robustness also leads to many duplicated efforts,
-slowing the pace of research~\cite{prabhu2011,lawlor2015}.
+which slows the pace of research~\cite{prabhu2011,lawlor2015}.
 Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
-hundreds of tools to perform the same task. In the Bioinformatics
-Links Directory as of 2016, there are 84 different multiple sequence aligners, 141 tools
+hundreds of tools to perform the same task:
+for example,
+in 2016 the Bioinformatics Links Directory included 84 different multiple sequence aligners, 141 tools
 to analyze transcript expression, and 182 pathway and interaction resources.
@@
 Best practices in software engineering specifically aim to increase software
-robustness. However, most bioinformaticians come from related disciplines and
-are trained on-the-job or otherwise informally~\cite{prins2015,atwood2015}. Even
-the existing training programs and initiatives rarely have the time to cover software engineering
-in-depth, especially since the field is so broad and rapidly
-developing~\cite{atwood2015,lawlor2015}. Regardless of whether or not the skills
-and knowledge are taught, developing robust software is not directly rewarded
-in science and funding is difficult to come by~\cite{prins2015}. Some proposed
-solutions to this problem, in addition to restructuring educational programs,
-include hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
+robustness. However, most bioinformaticians learn what they know about software development
+on the job or otherwise informally~\cite{prins2015,atwood2015}.
+Existing training programs and initiatives rarely have the time to cover software engineering
+in depth, especially since the field is so broad and developing so rapidly~\cite{atwood2015,lawlor2015}.
+In addition, making software robust is not directly rewarded
+in science, and funding is difficult to come by~\cite{prins2015}. Some proposed
+solutions to this problem include restructuring educational programs,
+hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
 partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
@@
 
-You don't need to be a full-time programmer to write robust software. In fact,
+The good news is,
+you don't need to be a professionally-trained programmer to write robust software. In fact,
 some of the best, most reliable pieces of software in many scientific
-communities are written by researchers~\cite{prabhu2011,sanders2008}. 
-The most widely-used software packages tend to adopt strong software
-engineering approaches, have high standards of reproducibility, good testing
-practices, and inter-institutional collaborations that grow into strong user
+communities are written by researchers~\cite{prabhu2011,sanders2008}
+who have adopted strong software
+engineering approaches, have high standards of reproducibility, use good testing
+practices, and foster inter-institutional collaborations that grow into strong user
 bases. In the bioinformatics community, Bioconductor and Galaxy follow this
@@
 
-That said, not \emph{every} coding effort needs to be engineered to last.
-Code that is used once to answer a specific question related to a specific dataset
-doesn't require comprehensive documentation or flexible configuration,
-and the only sensible way to test it may well be to run it on the dataset in
-question. Exploratory analysis is an iterative process that is developed
-quick and revised often~\cite{lawlor2015,sanders2008}. 
-However, if a script is dusted off and run three or four
-times for slightly different purposes,
-is crucial to a publication or a lab,
-or being passed on to someone else,
-it may be time to make your software more robust.
-
-In particular, robust software:
+So what \emph{is} ``robust'' software?
+In brief, it:
 
@@
 
-These simple rules are also necessary steps toward creating a reusable
+These simple rules don't just ensure that the software will run for other people;
+they are also necessary steps toward creating a reusable
 library that can be shared and reused through a site like CPAN or CRAN.  They are
@@
 
-\section{Have a README that explains in a few lines what the software does and what its dependencies are.}
+\section{Have a README that briefly explains what the software does and what its dependencies are.}
 
@@
 Numerous guidelines exist on how to write good
-READMEs\cite{Johnson1997,gnustandards}. We summarize the crucial points here.
+READMEs\cite{Johnson1997,gnustandards};
+their key common features are listed below.
 
@@
 There's nothing more frustrating
-than spending the time to download and install some software only to
+than downloading and installing something only to
 find out that it doesn't do what you thought it did.
@@
 
-\textbf{Provide compile/installation instructions:} If the software needs to be compiled
+\textbf{Provide compilation/installation instructions:}
+If the software needs to be compiled
 or installed, list those instructions in the README. Also
 mention if you recommend they use a pre-compiled binary instead
-through a system such as pip, yum or apt.
+through a system such as pip, yum, or apt.
 
 \textbf{List input and output files}: All possible input and output files
-should be listed in this section. Do the files conform to a particular
-standard, an extension of an existing format, or is it your own
-format? If using a standard format, link to the specification and
+should be listed in this section. Do the files use a particular
+standard, an extension of an existing format, or your own
+format?  If they use a standard format, link to the specification and
 version. If you extend the standard or have your own format, define it
@@
 a standard format and your file format). If there is no rigorous format
-(such as with log files), show an example file, or the first few lines,
-and explain what the sections mean.
+(which is common with log files), show at least a few lines from an
+example file and explain what the sections mean.
 
 Input files and their formats are included in most documentation, but
-intermediate, auxiliary and log files are often missing.
+intermediate, auxiliary, and log files are often missing.
 \emph{All} files should be listed in the README, even those considered self-explanatory. Log
@@
 \textbf{State attributions and licensing:} Attributions are how you credit
-your main contributors; licenses are how you want others to use and
-credit your software. READMEs are usually available \emph{before}
+your main contributors; licenses are how others may use and
+credit your work. These files should be available \emph{before}
 the software is installed (or even downloaded).
@@
 Ideally, usage is a terse, informative command-line help message that
-guides the user in the correct use of your software. Terseness is
+guides the user in the correct use of the software. Terseness is
 important: usage that extends for multiple screens is a nuisance, especially when
 printed to standard error instead of standard output (where it can
-easily be paged).
+easily be paged).  Long output is also much less likely to be read{\ldots}
 
@@
 software. It is invoked either by running the software without
-any arguments; running the software with incorrect arguments; or by
+any arguments, running it with incorrect arguments, or by
 explicitly choosing a help or usage option.
@@
 \textbf{Most commonly used arguments, a description of each, and the
-default values}: Not all arguments need to appear in the usage, but the most
-commonly used ones should be listed here. Users will rely on this for
-quick reference when working with your software.
+default values}: Not all arguments need to appear in the usage, but those most
+commonly used should be listed. Users will rely on this for
+quick reference when working with the software.
 
-\textbf{Where to find more information}: Whether an email address, web
-site or manual page, there should be an indication where the user can go
-to find more information about the software.
+\textbf{Where to find more information}: Whether it's an email address, web
+site, or manual page, there should be an indication where the user can go
+to find out more about the software.
 
-\textbf{Printed to standard output} : So that it can be piped into
+\textbf{Printed to standard output}: Usage should be displayed on standard output so that it can be piped into
 \texttt{less}, searched with \texttt{grep}, or compared to the previous
@@
 
-\section{Version your releases}
+\section{Version your releases.}
 
@@
 time so that the results can be reproduced.
+
 Most software has a version number composed of a decimal number that
 increments as new versions are released. 
-
 There are many different ways
-to construct and interpret the version, but most importantly for us, a
+to construct and interpret this number, but most importantly for us, a
 particular software version run with the same parameters should give
@@
 also be \textbf{included in all of the program's output}, particularly debugging
-traces: if someone needs help, it's important that they be able to tell
+traces.  If someone needs help, it's important that they be able to tell
 whoever's helping them which version of the software they're using.
@@
 and won't see any benefits from upgrading.
-A program's authors should therefore \textbf{ensure that every released version
-continues to be available.}
+A program's authors should therefore \textbf{ensure that old released versions
+continue to be available.}
 A number of mechanisms exist for
@@
 Despite these problems, software developers in research should
-re-use existing software wherever possible and reasonable to avoid
-``reinventing the wheel''. We fully support the reuse of software between
-projects provided three additional guidelines are adhered to.
-
+re-use existing software provided three guidelines are adhered to.
 First, 
@@
 possible, \textbf{use native functions for starting other processes}, such as
-Java's \texttt{Runtime.exec} call, Python's subprocess module, and Perl's system
+Java's \texttt{Runtime.exec} call, Python's \texttt{subprocess} module, and Perl's \texttt{system}
 command, and be sure to capture and report the output of the subprocess's standard error
@@
 
-In all but the simplest of scripts, \textbf{we highly recommend using a build
+In all but the simplest of scripts, \textbf{use a build
 automation utility to compile code and/or manage your software workflow}. The benefits of
-using a build utility like Make, Rake, Maven, Ant or MS Build vastly outweighs the
-drawbacks associated with the learning curve. Most importantly, a build system
+using tools like Make, Rake, Maven, Ant or MS Build vastly outweigh the
+their learning curves. Most importantly, a build system
 lets someone else build your software the same way you would. Most build
@@
 \textbf{document \emph{all} dependencies, preferably in a machine-readable form}.
-Package managers like apt and yum are available on most *nix systems, and
+Package managers like apt and yum are available on most Unix-like systems, and
 application package managers exist for specific languages like Python (pip),
@@
 
-This file can be read by pip package manager, which can check that the
-required software is available, and install it if it is not. 
+This file can be read by the pip package manager, which can check that the
+required software is available and install it if it is not. 
 Whatever is used,
 we recommend that developers should \emph{always} install dependencies
-using this mechanism, especially on their personal machines, so that
+using their dependency description, especially on their personal machines, so that
 they're sure it works.
@@
 privileges will frustrate such efforts, and thereby reduce uptake of the
-package. Requiring that software be installed under its own user account
-(e.g., that \texttt{packagename} be made a user, and all of the
-package's software be installed in that ``user's'' space 
-(see: Apache Tomcat))
-is similarly
-limiting, and makes side-by-side installation of multiple versions of
+package. Requiring, as Apache Tomcat does, that software be installed
+under its own user account---i.e.,
+that \texttt{packagename} be made a user, and all of the
+package's software be installed in that pseudo-user's space---is similarly limiting,
+and makes side-by-side installation of multiple versions of
 the package more difficult.
@@
 edit the software to do so.
-
 A corollary to this rule is
-\textbf{do not require users to navigate to a particular directory to do their work}.
-``Where I have to be'' is just another hard-coded path.
+\textbf{do not require users to navigate to a particular directory to do their work},
+since ``where I have to be'' is just another hard-coded path.
 
@@
 and
-any alternatives that you've built-in,
-e.g.,
-compress results, use a different calculation, verbose output, etc.
+alternatives such as compressing results,
+use a variant algorithm,
+or verbose output.
 
@@
 command line increases the flexibility and usability of the program. You
-may have determined early in development that 0.58 is an optimal seed for your
-original dataset, but that doesn't mean that is the best seed for every
+may have determined early in development that 0.58 is an optimal cluster threshold for your
+original dataset, but that doesn't mean it is best for every
 case. Being able to change parameters on the fly to determine if and how
 they change the results is important as your software gains more users,
-facilitating exploratory analysis and parameter sweeping.
+as it facilitates exploratory analysis and parameter sweeping.
 \textbf{Choose reasonable defaults where they exist},
@@
 after installation. Its purpose is not only to check that the software
-is working correctly (although that is extremely helpful), but rather to
-ensure that it will work at all. This test script can also serve as a
+is working correctly (although that is extremely helpful), but also to
+ensure that it works at all. This test script can also serve as a
 working example of how to run the software.
 
-In order to be useful, \textbf{this test script must be easy to find and run}. 
+In order to be useful, \textbf{make the test script easy to find and run}. 
 Many build systems will also run unit tests if provided them at compile time.
@@
 or something equally obvious is a much better solution than documenting
-test cases and requiring people to copy and paste them into the shell.
+test cases and requiring people to copy and paste them.
 
-Equally, \textbf{the test script's output must be easy to interpret}. Screens
+Equally, \textbf{make the test script's output easy to interpret}. Screens
 full of correlation coefficients do not qualify: instead, the script's
@@
 
-Studies have found that the ease with which people can start making
+Research has shown that the ease with which people can start making
 contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
@@
 debugging more difficult. If even the small test set (\#9) produces
-different results, new users may not be able to tell whether the software is
+different results for each run, new users may not be able to tell whether the software is
 working properly. When comparing
@@
 Any short list of rules for developing robust software will
-necessarily be incomplete.  A few of our more significant omissions
+necessarily be incomplete.   Our more significant omissions
 include:
@@
 \begin{description}
-
-\item[\textbf{Services}] The rules we present above are necessary, but
-  not sufficient, for software that needs to interact with services
-  such as database and web servers.  Such software is often stopped
-  and started under automatic control, may need to authenticate, etc.,
-  all of which are out of scope of this paper.
 
@@
   documentation~\cite{karimzadeh2016} and so we do not cover it here.
+
+\item[\textbf{Services}] The rules we present above are necessary, but
+  not sufficient, for software that needs to interact with services
+  such as database and web servers.  Such software is often stopped
+  and started under automatic control, may need to authenticate, etc.,
+  all of which are out of scope of this paper.
 
@@
 
+That said, not \emph{every} coding effort needs to be engineered to
+last.  Code that is used once to answer a specific question related to
+a specific dataset doesn't require comprehensive documentation or
+flexible configuration, and the only sensible way to test it may well
+be to run it on the dataset in question. Exploratory analysis is an
+iterative process that is developed quick and revised
+often~\cite{lawlor2015,sanders2008}.  However, if a script is dusted
+off and run three or four times for slightly different purposes, is
+crucial to a publication or a lab, or being passed on to someone else,
+it may be time to make your software more robust.
+
 \bibliography{robust-software}

### 2017-01-04 robust-software.tex
Clarifying our meaning of robust

@@
 or even off the primary developer's computer.
-We present ten simple rules to make such software robust enough to run anywhere,
-and inspire confidence in your reproducibility,
+We present ten simple rules to make such software robust enough to be run by anyone,
+anywhere,
 and thereby delight your users and collaborators.
@@
 The good news is,
-you don't need to be a professionally-trained programmer to write robust software. In fact,
+you don't need to be a professionally-trained programmer to write robust software.
+In fact,
 some of the best, most reliable pieces of software in many scientific
@@
 So what \emph{is} ``robust'' software?
-In brief, it:
+We implied above that it is software that works for people other than the original author,
+on machines other than its creator's.
+More specifically, we mean that:
 
@@
 \item
-  works for users other than the original author;
+  it can be installed on more than one computer with relative ease,
 \item
-  is kept under version control;
+  it works as advertised, and
 \item
-  can be installed on more than one computer or account with relative ease;
-\item
-  has well-defined input and output formats;
-\item
-  has documentation that describes what its dependencies are, how to
-  install it, and what the options are; and
-\item
-  comes with enough tests to show that it actually runs.
+  it can be integrated with other tools.
 \end{itemize}
 
-These simple rules don't just ensure that the software will run for other people;
-they are also necessary steps toward creating a reusable
-library that can be shared and reused through a site like CPAN or CRAN.  They are
-generic and can be applied to all languages, libraries, packages,
-documentation styles, and operating systems for both closed-source and
-open-source software.  
-Whether the aim is as simple as sharing the
-code with collaborators or as complex as using the software in a
-production analysis environment, increasing the robustness of your
-software decreases headaches all around.
+Our rules are generic and can be applied to all languages, libraries, packages,
+documentation styles, and operating systems for both closed-source and open-source software.  
+They are also necessary steps toward making computational research reproducible:
+after all,
+if the tools and libraries used to produce a result cannot be run by others,
+they cannot verify or build on published results.
 

### 2017-01-04 robust-software.tex
Clarifying note about version control

--- 2016-10-14 robust-software.tex
@@
 software decreases headaches all around.
+
+Note: one practice \emph{not} included in the list below is the use of version control.
+We think this is essential to building software,
+and indeed to doing reproducible scientific work of any kind
+\cite{wilson2014,wilson2016}.
+However,
+our focus in this short guide is on the software that is built,
+not on the process used to build it.
 

### 2017-01-04 robust-software.tex
Extending the note on version control and maintainability

@@
 Note: one practice \emph{not} included in the list below is the use of version control.
-We think this is essential to building software,
-and indeed to doing reproducible scientific work of any kind
+We think this is essential to building software of any kind
 \cite{wilson2014,wilson2016}.
@@
 not on the process used to build it.
+Similarly,
+we do not explicitly discuss code review,
+unit testing,
+and other practices essential to maintainability and sustainable software development.
 

### 2017-01-04 robust-software.tex
Reorganizing "print usage" and "configurability".
1.  Combining requirement to log actions with requirement to print usage.
2.  Moving configurability to immediately follow command-line usage requirement.

--- 2017-01-04 robust-software.tex
@@
 
-Note: one practice \emph{not} included in the list below is the use of version control.
-We think this is essential to building software,
-and indeed to doing reproducible scientific work of any kind
-\cite{wilson2014,wilson2016}.
-However,
-our focus in this short guide is on the software that is built,
-not on the process used to build it.
-
 \section{Have a README that briefly explains what the software does and what its dependencies are.}
@@
 
-\section{Print usage information when launching from the command line that explains the software's features.}
+\section{Tell the user what could be done and what was actually done.}
 
+A robust program should
+\textbf{print usage information when launching from the command line that explains the software's features}.
 Usage information provides the first line of help for both first-time and
@@
 
-\textbf{The syntax for running the program}: This includes the name of the
-program and defines the relative
-location of optional and required flags, arguments and values for execution.
-Arguments in {[}square brackets{]}
-are usually optional. An ellipsis (\ldots e.g. ``{[}OPTION{]}\ldots{}'')
-indicates that more than one value can be provided.
+The key features of this usage message are:
 
-\textbf{Description}: Similar to the README, the
-description reminds users of the software's primary function.
+\begin{itemize}
 
-\textbf{Most commonly used arguments, a description of each, and the
-default values}: Not all arguments need to appear in the usage, but those most
-commonly used should be listed. Users will rely on this for
-quick reference when working with the software.
+\item[\textbf{The syntax for running the program}] This includes the
+  name of the program and defines the relative location of optional
+  and required flags, arguments and values for execution.  Arguments
+  in {[}square brackets{]} are usually optional. An ellipsis (\ldots
+  e.g. ``{[}OPTION{]}\ldots{}'') indicates that more than one value
+  can be provided.
 
-\textbf{Where to find more information}: Whether it's an email address, web
-site, or manual page, there should be an indication where the user can go
-to find out more about the software.
+\item[\textbf{Description}] Similar to the README, the description
+  reminds users of the software's primary function.
 
-\textbf{Printed to standard output}: Usage should be displayed on standard output so that it can be piped into
-\texttt{less}, searched with \texttt{grep}, or compared to the previous
-version with \texttt{diff}.
+\item[\textbf{Most commonly used arguments, a description of each, and
+    the default values}] Not all arguments need to appear in the
+  usage, but those most commonly used should be listed. Users will
+  rely on this for quick reference when working with the software.
 
-\textbf{Exit with an appropriate exit code}: When usage is invoked by
-providing incorrect arguments, the program should exit with a non-zero
-code to indicate an error. However, when help is explicitly requested,
-the software should not exit with an error.
+\item[\textbf{Where to find more information}] Whether it's an email
+  address, web site, or manual page, there should be an indication
+  where the user can go to find out more about the software.
+
+\item[\textbf{Printed to standard output}] Usage should be displayed
+  on standard output so that it can be piped into \texttt{less},
+  searched with \texttt{grep}, or compared to the previous version
+  with \texttt{diff}.
+
+\item[\textbf{Exit with an appropriate exit code}] When usage is
+  invoked by providing incorrect arguments, the program should exit
+  with a non-zero code to indicate an error. However, when help is
+  explicitly requested, the software should not exit with an error.
+
+\end{itemize}
+
+The usage message tells users what the program could do.
+It is equally important for the program to tell users what it actually did.
+Accordingly,
+when the program starts, it should \textbf{echo all parameters and software
+versions to standard out or a log file alongside the results}. This
+feature supports greater reproducibility because any result can be
+replicated with only the previous output files as reference.
+
+(And as a corollary
+if the software is configurable in any way,
+it should \textbf{check that all input values are in a reasonable range at startup}. 
+Few things are as annoying as having a program announce after running for two hours
+that it isn't going to save its results
+because the requested directory doesn't exist.)
+
+\section{Allow configuration of all useful parameters from the command line.}
+
+Useful parameters are those values that a user will need to modify to suit
+their computer, dataset or application. 
+The full list of useful parameters is software-specific and so cannot be
+detailed here, but a short list includes providing
+input and reference files and directories,
+output files and directories,
+filtering,
+tuning,
+random number generation seeds,
+and
+alternatives such as compressing results,
+use a variant algorithm,
+or verbose output.
+
+Allowing parameters to be specific on the
+command line increases the flexibility and usability of the program. You
+may have determined early in development that 0.58 is an optimal cluster threshold for your
+original dataset, but that doesn't mean it is best for every
+case. Being able to change parameters on the fly to determine if and how
+they change the results is important as your software gains more users,
+as it facilitates exploratory analysis and parameter sweeping.
+
+\textbf{Choose reasonable defaults where they exist},
+and \textbf{set no defaults at all when there aren't any reasonable ones}.
+You can set reasonable default values
+as long as any command line arguments
+override those values. Configuration files should be used in preference
+to hard-coding the defaults directly. Configuration files can be in a standard
+location, e.g. \texttt{.packagerc} in the user's home directory or
+provided on the command line as an additional argument.
+Only values that are unlikely to
+change between runs belong in the config file, such as dependencies,
+servers, version numbers, network drives, and any other defaults for
+your lab or institution. 
+Use a default configuration file and specify all other parameters on the
+command line.
 
@@
 
-\section{Allow configuration of all useful parameters from the command line.}
-
-Useful parameters are those values that a user will need to modify to suit
-their computer, dataset or application. 
-The full list of useful parameters is software-specific and so cannot be
-detailed here, but a short list includes providing
-input and reference files and directories,
-output files and directories,
-filtering,
-tuning,
-random number generation seeds,
-and
-alternatives such as compressing results,
-use a variant algorithm,
-or verbose output.
-
-Providing parameters on the
-command line increases the flexibility and usability of the program. You
-may have determined early in development that 0.58 is an optimal cluster threshold for your
-original dataset, but that doesn't mean it is best for every
-case. Being able to change parameters on the fly to determine if and how
-they change the results is important as your software gains more users,
-as it facilitates exploratory analysis and parameter sweeping.
-\textbf{Choose reasonable defaults where they exist},
-and \textbf{set no defaults at all when there aren't any reasonable ones}.
-
-You can set reasonable default values
-as long as any command line arguments
-override those values. Configuration files should be used in preference
-to hard-coding the defaults directly. Configuration files can be in a standard
-location, e.g. \texttt{.packagerc} in the user's home directory or
-provided on the command line as an additional argument.
-Only values that are unlikely to
-change between runs belong in the config file, such as dependencies,
-servers, version numbers, network drives, and any other defaults for
-your lab or institution. 
-Use a default configuration file and specify all other parameters on the
-command line.
-
-When the software starts, it should \textbf{echo all parameters and software
-versions to standard out or a log file alongside the results}. This
-feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
-Finally,
-if the software is configurable in any way,
-it should \textbf{check that all input values are in a reasonable range at startup}. 
-Few things are as annoying as having a program announce after running for two hours
-that it isn't going to save its results
-because the requested directory doesn't exist.
-
 \section{Include a small test set that can be run to ensure the software is actually working.}

### 2017-01-04 robust-software.tex
Clarifying configuration files

@@
 
-\section{Allow configuration of all useful parameters from the command line.}
+\section{Make common operations easy to control.}
 
-Useful parameters are those values that a user will need to modify to suit
-their computer, dataset or application. 
-The full list of useful parameters is software-specific and so cannot be
-detailed here, but a short list includes providing
-input and reference files and directories,
+Being able to change parameters on the fly to determine if and how
+they change the results is important as your software gains more users,
+as it facilitates exploratory analysis and parameter sweeping.
+Programs should therefore
+\textbf{allow the most commonly changed parameters to be configured from the command line}.
+
+Users will want to change some values more often than others.
+Which ones is software-specific and so cannot be detailed here,
+but a short list includes input and reference files and directories,
 output files and directories,
-filtering,
-tuning,
+filtering parameters,
 random number generation seeds,
@@
 
-Allowing parameters to be specific on the
-command line increases the flexibility and usability of the program. You
-may have determined early in development that 0.58 is an optimal cluster threshold for your
-original dataset, but that doesn't mean it is best for every
-case. Being able to change parameters on the fly to determine if and how
-they change the results is important as your software gains more users,
-as it facilitates exploratory analysis and parameter sweeping.
-
-\textbf{Choose reasonable defaults where they exist},
+To make programs even easier to use,
+\textbf{choose reasonable defaults where they exist}
 and \textbf{set no defaults at all when there aren't any reasonable ones}.
@@
 as long as any command line arguments
-override those values. Configuration files should be used in preference
-to hard-coding the defaults directly. Configuration files can be in a standard
-location, e.g. \texttt{.packagerc} in the user's home directory or
-provided on the command line as an additional argument.
-Only values that are unlikely to
-change between runs belong in the config file, such as dependencies,
-servers, version numbers, network drives, and any other defaults for
-your lab or institution. 
-Use a default configuration file and specify all other parameters on the
-command line.
+override those values.
+
+Changeable values should \emph{never} be hard-coded:
+if users have to edit your software in order to run it,
+you have done something wrong.
+Changeable but infrequently-changed values should therefore be stored in configuration files.
+These can be in a standard location,
+e.g. \texttt{.packagerc} in the user's home directory,
+or provided on the command line as an additional argument.
+Configuration files are often created during installation
+to set up such things as server names,
+network drives,
+and other defaults for your lab or institution.
 

### 2017-01-11 robust-software.tex

--- 2017-01-04 robust-software.tex
@@
 software decreases headaches all around.
-
-Note: one practice \emph{not} included in the list below is the use of version control.
-We think this is essential to building software of any kind
-\cite{wilson2014,wilson2016}.
-However,
-our focus in this short guide is on the software that is built,
-not on the process used to build it.
-Similarly,
-we do not explicitly discuss code review,
-unit testing,
-and other practices essential to maintainability and sustainable software development.
 
@@
 
-\section{Use a build utility and package manager.}
+\section{Rely on build tools and package managers for installation.}
 
-In all but the simplest of scripts, \textbf{use a build
-automation utility to compile code and/or manage your software workflow}. The benefits of
-using tools like Make, Rake, Maven, Ant or MS Build vastly outweigh the
-their learning curves. Most importantly, a build system
-lets someone else build your software the same way you would. Most build
-utilities can also be configured to run tests during every build and
-re-generate documentation, and can be hooked into continuous integration,
-cutting releases, and deploying the software to other environments.
+Programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build
+to compile code, deploy applications, and automate other tasks.
+These tools can also be used to manage runtime environments,
+i.e.,
+to check that the right versions of required packages are installed
+and install or upgrade them if they are not.
 
-Build utilities rely on any dependencies being present in the environment for a
-successful build, and so developers should
-\textbf{document \emph{all} dependencies, preferably in a machine-readable form}.
+The same tools can and should be used to manage runtime environments on users' machines as well.
+Accordingly,
+developed should
+\textbf{document all dependencies in a machine-readable form}.
 Package managers like apt and yum are available on most Unix-like systems, and
@@
 compile/run time.
+
 For example, it is common for Python projects to include a file called
@@
 Whatever is used,
-we recommend that developers should \emph{always} install dependencies
+developers should \emph{always} install dependencies
 using their dependency description, especially on their personal machines, so that

### 2017-01-11 robust-software.tex
Fixup typo from pull request #43
Fixup awkward sentence from PR #42

@@
 Users will want to change some values more often than others.
-Which ones is software-specific and so cannot be detailed here,
+Since parameters are software-specific, the appropriate 'tunable' ones cannot be detailed here,
 but a short list includes input and reference files and directories,
@@
 Accordingly,
-developed should
+developers should
 \textbf{document all dependencies in a machine-readable form}.

### 2017-01-11 robust-software.tex

--- 2017-01-11 robust-software.tex
@@
 they cannot verify or build on published results.
+
+Note: one practice \emph{not} included in the list below is the use of version control.
+We think this is essential to building software of any kind
+\cite{wilson2014,wilson2016}.
+However,
+our focus in this short guide is on the software that is built,
+not on the process used to build it.
+Similarly,
+we do not explicitly discuss code review,
+unit testing,
+and other practices essential to maintainability and sustainable software development.
 

### 2017-01-11 robust-software.tex
Added environment for highlighting changes; changed the Usage items from 'itemize' to 'description' because it renders better

--- 2017-01-11 robust-software.tex
@@
 \noalign{\global\arrayrulewidth\savedwidth}}
+
+\usepackage{color}
+
+\newenvironment{reviewed}
+{ \color{orange} }
 
@@
 
+Note: one practice \emph{not} included in the list below is the use of version control.
+We think this is essential to building software of any kind
+\cite{wilson2014,wilson2016}.
+However,
+our focus in this short guide is on the software that is built,
+not on the process used to build it.
+Similarly,
+we do not explicitly discuss code review,
+unit testing,
+and other practices essential to maintainability and sustainable software development.
+
 \section{Have a README that briefly explains what the software does and what its dependencies are.}
@@
 
-\begin{itemize}
+\begin{description}
 
@@
 
-\end{itemize}
+\end{description}
 

### 2017-01-11 robust-software.tex
Added emphasis and citations for issue #29 reproducibility; emphasized the prototype nature of scientific software

--- 2017-01-11 robust-software.tex
@@
 Many common challenges can be avoided by following a few simple rules; doing so
-not only accelerates research, but also improves reproducibility.
+not only improves reproducibility, but can accelerate research.
 
@@
 Being unable to
-replicate results is so common that one publication refers to it as "a rite of
-passage"~\cite{baker2016}. 
+replicate results is so common that one publication refers to it as ``a rite of
+passage''~\cite{baker2016}. 
 The root cause of this problem is that most research software
-isn't \emph{robust}. The difference between running and being
-robust is the difference between ``works for me on my machine'' and
-``works for other people on a cluster I've never used''. The lack of
-robustness also leads to many duplicated efforts,
+is essentially a prototype, and therefore is not 
+\emph{robust}. The lack of
+robustness in published, distributed software
+leads to duplicated efforts with little practical benefit,
 which slows the pace of research~\cite{prabhu2011,lawlor2015}. 
+Not all software needs to be robust enough to run for other 
 Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
@@
 path~\cite{gentleman2004,afgan2016}.
+Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
+publish a paper about your software, it should, at minimum, satisfy these
+rules.
 
@@
 \item
-  it works as advertised, and
+  it works consistently as advertised, and
 \item
@@
 documentation styles, and operating systems for both closed-source and open-source software.  
-They are also necessary steps toward making computational research reproducible:
+They are also necessary steps toward making computational research replicable and reproducible:
 after all,
-if the tools and libraries used to produce a result cannot be run by others,
-they cannot verify or build on published results.
+if your tools and libraries cannot be run by others,
+they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.
+
+Note: practices \emph{not} included in the list below include version control,
+code review, unit testing, and other practices essential to maintainability and
+sustainable software development.
+These elements are essential to building software of any kind
+\cite{wilson2014,wilson2016}.
+However,
+our focus in this short guide is on the software that is already built,
+not on the process used to build it.
 

### 2017-01-11 robust-software.tex
Reworked the Reuse section with the package manager rule in mind; added a guideline to depend on stable software. Issue #31

@@
 With a few lines, a call
-is made out to the other program and the results are incorporated into the
+is made out to another library or program and the results are incorporated into the
 primary script. Using popular projects reduces the amount of code that
@@
 
-Unfortunately, reusing software introduces dependencies, which can
+Unfortunately, reusing software (whether software libraries or separate executables)
+introduces dependencies, which can
 bring their own special pain. The interface between two software
-packages can also be a source of considerable frustration: all too
+packages can be a source of considerable frustration: all too
 often, support requests descend into debugging errors produced by the
-other project. 
-In addition, every package someone has to install
-before being able to use yours is a
-possible (some would say ``likely'') source of frustration for some
-potential user. The way the second program is invoked can throw up errors.
-The program may not be in the user's path, or it may be an
-older or newer version and produce results different than expected.
-Windows users will be frustrated if you invoke \texttt{bash} or \texttt{sh} explicitly or
-use shell-specific conventions like `*' expansion.
-Even Linux-standard functions available vary slightly between
-installs. For example, GNU \texttt{sort} is available on almost every
-*nix distribution, but sorts differently depending on locale.
+other project due to incompatible libraries, versions, or operating
+systems~\cite{brown2013}. Even introducing libraries in the same programming
+language can rely on software installed in the environment.
 
 Despite these problems, software developers in research should
-re-use existing software provided three guidelines are adhered to.
+re-use existing software provided a few guidelines are adhered to.
+
 First, 
@@
 executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration.
+it may not be worth the pain of integration. Reuse software that offers some
+measurable improvement to your build environment.
 
-Second, \textbf{ensure the appropriate software and version is available}.
-Early in execution, check whether the dependency is executable and whether the
-version is compatible.
+Second, if launching an executable, \textbf{ensure the appropriate software and version is available}.
 Either allow the user to
 configure the exact path to the package, distribute the program with the
-dependent software, or download it during installation using a
-dependency management system (see Rule 5).
+dependent software, or download it during installation using your
+package manager.
 
-Finally, to ensure support on as many different operating systems as
-possible, \textbf{use native functions for starting other processes}, such as
-Java's \texttt{Runtime.exec} call, Python's \texttt{subprocess} module, and Perl's \texttt{system}
-command, and be sure to capture and report the output of the subprocess's standard error
-to facilitate debugging.
+Third, \textbf{ensure that reused software is robust}. Relying on erratic third
+party libraries or software is a recipe for tears. Prefer software that follows
+good software development practices, is open for support questions, and is
+available from a stable repository using your package manager.
+
+Exercise caution especially when transitioning across languages or using
+separate executables, as they tend to be especially sensitive to operating
+systems, environments, and locales.
 
@@
 
-Programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build
-to compile code, deploy applications, and automate other tasks.
+To compile code, deploy applications, and automate other tasks, 
+programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build.
 These tools can also be used to manage runtime environments,
@@
 and install or upgrade them if they are not.
+As mentioned in the previous rule,
+a package manager can mitigate some of the difficulties in software reuse.
 

### 2017-01-11 robust-software.tex
Removed authentication sentence for #33; added web service mention for #31

@@
 systems~\cite{brown2013}. Even introducing libraries in the same programming
-language can rely on software installed in the environment.
+language can rely on software installed in the environment, and the problem
+becomes much more difficult when relying on executables, or even on web
+services.
 
@@
 it may not be worth the pain of integration. Reuse software that offers some
-measurable improvement to your build environment.
+measurable improvement to your project.
 
@@
 dependent software, or download it during installation using your
-package manager.
+package manager. If the executable requires internet access, check for that
+early in execution.
 
@@
 good software development practices, is open for support questions, and is
-available from a stable repository using your package manager.
+available from a stable location or repository using your package manager.
 
@@
   not sufficient, for software that needs to interact with services
-  such as database and web servers.  Such software is often stopped
-  and started under automatic control, may need to authenticate, etc.,
-  all of which are out of scope of this paper.
+  such as databases and web servers, or coordinates multiple hosts or interfaces, 
+  and is out of scope of this paper.
 

### 2017-01-11 robust-software.tex
Removed inter-institutional collaborations as per #34

@@
 engineering approaches, have high standards of reproducibility, use good testing
-practices, and foster inter-institutional collaborations that grow into strong user
-bases. In the bioinformatics community, Bioconductor and Galaxy follow this
+practices, and foster strong user
+bases through constantly evolving, clearly documented, useful, and useable software.
+In the bioinformatics community, Bioconductor and Galaxy follow this
 path~\cite{gentleman2004,afgan2016}.

### 2017-01-11 robust-software.tex
Combined all references to documentation #38 . Moved non-documentation parts around the manuscript into appropriate sections.

@@
 
-\section{Have a README that briefly explains what the software does and what its dependencies are.}
+\section{Document your code and usage}
 
-The README is the first stop for most new users.
+Numerous resources exist for writing high quality
+documentation~\cite{karimzadeh2016} and so we do not cover it here, except for
+two minimal elements: the README and usage.
+
+The README is the first stop for most new users, usually available even before
+the software is installed.
 At a minimum, it needs to get a new user started and point them towards more
-help, if they need it. 
+help, if they need it. Usage is a terse, informative command-line help message that
+guides the user in the correct use of the software.
+
 Numerous guidelines exist on how to write good
@@
 their key common features are listed below. 
+The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
+is a good model to emulate.
 
-\textbf{Explain what the software does:} At the beginning of the README,
-explain what the software does in one or two sentences.
+\begin{description}
+
+\item[\textbf{Explain what the software does.}] 
 There's nothing more frustrating
@@
 
-\textbf{List required dependencies:} Often, software depends on
-specific versions of libraries, modules, or operating systems. 
-This is
-entirely reasonable as long as it is properly documented. Include the name and
-version number for each dependency.  Several methods exist for acquiring
-required packages: either provide the commands necessary to download the
-dependencies; link to the software homepage; or (our preference) use a
-dependency manager. We address dependencies in more detail in rule 5.
+\item[\textbf{List required dependencies.}] Often, software depends on
+specific versions of libraries, modules, or operating systems. Include the name and
+version number for each dependency or a link to the requirements file. 
+We address dependencies in more detail in rule 5.
 
-\textbf{Provide compilation/installation instructions:}
+\item[\textbf{Provide compilation/installation instructions:}]
 If the software needs to be compiled
-or installed, list those instructions in the README. Also
-mention if you recommend they use a pre-compiled binary instead
-through a system such as pip, yum, or apt.
+or installed, list those instructions in the README.
 
-\textbf{List input and output files}: All possible input and output files
-should be listed in this section. Do the files use a particular
-standard, an extension of an existing format, or your own
-format?  If they use a standard format, link to the specification and
+\item[\textbf{List input and output files.}] All possible input and output files
+should be listed in this section. 
+If they use a standard format, link to the specification and
 version. If you extend the standard or have your own format, define it
 explicitly, listing all the required fields and acceptable values.
-(You get bonus points if you include a script to convert between
-a standard format and your file format). If there is no rigorous format
-(which is common with log files), show at least a few lines from an
+If there is no rigorous format, as is common with log files, show at least a
+few lines from an
 example file and explain what the sections mean.
 
-Input files and their formats are included in most documentation, but
-intermediate, auxiliary, and log files are often missing.
-\emph{All} files should be listed in the README, even those considered self-explanatory. Log
+\emph{All} files should be listed, even those considered self-explanatory. Log
 files are often full of valuable information that can be
@@
 
-\textbf{State attributions and licensing:} Attributions are how you credit
+\item[\textbf{State attributions and licensing.}] Attributions are how you credit
 your main contributors; licenses are how others may use and
-credit your work. These files should be available \emph{before}
-the software is installed (or even downloaded).
+credit your work.
 Leave no
@@
 needs to credit you. If your software is not open source, state that clearly.
-Attributions can also contain a list of ``expert'' users
-that can be contacted if new users have problems with the software.
+\end{description}
 
-The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
-is a good model:
-it explains the software's purpose,
-tells readers where to get help,
-and includes links to a CITATION file (explaining how to cite the project)
-and the license file.
-
-\section{Tell the user what could be done and what was actually done.}
-
-A robust program should
+In addition to a README, a robust program should
 \textbf{print usage information when launching from the command line that explains the software's features}.
 Usage information provides the first line of help for both first-time and
-experienced users of command-line applications.
-Ideally, usage is a terse, informative command-line help message that
-guides the user in the correct use of the software. Terseness is
-important: usage that extends for multiple screens is a nuisance, especially when
-printed to standard error instead of standard output (where it can
-easily be paged).  Long output is also much less likely to be read{\ldots}
+experienced users of command-line applications. Terseness is
+important: usage that extends for multiple screens is 
+difficult to read or refer to on the fly.
 
-Usage should provide all of the information necessary to run the
-software. It is invoked either by running the software without
+Usage is invoked either by running the software without
 any arguments, running it with incorrect arguments, or by
-explicitly choosing a help or usage option.
+explicitly choosing a help option.
 More standard command-line behaviours are detailed in \cite{Seemann2013}.
@@
 POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for usage. 
-An example of good usage is GNU's \texttt{mkdir} command, which makes
-new directories:
-
-\begin{small}
-\begin{verbatim}
-$ mkdir --help
-Usage: mkdir [OPTION]... DIRECTORY...
-Create the DIRECTORY(ies), if they do not already exist.
-
-Mandatory arguments to long options are mandatory for short options too.
-  -m, --mode=MODE   set file mode (as in chmod), not a=rwx - umask
-  -p, --parents     no error if existing, make parent directories as needed
-  -v, --verbose     print a message for each created directory
-  -Z, --context=CTX  set the SELinux security context of each created
-                      directory to CTX
-      --help     display this help and exit
-      --version  output version information and exit
-
-Report mkdir bugs to bug-coreutils@gnu.org
-GNU coreutils home page: <http://www.gnu.org/software/coreutils/>
-General help using GNU software: <http://www.gnu.org/gethelp/>
-For complete documentation, run: info coreutils 'mkdir invocation'
-\end{verbatim}
-\end{small}
-
-The key features of this usage message are:
+The key features of a usage message are:
 
@@
   address, web site, or manual page, there should be an indication
-  where the user can go to find out more about the software.
+  where the user can go to find out more.
 
@@
 
-The usage message tells users what the program could do.
-It is equally important for the program to tell users what it actually did.
-Accordingly,
-when the program starts, it should \textbf{echo all parameters and software
-versions to standard out or a log file alongside the results}. This
-feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
-
-(And as a corollary
-if the software is configurable in any way,
-it should \textbf{check that all input values are in a reasonable range at startup}. 
-Few things are as annoying as having a program announce after running for two hours
-that it isn't going to save its results
-because the requested directory doesn't exist.)
+Documentation beyond the README and usage is up to the developer's discretion.
+We think it is very important for developers to document their work, but our
+experience is that people are unlikely do it during normal development, and we
+don’t want to recommend practices that people won’t actually adopt. However, it
+is worth noting that software that is widely used and contributed to has and
+enforces the need for good documentation~\cite{gentleman2004}.
 
@@
 or verbose output.
+
+\textbf{Check that all input values are in a reasonable range at startup}. 
+Few things are as annoying as having a program announce after running for two
+hours that it isn't going to save its results because the requested directory
+doesn't exist.
 
@@
 
+The usage message tells users what the program could do.
+It is equally important for the program to tell users what it actually did.
+Accordingly,
+when the program starts, it should \textbf{echo all parameters and software
+versions to standard out or a log file alongside the results}. This
+feature supports greater reproducibility because any result can be
+replicated with only the previous output files as reference.
+
 Given a set of parameters and a dataset, \textbf{a particular version of a program
@@
 
-\item[\textbf{Documentation}] We think it is very important for
-  developers to document their work, but our experience is that people
-  are unlikely do it during normal development, and we don't want to
-  recommend practices that people won't actually adopt. However, it is worth
-  noting that software that is widely used and contributed to has and enforces
-  the need for good documentation~\cite{gentleman2004}.
-  Numerous resources exist for writing high quality
-  documentation~\cite{karimzadeh2016} and so we do not cover it here.
-
 \item[\textbf{Services}] The rules we present above are necessary, but

### 2017-01-11 robust-software.tex
Added reference to git paper

@@
 controlled release that range from as simple as adding an appropriate
-commit message or tag to version control, to official releases alongside
+commit message or tag to version control~\cite{blischak2016}, to official releases alongside
 code on Bitbucket or GitHub, to depositing into a

### 2017-01-13 response-to-reviewers.md

> Reviewer 1
> 
> This manuscript provides rules aimed at novice programmers to help
> them write more robust bioinformatics programs. The rules are all
> sensible, and biologists turned programmers (and some people with
> formal software engineering training) would do well to read them.
> 
> With any set of guidelines, and especially guidelines about
> programming, there is room for subjectivity and differences of
> opinion, so here are mine. The authors are free to include or exclude
> these comments in their revision.

>  Robustness
>
>The rules are framed in terms of increasing robustness.
>
> The piece starts out defining robustness in terms of portability,
> specifically outside the programmer's own environment - but then goes
> on to define it more broadly in terms of multiple criteria such as
> being kept under version control (which I don't think is a necessary
> condition for robustness - many robust programs were made before VCSs
> although I think using VCSs is excellent advice for a modern
> programmer)
> 
> The definition here is somewhat different from Wikipedia's, which defines it as:
> 
> "the ability of a computer system to cope with errors during execution[1][2] and cope with erroneous input."
> 
> I am not sure of the best name for the property the authors are aiming at, but it seems some combination of robustness, maintainability, quality, predictability - essentially better software.

Thanks for the feedback. We have clarified and condensed our definition of 'robust' software in the introduction, in particular removing the reference to version control.

>     Documentation and maintainability.
> 
> I find the relegation of documentation to an additional note rather
> than a rule surprising. It's also telling the authors don't call out
> maintainability as a desirable property (except in passing in the
> conclusions). If software is not thrown away, it frequently needs to
> be updated - e.g. to accomodate changes in input file types. This
> maintenance task often falls on people who are not the original
a> uthor. If software does not have adequate documentation, it will be
> harder to maintain (either by the original author or by others).

We've promoted documentation to a full rule (rule #1 in fact).

>    Good software development practice
> 
> The rules do not mention anything about good software engineering
> practice, such as structured programming. While it's good to not be
> too specific and dictate specific fads or paradigms (and the intended
> audience is individual programmers not teams), no one would disagree
> that some kind of good structured programming practices are
> required to write programs where clarity, maintainability,
> extensibility and quality are of concern.
> 
> In general, the program should not repeat large pieces of code with
> the same logic - these should be abstracted into
> subroutines/functions/methods/classes etc. The program should not be a
> giant monolithic main() method. Behavior of subroutines should not be
> controlled through globals. etc.
> 
> While it's not necessary to replicate Knuth's three volumes, I think
> good software engineering practice deserves to be a rule, or at least
> called out.

 We've added a paragraph to the introduction to mention code quality, maintainability, unit tests, and version control as important but out of scope.

> Reproducibility
> 
> Reproducibility of software and workflows is increasingly in the
> spotlight. There are passing mentions of reproducibility, and it is
> implicit in many of the guidelines, but it would be good to call it
> out explicitly as a goal beyond making the software robust. For
> example, in the list on p2.
> 
> People like Titus Brown have written a lot about this in blogs. It
> would be useful to reference some of this.

We've added more mention of reproducibility and citations to posts written by C. Titus Brown and by Gaël Varoquaux.

> README
> 
> Excellent advice, although there is some overlap with 2.

These two rules have been combined.

> Usage info
> 
> Good advice, but I wonder if 2 and 8 could be combined?

We consider usage to be part of the documentation, but a section of this rule referred to checking input parameters and echoing out parameters to the command line. These sections have been moved to the rules on controlling common operations and producing identical results.

> Versioning
> 
> All good advice.
> 
> As the authors note, GitHub makes this very easy. The advice is a bit succinct:
> 
> "as simple as adding an appropriate commit message or tag to version control"
> 
> As this article seems to be aimed as novice programmers who may not
> have been immersed in good, modern, open software development
> practice, it would be a good idea to include more details or pointers here.
> 
> In fact, I would recommend every programmer learns how to use a VCS
> and a hosted VCS like GitHub/GitLab/Bitbucket, to commit early and
> commit often, to use issue trackers, etc. Not just a matter of making
> public releases (which is important) but also as a matter of good
> practice even for software not intended to be released.

We think it's important too, but it's a tough sell in some crowds. We've added a reference to another educational article, "A Quick Introduction to Version Control with Git and GitHub".

> Reuse software
> 
> At first I assumed this section was about reusing software
> libraries, but it transpired to be about reusing executable programs.
> 
> This seems an odd guideline. While it's certainly convenient to launch
> a program using commands like 'system' in perl, it's mostly better to
> use a library (or service). In fact the authors agree with this in 5,
> in which case why are we recommending them do this here?
> 
> The advice seems a bit confused in general. It's not really clear if
> the authors are saying if it's a good idea or bad idea to call external programs.
> 
> I think general advice to 'take care when calling external programs'
> may be more appropriate.

We've re-written this section more mindful of the package manager rule to follow and more explicitly pushing back against using external executables and services.

> built utility / package manager
> 
> Excellent advice that is all too often ignored.
> 
> As noted above the presentation in combo with 4 is
> confusing. Generally a package manager is used when an external
> library, module or package is re-used. But as noted there is generally
> no advice on reusing these artefacts, as 4 is about running programs.

We believe we have addressed the discrepencies between these two rules with our re-write.

> Do not require root privileges
> 
> I certainly wouldn't contradict this but I wonder why it's called
> out. There are many possible inadvisable things we've all seen in years of programming, but maybe this could be combined with advice about running external programs (4) or making the software runnable outside a default user environment (7)

(Greg, you have something for this? My motivation is painful experience.)
..................

> Eliminate hard-coded paths
> 
> Excellent advice, and worth calling out as it is such a frequent problem

:)

> Configuration from command line
> 
> Good advice, but could be combined with 7, as they are both about configuration.

................

> Include a small test set
> 
> Good advice, but there is no advice about including software unit
> tests. This is good general software engineering advice, and the
> existence of simple cloud-based CI frameworks like Travis that
> integrate with GitHub make it very easy to ensure that every change to
> a program does not break thing.
> 
> Produce identical results when given identical inputs
> 
> Generally good advice, but perhaps overly specific. There is a lengthy
> discussion about what to do in the case of algorithms that involve
> random numbers. What about software that calls external services?
> 
> In my opinion this is a judgment call by the software
> author. Certainly, if there is no deliberate randomization or calling
> of external services then nondetermistic output would be troubling.

If you call external services, then according to the (new!) guideline in "Reuse software", that software should itself be robust and therefore produce the same results on every run. Some services are updated to change, add or remove data or results and this is unfortunate, but expected. Two subsequent runs of the software, barring any other changes, should still produce identical results.

In addition, software that doesn't produce identical results is much more difficult to test. Depending on the amount of variation, it can be nearly impossible.

..................
(Greg, any other changes you want to make here?)

> What we left out
> 
> As mentioned above, documentation is important, and some of the
> existing 10 could be merged to make room for this.

We merged all mentions of documentation into the first rule.

> For services, I am not sure why authentication is mentioned, even in
> passing. It doesn't really pertain to robustness per se.

This reference has been removed.

> Reviewer 3
> 
> Comments to the author:
> 
> This paper sets out to describe 10 simple rules which, if applied, can contribute to making software research more robust. It provides a useful set of guidelines that should generally improve the robustness of research software and it should be of interest to most graduate students who are writing software in the process of their research.
> 
> I recommend the publication of this paper but would request the addition or additional emphasis on one point which is not called out as a rule but which is mentioned in passing in several places within the paper. In particular I am referring to the recommendation to put all software into a source control system. As this is an opinion paper, I do not think this is a necessary change for publication (in my opinion it’s a vital rule, that is just my opinion) but it is referenced in several places and in particular the author’s definition of “robust software” includes “...is kept under version control”. As this is an issue frequently ignored or addressed “too late” by graduate students I would like to see this get additional emphasis, even if it is not promoted to being a full rule.

..............

> The remainder of my comments are meant to provide (hopefully) useful feedback to the authors in the line of suggestions that may enhance the manuscript.
> 
> In describing successful software research projects the authors state “...who have adopted strong software engineering approaches, have high standards of reproducibility, use good testing practices, and foster inter-institutional collaborations that grow into strong user bases”.. All of these points can be clearly tied to their rules with the exception of the final statement about inter-institutional collaborations. While such collaborations are desirable, particularly to establish project continuity and longevity, they are not strictly necessary for robustness and this is not further referenced in the manuscript.

We removed this line from the text.

> In several cases, it seems that the paper might violate the first rule of “The 10 simple rules for writing a 10 simple rules paper” (http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003858) with multiple rules combined into a single rule.

We reviewed all the sections you mentioned. In general, we believe that most of our *bolded* sections within each rule are refinements of the rule and not separate rules. We address each one of your concerns in-line below.

> For example rule 8 combines putting parameters on the command line with echoing all parameters to the log file to support reproducibility and provenance.

................

> Similarly rule 5 covers both automating the build process and automating the deployment and installation.

..............

> Rule 4 also tends to stuff multiple rules (bolded) into one rule (( limit dependencies, ensure dependencies are available) these 2 are closely related), with “ use native functions” which is a very different idea and not what I would expect from a ‘reuse’ rule.

We removed this last 'native functions' rule and replaced it with something more appropriate.

> Finally the ‘What we left out’ section implicitly defines an additional rule (documentation). In this particular case there could have been a single rule (Document your code and usage) that could encompass the points of both Rules 1 and 2 (Readme, usage) as well as the points from ‘What we left out’.

We combined all mentions of documentation into a single rule.

> In rule 8, I would tend to suggest generalizing the rule somewhat to suggest externalizing parameters rather than necessarily requiring them to be available as command line options. For example if a configuration file is used to provide all parameter values, this could accomplish the task of separating the parameters out from the code as well as providing a record of the values used for a run as they describe through echoing parameters to a log file.

....

### 2017-01-13 robust-software.tex
Updating rule numbers and references, and making space for a new rule about version control.
1.  Added a new "Rule 1: Use version control" as a placeholder.
2.  Updated rule numbers for all other rules.
3.  Checked cross-rule references.
4.  Added new commands `\rulemajor` and `\ruleminor` to make rules and sub-rules easier to find.
5.  Updated the checklist to match the new rule set.
6.  Removed the "what we left out" section - it now had only one entry, so it looked a little lonely, and the full list of what we've left out would be very long.
7.  Removed the "bonus rules" section of the chec

--- 2017-01-11 robust-software.tex
@@
 \newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}
+\newcommand{\rulemajor}[1]{\section{#1}}
+\newcommand{\ruleminor}[1]{\textbf{#1}}
 
@@
 
-Note: practices \emph{not} included in the list below include version control,
-code review, unit testing, and other practices essential to maintainability and
-sustainable software development.
-These elements are essential to building software of any kind
+\rulemajor{Rule 1: Use version control.}
+
+FIXME: placeholder while renumbering.
+
+Our focus in this short guide is on the software that is already built,
+not on the process used to build it.
+However,
+we believe that version control is essential to sustainable software development
 \cite{wilson2014,wilson2016}.
-However,
-our focus in this short guide is on the software that is already built,
-not on the process used to build it.
 
-\section{Document your code and usage}
+\rulemajor{Rule 2: Document your code and usage}
 
@@
 
-\item[\textbf{Explain what the software does.}] 
+\item[\ruleminor{Explain what the software does.}] 
 There's nothing more frustrating
@@
 
-\item[\textbf{List required dependencies.}] Often, software depends on
+\item[\ruleminor{List required dependencies.}] Often, software depends on
 specific versions of libraries, modules, or operating systems. Include the name and
 version number for each dependency or a link to the requirements file. 
-We address dependencies in more detail in rule 5.
+We address dependencies in more detail in Rule~5.
 
-\item[\textbf{Provide compilation/installation instructions:}]
+\item[\ruleminor{Provide compilation/installation instructions:}]
 If the software needs to be compiled
@@
 
-\item[\textbf{List input and output files.}] All possible input and output files
+\item[\ruleminor{List input and output files.}] All possible input and output files
 should be listed in this section. 
@@
 
-\item[\textbf{State attributions and licensing.}] Attributions are how you credit
+\item[\ruleminor{State attributions and licensing.}] Attributions are how you credit
 your main contributors; licenses are how others may use and
@@
 In addition to a README, a robust program should
-\textbf{print usage information when launching from the command line that explains the software's features}.
+\ruleminor{print usage information when launching from the command line that explains the software's features}.
 Usage information provides the first line of help for both first-time and
@@
 
-\item[\textbf{The syntax for running the program}] This includes the
+\item[\ruleminor{The syntax for running the program}] This includes the
   name of the program and defines the relative location of optional
@@
 
-\item[\textbf{Description}] Similar to the README, the description
+\item[\ruleminor{Primary function}] Similar to the README, the description
   reminds users of the software's primary function.
 
-\item[\textbf{Most commonly used arguments, a description of each, and
+\item[\ruleminor{Most commonly used arguments, a description of each, and
     the default values}] Not all arguments need to appear in the
@@
 
-\item[\textbf{Where to find more information}] Whether it's an email
+\item[\ruleminor{Where to find more information}] Whether it's an email
   address, web site, or manual page, there should be an indication
@@
 
-\item[\textbf{Printed to standard output}] Usage should be displayed
+\item[\ruleminor{Printed to standard output}] Usage should be displayed
   on standard output so that it can be piped into \texttt{less},
@@
 
-\item[\textbf{Exit with an appropriate exit code}] When usage is
+\item[\ruleminor{Exit with an appropriate exit code}] When usage is
   invoked by providing incorrect arguments, the program should exit
@@
 
-\section{Make common operations easy to control.}
+\rulemajor{Rule 3: Make common operations easy to control.}
 
@@
 Programs should therefore
-\textbf{allow the most commonly changed parameters to be configured from the command line}.
+\ruleminor{allow the most commonly changed parameters to be configured from the command line}.
 
@@
 
-\textbf{Check that all input values are in a reasonable range at startup}. 
+\ruleminor{Check that all input values are in a reasonable range at startup}. 
 Few things are as annoying as having a program announce after running for two
@@
 To make programs even easier to use,
-\textbf{choose reasonable defaults where they exist}
-and \textbf{set no defaults at all when there aren't any reasonable ones}.
+\ruleminor{choose reasonable defaults where they exist}
+and \ruleminor{set no defaults at all when there aren't any reasonable ones}.
 You can set reasonable default values
@@
 
-\section{Version your releases.}
+\rulemajor{Rule 4: Version your releases.}
 
@@
 output as well as any errors.
-\textbf{Increment your version number every time you release your software to
+\ruleminor{Increment your version number every time you release your software to
 other people}.
@@
 
-\textbf{The version of your software should be easily available by 
+\ruleminor{The version of your software should be easily available by 
 supplying \texttt{-\/-version} or \texttt{-v} on the command line}. This command should print
 the software name and version number, and it should
-also be \textbf{included in all of the program's output}, particularly debugging
+also be \ruleminor{included in all of the program's output}, particularly debugging
 traces.  If someone needs help, it's important that they be able to tell
@@
 and won't see any benefits from upgrading.
-A program's authors should therefore \textbf{ensure that old released versions
+A program's authors should therefore \ruleminor{ensure that old released versions
 continue to be available.}
@@
 
-\section{Reuse software (within reason).}
+\rulemajor{Rule 5: Reuse software (within reason).}
 
@@
 First, 
-\textbf{make sure that you really need the auxiliary program}. If you are
+\ruleminor{make sure that you really need the auxiliary program}. If you are
 executing GNU sort instead of figuring out how to sort lists in Python,
@@
 
-Second, if launching an executable, \textbf{ensure the appropriate software and version is available}.
+Second, if launching an executable, \ruleminor{ensure the appropriate software and version is available}.
 Either allow the user to
@@
 
-Third, \textbf{ensure that reused software is robust}. Relying on erratic third
+Third, \ruleminor{ensure that reused software is robust}. Relying on erratic third
 party libraries or software is a recipe for tears. Prefer software that follows
@@
 
-\section{Rely on build tools and package managers for installation.}
+\rulemajor{Rule 6: Rely on build tools and package managers for installation.}
 
@@
 and install or upgrade them if they are not.
-As mentioned in the previous rule,
+As mentioned in Rule~5,
 a package manager can mitigate some of the difficulties in software reuse.
@@
 developers should
-\textbf{document all dependencies in a machine-readable form}.
+\ruleminor{document all dependencies in a machine-readable form}.
 Package managers like apt and yum are available on most Unix-like systems, and
@@
 Conversely, developers should
-\textbf{avoid depending on scripts and tools which are not available as packages}.
+\ruleminor{avoid depending on scripts and tools which are not available as packages}.
 In many cases, a program's author may not realize that some tool was built locally, and
@@
 
-\section{Do not require root or other special privileges.}
+\rulemajor{Rule 7: Do not require root or other special privileges to install or run.}
 
@@
 very unusual circumstances,
-\textbf{packages should not require root privileges to set up or use}.
+\ruleminor{packages should not require root privileges to set up or use}.
 
@@
 Developers should therefore
-\textbf{allow packages to be installed in an arbitrary location},
+\ruleminor{allow packages to be installed in an arbitrary location},
 e.g., under a user's home directory in
@@
 Lightweight virtualization containers like Docker make this much easier as well, 
-or simply \textbf{ask another person to try and build your software}.
+or simply \ruleminor{ask another person to try and build your software before releasing it}.
 
-\section{Eliminate hard-coded paths.}
+\rulemajor{Rule 8: Eliminate hard-coded paths.}
 
@@
 For these reasons, users should be able to
-\textbf{set the names and locations of input and output files as command-line parameters}.
+\ruleminor{set the names and locations of input and output files as command-line parameters}.
 This rule applies to reference data sets as well as the user's own
@@
 A corollary to this rule is
-\textbf{do not require users to navigate to a particular directory to do their work},
+\ruleminor{do not require users to navigate to a particular directory to do their work},
 since ``where I have to be'' is just another hard-coded path.
@@
 
-\section{Include a small test set that can be run to ensure the software is actually working.}
+\rulemajor{Rule 9: Include a small test set that can be run to ensure the software is actually working.}
 
@@
 
-In order to be useful, \textbf{make the test script easy to find and run}. 
+In order to be useful, \ruleminor{make the test script easy to find and run}. 
 Many build systems will also run unit tests if provided them at compile time.
@@
 
-Equally, \textbf{make the test script's output easy to interpret}. Screens
+Equally, \ruleminor{make the test script's output easy to interpret}. Screens
 full of correlation coefficients do not qualify: instead, the script's
@@
 
-\section{Produce identical results when given identical inputs.}
+\rulemajor{Rule 10: Produce identical results when given identical inputs.}
 
@@
 Accordingly,
-when the program starts, it should \textbf{echo all parameters and software
+when the program starts, it should \ruleminor{echo all parameters and software
 versions to standard out or a log file alongside the results}. This
@@
 
-Given a set of parameters and a dataset, \textbf{a particular version of a program
+Given a set of parameters and a dataset, \ruleminor{a particular version of a program
 should produce the same results every time it is run}
@@
 approximate random numbers. Setting the seed to a consistent value
-can remove randomness between runs. \textbf{Allow the user to optionally provide
+can remove randomness between runs. \ruleminor{Allow the user to optionally provide
 the random seed as an input parameter}, thus rendering the program deterministic
@@
 using clock time), echo it to the output for re-use later. 
-If setting the seed is not possible, \textbf{make sure the acceptable tolerance is
+If setting the seed is not possible, \ruleminor{make sure the acceptable tolerance is
 known and detailed in documentation and in the tests}.
-
-\section*{What We Left Out}
-
-Any short list of rules for developing robust software will
-necessarily be incomplete.   Our more significant omissions
-include:
-
-\begin{description}
-
-\item[\textbf{Services}] The rules we present above are necessary, but
-  not sufficient, for software that needs to interact with services
-  such as databases and web servers, or coordinates multiple hosts or interfaces, 
-  and is out of scope of this paper.
-
-\end{description}
 

### 2017-01-17 robust-checks.tex
Submission to arXiv

@@
 \begin{easylist}[checklist]
-& Have a README
-&& Explain what the software does
-&& List required dependencies
-&& Provide compile/installation instructions
-&& List input and output files
-&& State attributions and licensing
-& Print usage information from the command line
-&& Include:
-&&& the syntax for running the program in GNU/POSIX format
-&&& a one line description
-&&& the most commonly used arguments, a description of each, and the default values
-&&&& Where to find more information
-&& Print to standard output
+& Use version control.
+
+& Document your code and usage
+&& Explain what the software does.
+&& List required dependencies.
+&& Provide compilation/installation instructions.
+&& List input and output files.
+&& State attributions and licensing.
+&& Print usage information when launching from the command line.
+&& Desribe the syntax for running the program.
+&& Explain the software's primary function.
+&& Describe the most commonly used arguments, with a description of each and default values.
+&& Tell users where to find more information.
+&& Print to standard output.
 && Exit with an appropriate exit code
-& Version your releases
-&& Increment your version number every time you release your software to other people
-&& Print when supplying \texttt{-\/-version} or \texttt{-v} on the command line
-&& Include version number in output
-&& Deposit releases in a stable location so they are available in perpetuity
-& Reuse software (within reason)
-&& Make sure that you really need the auxiliary program
-&& Check for dependent software and version early in execution
-&& Use native functions for starting other processes
-& Use a build utilty and package manager
-&& Document \emph{all} dependencies, preferably in a machine-readable form
-&& Avoid depending on scripts and tools which are not available as packages
-& Do not require root or other special privileges
-&& Allow packages to be installed in an arbitrary location
-&& Ask another person to try and build your software
-& Eliminate hard-coded paths
-&& Set the names and locations of input and output files as command-line parameters
-&& Do not require users to navigate to a particular directory to do their work
-& Allow configuration of all useful parameters from the command line.
-&& Choose reasonable defaults where they exist
-&& Set no defaults at all when there aren't any reasonable ones
-&& Echo all parameters and software versions to standard out or a log file alongside the results
-&& Check that all input values are in a reasonable range near startup
+
+& Make common operations easy to control.
+&& Allow the most commonly changed parameters to be configured from the command line.
+&& Check that all input values are in a reasonable range at startup.
+&& Choose reasonable defaults where they exist.
+&& Set no defaults at all when there aren't any reasonable ones.
+
+& Version your releases.
+&& Increment your version number every time you release your software to other people.
+&& Print the version number when given the appropriate flag.
+&& Include the version number in all of the program's output.
+&& Ensure that old released versions continue to be available.
+
+& Reuse software (within reason).
+&& Make sure that you really need auxiliary programs.
+&& Ensure the appropriate software and version is available.
+&& Ensure that reused software is robust.
+
+& Rely on build tools and package managers for installation.
+&& Document all dependencies in a machine-readable form.
+&& Avoid depending on scripts and tools which are not available as packages.
+
+& Do not require root or other special privileges to install or run.
+&& Packages should not require root privileges to set up or use.
+&& Allow packages to be installed in an arbitrary location.
+&& Ask another person to try and build your software before releasing it.
+
+& Eliminate hard-coded paths.
+&& Set the names and locations of input and output files as command-line parameters.
+&& Do not require users to navigate to a particular directory to do their work.
+
 & Include a small test set that can be run to ensure the software is actually working.
-&& Tests are easy to find and run
-&& Test results are easy to interpret
-& Produce identical results when given identical inputs
-&& For randomized algorithms: allow the user to optionally provide the seed as an input parameter; or
-&& Make sure the acceptable tolerance is known and detailed in documentation and in the tests
-\end{easylist}
+&& Make the test script easy to find and run.
+&& Make the test script's output easy to interpret.
 
-\section*{Bonus points}
-
-\begin{easylist}[checklist]
-& Conform to command-line conventions~\cite{Seemann2013}
-& Write high quality documentation~\cite{karimzadeh2016}
+& Produce identical results when given identical inputs.
+&& Echo all parameters and software versions to standard out or a log file alongside the results.
+&& A particular version of a program should produce the same results every time it is run.
+&& Allow the user to optionally provide the random seed as an input parameter.
+&& Make sure the acceptable tolerance is known and detailed in documentation and in the tests.
 
@@
 
-\bibliography{robust-checks}
-
 \end{document}

### 2017-01-17 robust-software.tex

@@
 
-FIXME: placeholder while renumbering.
-
 Our focus in this short guide is on the software that is already built,
-not on the process used to build it.
-However,
-we believe that version control is essential to sustainable software development
+not on the process used to build it,
+but we believe that version control is essential to sustainable software development
 \cite{wilson2014,wilson2016}.
+In particular,
+developers will struggle to understand what they have actually built,
+what it actually does,
+and what they have actually released
+without some mechanical way to keep track of changes.
+They should therefore
+\ruleminor{put everything created manually into version control as soon as it is created},
+including programs, original field observations, and the source files for papers.
+Files that can be regenerated at need,
+such as the binaries for compiled programs or intermediate files generated during data analysis,
+should not be versioned;
+instead,
+it is often more sensible to use an archiving system for them,
+and store the metadata describing their contents in version control instead~\cite{noble2009}.
 

### 2017-01-17 robust-software.tex
Shortened documentation rule

@@
 
-Numerous resources exist for writing high quality
-documentation~\cite{karimzadeh2016} and so we do not cover it here, except for
-two minimal elements: the README and usage.
-
-The README is the first stop for most new users, usually available even before
-the software is installed.
-At a minimum, it needs to get a new user started and point them towards more
-help, if they need it. Usage is a terse, informative command-line help message that
+How to write high quality documentation has been described elsewhere~\cite{karimzadeh2016}
+and so here we only cover two minimal types: the README and usage. 
+The README is usually available even before the software is installed, 
+and exists
+to get a new user started and point them towards more
+help. Usage is a terse, informative command-line help message that
 guides the user in the correct use of the software.
 
-Numerous guidelines exist on how to write good
-READMEs\cite{Johnson1997,gnustandards};
-their key common features are listed below. 
-The README file for \withurl{khmer}{https://github.com/dib-lab/khmer/blob/master/README.rst}
-is a good model to emulate.
+Numerous guidelines exist on how to \ruleminor{write a good
+README file}~\cite{Johnson1997,gnustandards}.
+Some key common features are listed below.
 
-\begin{description}
+\begin{enumerate}
 
-\item[\ruleminor{Explain what the software does.}] 
+\item \textit{Explain what the software does}.
 There's nothing more frustrating
@@
 
-\item[\ruleminor{List required dependencies.}] Often, software depends on
-specific versions of libraries, modules, or operating systems. Include the name and
-version number for each dependency or a link to the requirements file. 
+\item \textit{List required dependencies}.
 We address dependencies in more detail in Rule~5.
 
-\item[\ruleminor{Provide compilation/installation instructions:}]
-If the software needs to be compiled
-or installed, list those instructions in the README.
+\item \textit{Provide compilation or installation instructions}.
 
-\item[\ruleminor{List input and output files.}] All possible input and output files
-should be listed in this section. 
-If they use a standard format, link to the specification and
-version. If you extend the standard or have your own format, define it
-explicitly, listing all the required fields and acceptable values.
-If there is no rigorous format, as is common with log files, show at least a
-few lines from an
-example file and explain what the sections mean.
+\item \textit{List all input and output files},
+even those considered self-explanatory. 
+Link to specifications for standard formats
+and list the required fields and acceptable values in other files.
+If there is no rigorous format, explain what the sections mean.
+These files are often full of valuable information that can be
+mined for the user's specific purpose.
 
-\emph{All} files should be listed, even those considered self-explanatory. Log
-files are often full of valuable information that can be
-mined for the user's specific purpose. If your users might need to know,
-``Does this program report the percentage of reads trimmed to remove
-adapter sequences?'' they should be able to check the README and confidently
-say, ``Yes, it is in the log file''.
+\item \textit{List a few example commands} to get a user started quickly.
 
-\item[\ruleminor{State attributions and licensing.}] Attributions are how you credit
-your main contributors; licenses are how others may use and
-credit your work.
-Leave no
-question in anyone's mind about whether your software can be used
-commercially, how much modification is permitted, and how other software
-needs to credit you. If your software is not open source, state that clearly.
-\end{description}
+\item \textit{State attributions and licensing}. Attributions are how you credit
+your contributors; licenses dictate how others may use and
+need to credit your work.
+\end{enumerate}
 
-In addition to a README, a robust program should
-\ruleminor{print usage information when launching from the command line that explains the software's features}.
-Usage information provides the first line of help for both first-time and
-experienced users of command-line applications. Terseness is
-important: usage that extends for multiple screens is 
+The program should also 
+\ruleminor{print usage information} when launching from the command line.
+Usage provides the first line of help for both new and experienced users. 
+Terseness is important: usage that extends for multiple screens is 
 difficult to read or refer to on the fly.
-
-Usage is invoked either by running the software without
-any arguments, running it with incorrect arguments, or by
-explicitly choosing a help option.
-More standard command-line behaviours are detailed in \cite{Seemann2013}.
 
@@
 POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for usage. 
-The key features of a usage message are:
+More standard command-line behaviours are detailed in \cite{Seemann2013}.
+The key features of a usage message include:
 
-\begin{description}
+\begin{enumerate}
 
-\item[\ruleminor{The syntax for running the program}] This includes the
-  name of the program and defines the relative location of optional
-  and required flags, arguments and values for execution.  Arguments
-  in {[}square brackets{]} are usually optional. An ellipsis (\ldots
-  e.g. ``{[}OPTION{]}\ldots{}'') indicates that more than one value
-  can be provided.
+\item \textit{Describe the syntax for running the program}, including the
+  name of the program, the relative location of optional
+  and required flags, other arguments, and values for execution.
 
-\item[\ruleminor{Primary function}] Similar to the README, the description
-  reminds users of the software's primary function.
+\item \textit{Give a short description} to remind users of the software's primary function.
 
-\item[\ruleminor{Most commonly used arguments, a description of each, and
-    the default values}] Not all arguments need to appear in the
-  usage, but those most commonly used should be listed. Users will
-  rely on this for quick reference when working with the software.
+\item \textit{List the most commonly used arguments}, a description of each, and
+    the default values.
 
-\item[\ruleminor{Where to find more information}] Whether it's an email
-  address, web site, or manual page, there should be an indication
-  where the user can go to find out more.
+\item \textit{State where to find more information}.
 
-\item[\ruleminor{Printed to standard output}] Usage should be displayed
-  on standard output so that it can be piped into \texttt{less},
-  searched with \texttt{grep}, or compared to the previous version
-  with \texttt{diff}.
+\end{enumerate}
 
-\item[\ruleminor{Exit with an appropriate exit code}] When usage is
-  invoked by providing incorrect arguments, the program should exit
-  with a non-zero code to indicate an error. However, when help is
-  explicitly requested, the software should not exit with an error.
-
-\end{description}
+Usage should be 
+printed to standard output so that it can be 
+combined with other bash utilities like \texttt{grep},
+and it should finish with an appropiate exit code
+depending upon whether it is a response to an error,
+or was explicitly requested.
 
@@
 We think it is very important for developers to document their work, but our
-experience is that people are unlikely do it during normal development, and we
-don’t want to recommend practices that people won’t actually adopt. However, it
+experience is that people are unlikely do it during normal development
+However, it
 is worth noting that software that is widely used and contributed to has and

### 2017-01-18 robust-software.tex

@@
 The README is usually available even before the software is installed, 
-and exists
-to get a new user started and point them towards more
+exists
+to get a new user started, and points them towards more
 help. Usage is a terse, informative command-line help message that
@@
 README file}~\cite{Johnson1997,gnustandards}. 
-Some key common features are listed below.
+At a minimum, your README should:
 
@@
 More standard command-line behaviours are detailed in \cite{Seemann2013}.
-The key features of a usage message include:
+Your software's usage should:
 
@@
 combined with other bash utilities like \texttt{grep},
-and it should finish with an appropiate exit code
-depending upon whether it is a response to an error,
-or was explicitly requested.
+and it should finish with an appropiate exit code.
 

### 2017-01-18 robust-checks.tex
Minor tidying up and reformatting the checklist

@@
 & Use version control.
+&& Put everything created manually into version control as soon as it is created.
 
-& Document your code and usage
-&& Explain what the software does.
-&& List required dependencies.
-&& Provide compilation/installation instructions.
-&& List input and output files.
-&& State attributions and licensing.
-&& Print usage information when launching from the command line.
-&& Desribe the syntax for running the program.
-&& Explain the software's primary function.
-&& Describe the most commonly used arguments, with a description of each and default values.
-&& Tell users where to find more information.
-&& Print to standard output.
-&& Exit with an appropriate exit code
+& Document your code and usage.
+&& Write a good README file.
+&& Print usage information.
 
@@
 && Increment your version number every time you release your software to other people.
-&& Print the version number when given the appropriate flag.
-&& Include the version number in all of the program's output.
+&& Make the version of your software easily available by supplying \texttt{-\/-version} or \texttt{-v} on the command line.
+&& Include the version number in in all of the program's output.
 && Ensure that old released versions continue to be available.
@@
 & Reuse software (within reason).
-&& Make sure that you really need auxiliary programs.
+&& Make sure that you really need the auxiliary program.
 && Ensure the appropriate software and version is available.
@@
 & Do not require root or other special privileges to install or run.
-&& Packages should not require root privileges to set up or use.
+&& Do not require root privileges to set up or use packages.
 && Allow packages to be installed in an arbitrary location.
@@
 && Echo all parameters and software versions to standard out or a log file alongside the results.
-&& A particular version of a program should produce the same results every time it is run.
+&& Produce the same results each time the same version of the program is run with the same inputs.
 && Allow the user to optionally provide the random seed as an input parameter.
-&& Make sure the acceptable tolerance is known and detailed in documentation and in the tests.
+&& Make sure acceptable tolerances are known and detailed in documentation and tests.
 

### 2017-01-19 robust-software.tex
Minor tidying up and reformatting the checklist

@@
 and store the metadata describing their contents in version control instead~\cite{noble2009}.
+
+If you are new to version control,
+it is simplest to treat it as ``a better \withurl{Dropbox}{http://dropbox.com}''
+(or, if you are of a certain age, a better FTP),
+and to use it simply to synchronize files between multiple developers and machines.
+Once you are comfortable working that way,
+you should \ruleminor{use a feature branch workflow}:
+designate one parallel copy (or ``branch'') of the repository as the master,
+and create a new branch from it each time you want to fix a bug or add a new feature.
+This allows work on independent changes to proceed in isolation;
+once the work has been completed and tested,
+it can be merged into the master branch for release.
+
+We also recommend that you \ruleminor{tag your releases}.
+A ``tag'' is a label that marks a particular revision in a repository's history.
+Adding such labels makes it much easier to find desired revisions at a later date,
+and also makes it clear exactly what was released.
 

### 2017-01-19 robust-software.tex
Added example set to rule 9, some other tweaks to grammar and flow.

--- 2017-01-18 robust-software.tex
@@
 which slows the pace of research~\cite{prabhu2011,lawlor2015}.
-Not all software needs to be robust enough to run for other 
+Not all software needs to be robust enough to run for other people,
+but software that is published and distributed to the community should be.
 Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
-hundreds of tools to perform the same task:
+hundreds of tools that perform similar tasks:
 for example,
@@
 
-Our focus in this short guide is on the software that is already built,
-not on the process used to build it,
-but we believe that version control is essential to sustainable software development
+Version control is essential to sustainable software development
 \cite{wilson2014,wilson2016}.
@@
 and store the metadata describing their contents in version control instead~\cite{noble2009}.
+
+If you are new to version control,
+it is simplest to treat it as ``a better \withurl{Dropbox}{http://dropbox.com}''
+(or, if you are of a certain age, a better FTP),
+and to use it simply to synchronize files between multiple developers and machines~\cite{blischak2016}.
+Once you are comfortable working that way,
+you should \ruleminor{use a feature branch workflow}:
+designate one parallel copy (or ``branch'') of the repository as the master,
+and create a new branch from it each time you want to fix a bug or add a new feature.
+This allows work on independent changes to proceed in isolation;
+once the work has been completed and tested,
+it can be merged into the master branch for release.
 
@@
 
-\rulemajor{Rule 5: Reuse software (within reason).}
+\rulemajor{Rule 5: Reuse software (within reason)}
 
@@
 single file in the project's root directory named \texttt{runtests.sh}
-or something equally obvious is a much better solution than documenting
-test cases and requiring people to copy and paste them.
+or something equally obvious is a 
+good solution that lets new users build their analysis from a working script.
+For example, with its distribution, 
+HISAT2 includes a full set of very small files,
+and a 'Getting Started with HISAT2' section in its manual 
+that leads you through the entire data lifecycle~\cite{pertea2016}.
 

### 2017-01-19 response-to-reviewers.md

@@
 
-Thanks for the feedback. We have clarified and condensed our definition of 'robust' software in the introduction, in particular removing the reference to version control.
+Thanks for the feedback. We have clarified and condensed our definition of 'robust' software in the introduction, and have made version control a full rule.
 
@@
 
-We've promoted documentation to a full rule (rule #1 in fact).
+We've promoted documentation to a full rule.
 

### 2017-01-20 response-to-reviewers.md
Added response to reviewers markdown #24

@@
+Dear Editor and Reviewers,
+
+Please find attached our revised paper for "Ten Simple Rules for Making Research Software More Robust".
+We have taken all reviewer comments into account in this revision and have produced a stronger paper.
+Each of the reviewer's concerns is addressed in-line below.
+Attached is also a version of the document with changes highlighted in orange.
+
+Thank you and best regards,
+
+Greg Wilson & Morgan Taschuk
+
 > Reviewer 1
@@
 >
->The rules are framed in terms of increasing robustness.
+> The rules are framed in terms of increasing robustness.
 >
@@
 > maintenance task often falls on people who are not the original
-a> uthor. If software does not have adequate documentation, it will be
+> author. If software does not have adequate documentation, it will be
 > harder to maintain (either by the original author or by others).
 
-We've promoted documentation to a full rule.
+We've promoted documentation to a full rule by merging the README and Usage rules into one and explicitly mentioning other types of documentation with appropriate references.
 
@@
 
-We've added a paragraph to the introduction to mention code quality, maintainability, unit tests, and version control as important but out of scope.
+Thanks for the feedback.
+We are focusing mostly on software that is preparing to be published or otherwise distributed, and so discussing the details of program flow and execution are out of scope.
+We do mention other good software engineering practices, such as source control, testing, and versioning as an introductory taste of software engineering.
 
@@
 
-We think it's important too, but it's a tough sell in some crowds. We've added a reference to another educational article, "A Quick Introduction to Version Control with Git and GitHub".
+We think it's important too, but it's a tough sell in some crowds.
+Version control is its own rule now, and we've added a reference to another educational article, "A Quick Introduction to Version Control with Git and GitHub".
 
@@
 We've re-written this section more mindful of the package manager rule to follow and more explicitly pushing back against using external executables and services.
+We also include web services now for consideration.
 
@@
 
-(Greg, you have something for this? My motivation is painful experience.)
-..................
+With the increasing torrent of 'big data', more and more calculations are outsourced into high performance compute clusters or cloud environments.
+When spawning jobs across tens to thousands of machines that may or may not share any storage, and likely only gives their users privileges in their own space, a requirement for needing root to install a package severely limits its reuse and distribution.
+Other people without administrative access to their machines are also cut off from using such software.
+Mostly, we've been bitten by this often enough to feel strongly that it should be its own rule.
 
@@
 > Good advice, but could be combined with 7, as they are both about configuration.
-
-................
-
 > Include a small test set
@@
 
-..................
-(Greg, any other changes you want to make here?)
-
 > What we left out
@@
 
-This reference has been removed.
+With the other changes, this section has been removed.
 
@@
 
-..............
+We added Rule #1, which is version control.
 
@@
 
-................
+We've moved these recommendations around so that they fit better into their major rule than they did before.
 
@@
 
-..............
+We've clarified this section to focus mostly on build automation and dependency acquisition.
 
@@
 
-We removed this last 'native functions' rule and replaced it with something more appropriate.
+We removed this last 'native functions' rule and replaced it with a more appropriate discussion of robust dependencies.
 
@@
 
-....
+This section has been re-written to take your concerns into account.
+
+---
+
+Thank you again for your time and effort.

### 2017-01-20 robust-software.tex

@@
 
-\newenvironment{reviewed}
-{ \color{orange} }
-
 % Remove comment for double spacing
@@
 \newcommand{\ruleminor}[1]{\textbf{#1}}
-
+\newcommand{\reviewed}[1]{{\color{black}#1}}
 \begin{document}
@@
 The root cause of this problem is that most research software
-is essentially a prototype, and therefore is not
+\reviewed{ is essentially a prototype, and therefore is not
 \emph{robust}. The lack of
 robustness in published, distributed software
-leads to duplicated efforts with little practical benefit,
+leads to duplicated efforts with little practical benefit,}
 which slows the pace of research~\cite{prabhu2011,lawlor2015}.
-Not all software needs to be robust enough to run for other people,
-but software that is published and distributed to the community should be.
 Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
@@
 engineering approaches, have high standards of reproducibility, use good testing
-practices, and foster strong user
-bases through constantly evolving, clearly documented, useful, and useable software.
+practices, and foster 
+\reviewed{strong user
+bases through constantly evolving, clearly documented, useful, and useable software.}
 In the bioinformatics community, Bioconductor and Galaxy follow this
 path~\cite{gentleman2004,afgan2016}.
-Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
+\reviewed{Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
 publish a paper about your software, it should, at minimum, satisfy these
-rules.
+rules.}
 
 So what \emph{is} ``robust'' software?
-We implied above that it is software that works for people other than the original author,
+\reviewed{We implied above that it is software that works for people other than the original author,
 on machines other than its creator's.
@@
 if your tools and libraries cannot be run by others,
-they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.
+they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.}
 
-\rulemajor{Rule 1: Use version control.}
+\reviewed{\rulemajor{Rule 1: Use version control.}
 
@@
 once the work has been completed and tested,
-it can be merged into the master branch for release.
+it can be merged into the master branch for release.}
 
-\rulemajor{Rule 2: Document your code and usage}
+\reviewed{\rulemajor{Rule 2: Document your code and usage}
 
@@
 
-\item \textit{List a few example commands} to get a user started quickly.
+\reviewed{\item \textit{List a few example commands} to get a user started quickly.}
 
@@
 is worth noting that software that is widely used and contributed to has and
-enforces the need for good documentation~\cite{gentleman2004}.
+enforces the need for good documentation~\cite{gentleman2004}.}
 
@@
 With a few lines, a call
-is made out to another library or program and the results are incorporated into the
+is made out to \reviewed{another library or program} and the results are incorporated into the
 primary script. Using popular projects reduces the amount of code that
@@
 
-Unfortunately, reusing software (whether software libraries or separate executables)
+Unfortunately, reusing software \reviewed{(whether software libraries or separate executables)}
 introduces dependencies, which can
@@
 often, support requests descend into debugging errors produced by the
-other project due to incompatible libraries, versions, or operating
+other project 
+\reviewed{due to incompatible libraries, versions, or operating
 systems~\cite{brown2013}. Even introducing libraries in the same programming
@@
 becomes much more difficult when relying on executables, or even on web
-services.
+services.}
 
@@
 executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration. Reuse software that offers some
-measurable improvement to your project.
+it may not be worth the pain of integration. \reviewed{Reuse software that offers some
+measurable improvement to your project.}
 
-Second, if launching an executable, \ruleminor{ensure the appropriate software and version is available}.
+Second, \reviewed{if launching an executable,}
+\ruleminor{ensure the appropriate software and version is available}.
 Either allow the user to
-configure the exact path to the package, distribute the program with the
+configure the exact path to the package, \reviewed{distribute the program with the
 dependent software, or download it during installation using your
@@
 separate executables, as they tend to be especially sensitive to operating
-systems, environments, and locales.
+systems, environments, and locales.}
 
@@
 
-To compile code, deploy applications, and automate other tasks,
+\reviewed{To compile code, deploy applications, and automate other tasks,
 programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build.
@@
 As mentioned in Rule~5,
-a package manager can mitigate some of the difficulties in software reuse.
+a package manager can mitigate some of the difficulties in software reuse.}
 
@@
 
-Every package should come with a small test script for users to run
+Every package should come with a set of tests for users to run
 after installation. Its purpose is not only to check that the software
@@
 
-In order to be useful, \ruleminor{make the test script easy to find and run}.
+In order to be useful, \ruleminor{make the tests easy to find and run}.
 Many build systems will also run unit tests if provided them at compile time.
-If the build system is not amenable to testing, a
-single file in the project's root directory named \texttt{runtests.sh}
-or something equally obvious is a 
-good solution that lets new users build their analysis from a working script.
+\reviewed{For users, or
+if the build system is not amenable to testing, 
+provide a working script
+in the project's root directory named \texttt{runtests.sh}
+or something equally obvious. 
+This lets new users build their analysis from a working script.
 For example, with its distribution, 
@@
 and a 'Getting Started with HISAT2' section in its manual 
-that leads you through the entire data lifecycle~\cite{pertea2016}.
+that leads you through the entire data lifecycle~\cite{pertea2016}.}
 
@@
 
-The usage message tells users what the program could do.
+\reviewed{The usage message tells users what the program could do.
 It is equally important for the program to tell users what it actually did.
@@
 feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
+replicated with only the previous output files as reference.}
 

### 2017-02-13 robust-software.tex
Tidying up the formatting a bit

--- 2017-01-19 robust-software.tex
@@
 \newcommand{\ruleminor}[1]{\textbf{#1}}
-
 \begin{document}
@@
 
-\linenumbers
+% \linenumbers
 
@@
 which slows the pace of research~\cite{prabhu2011,lawlor2015}.
-Not all software needs to be robust enough to run for other people,
-but software that is published and distributed to the community should be.
 Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
@@
 engineering approaches, have high standards of reproducibility, use good testing
-practices, and foster strong user
+practices, and foster 
+strong user
 bases through constantly evolving, clearly documented, useful, and useable software.
@@
 and list the required fields and acceptable values in other files.
-If there is no rigorous format, explain what the sections mean.
-These files are often full of valuable information that can be
-mined for the user's specific purpose.
+If there is no rigorous definition for a format, explain its parts
+as clearly as possible in plain English.
 
@@
 often, support requests descend into debugging errors produced by the
-other project due to incompatible libraries, versions, or operating
+other project 
+due to incompatible libraries, versions, or operating
 systems~\cite{brown2013}. Even introducing libraries in the same programming
@@
 
-Second, if launching an executable, \ruleminor{ensure the appropriate software and version is available}.
+Second, if launching an executable,
+\ruleminor{ensure the appropriate software and version is available}.
 Either allow the user to
@@
 specify an input or output \emph{directory}, and then require that there
-be files with particular names in that directory. This practice, which
-is sometimes called ``convention over configuration'', is used by many
-software frameworks, such as WordPress and Ruby on Rails, and often
+be files with particular names in that directory. This practice is an
+example of ``convention over configuration'', a principle is used by
+software frameworks such as WordPress and Ruby on Rails that often
 strikes a good balance between adaptability and consistency.
@@
 
-In order to be useful, \ruleminor{make the test script easy to find and run}.
+In order to be useful, \ruleminor{make the tests easy to find and run}.
 Many build systems will also run unit tests if provided them at compile time.
-If the build system is not amenable to testing, a
-single file in the project's root directory named \texttt{runtests.sh}
-or something equally obvious is a 
-good solution that lets new users build their analysis from a working script.
+For users, or
+if the build system is not amenable to testing, 
+provide a working script
+in the project's root directory named \texttt{runtests.sh}
+or something equally obvious. 
+This lets new users build their analysis from a working script.
 For example, with its distribution, 
@@
 when the program starts, it should \ruleminor{echo all parameters and software
-versions to standard out or a log file alongside the results}. This
-feature supports greater reproducibility because any result can be
-replicated with only the previous output files as reference.
+versions to standard out or a log file alongside the results} to increase
+the reproducibility of that step.
 
@@
 
-\bibliography{robust-software}
+\begin{thebibliography}{10}
+
+\bibitem{prins2015}
+Prins P, de~Ligt J, Tarasov A, Jansen RC, Cuppen E, Bourne PE.
+\newblock Toward effective software solutions for big biology.
+\newblock Nature Biotechnology. 2015;33(7):686--687.
+\newblock doi:{10.1038/nbt.3240}.
+
+\bibitem{baker2016}
+Baker M.
+\newblock 1,500 scientists lift the lid on reproducibility.
+\newblock Nature. 2016;533(7604):452--454.
+\newblock doi:{10.1038/533452a}.
+
+\bibitem{prabhu2011}
+Prabhu P, Jablin TB, Raman A, Zhang Y, Huang J, Kim H, et~al.
+\newblock A Survey of the Practice of Computational Science.
+\newblock In: State of the Practice Reports. SC '11. New York, NY, USA: ACM;
+  2011. p. 19:1--19:12.
+
+\bibitem{lawlor2015}
+Lawlor B, Walsh P.
+\newblock Engineering bioinformatics: building reliability, performance and
+  productivity into bioinformatics software.
+\newblock Bioengineered. 2015;6(4).
+\newblock doi:{10.1080/21655979.2015.1050162}.
+
+\bibitem{ison2016}
+Ison J, Rapacki K, Ménager H, Kalaš M, Rydza E, Chmura P, et~al.
+\newblock Tools and data services registry: a community effort to document
+  bioinformatics resources.
+\newblock Nucleic Acids Research. 2016;44(D1):D38--D47.
+\newblock doi:{10.1093/nar/gkv1116}.
+
+\bibitem{brazas2012}
+Brazas MD, Yim D, Yeung W, Ouellette BFF.
+\newblock A decade of web server updates at the bioinformatics links directory:
+  2003-–2012.
+\newblock Nucleic Acids Research. 2012;doi:{10.1093/nar/gks632}.
+
+\bibitem{stajich2002}
+Stajich JE, Block D, Boulez K, Brenner SE, Chervitz SA, Dagdigian C, et~al.
+\newblock The Bioperl Toolkit: Perl Modules for the Life Sciences.
+\newblock Genome Research. 2002;12(10):1611--1618.
+\newblock doi:{10.1101/gr.361602}.
+
+\bibitem{Seemann2013}
+Seemann T.
+\newblock Ten recommendations for creating usable bioinformatics command line
+  software.
+\newblock GigaScience. 2013;2(1):15.
+\newblock doi:{10.1186/2047-217X-2-15}.
+
+\bibitem{nekrutenko2012}
+Nekrutenko A, Taylor J.
+\newblock Next-generation sequencing data interpretation: enhancing
+  reproducibility and accessibility.
+\newblock Nature ReviewsGenetics. 2012;13(9):667--72.
+\newblock doi:{10.1038/nrg3305}.
+
+\bibitem{atwood2015}
+Atwood TK, Bongcam-Rudloff E, Brazas ME, Corpas M, Gaudet P, Lewitter F, et~al.
+\newblock GOBLET: The Global Organisation for Bioinformatics Learning,
+  Education and Training.
+\newblock PLoS Computational Biology. 2015;doi:{10.1371/journal.pcbi.1004143}.
+
+\bibitem{sanders2008}
+Sanders R, Kelly D.
+\newblock Dealing with Risk in Scientific Software Development.
+\newblock Software, IEEE. 2008;25(4):21--28.
+\newblock doi:{10.1109/ms.2008.84}.
+
+\bibitem{afgan2016}
+Afgan E, Baker D, van~den Beek M, Blankenberg D, Bouvier D, Čech M, et~al.
+\newblock The Galaxy platform for accessible, reproducible and collaborative
+  biomedical analyses: 2016 update.
+\newblock Nucleic Acids Research. 2016;44(W1):W3--W10.
+\newblock doi:{10.1093/nar/gkw343}.
+
+\bibitem{howe2012}
+Howe B.
+\newblock Virtual Appliances, Cloud Computing, and Reproducible Research.
+\newblock Computing in Science Engineering. 2012;14(4):36--41.
+\newblock doi:{10.1109/MCSE.2012.62}.
+
+\bibitem{gentleman2004}
+Gentleman RC, Carey VJ, Bates DM, Bolstad B, Dettling M, Dudoit S, et~al.
+\newblock Bioconductor: open software development for computational biology and
+  bioinformatics.
+\newblock Genome Biology. 2004;5(10):1--16.
+\newblock doi:{10.1186/gb-2004-5-10-r80}.
+
+\bibitem{varoquaux2015}
+Varoquaux G. Software for reproducible science: let’s not have a
+  misunderstanding; 2015.
+\newblock
+  \url{http://gael-varoquaux.info/programming/software-for-reproducible-science-lets-not-have-a-misunderstanding.html}.
+
+\bibitem{brown2013}
+Brown CT. Replication, reproduction, and remixing in research software; 2013.
+\newblock \url{http://ivory.idyll.org/blog/research-software-reuse.html}.
+
+\bibitem{wilson2014}
+Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et~al.
+\newblock Best Practices for Scientific Computing.
+\newblock PLoS Biology. 2014;12(1):e1001745.
+\newblock doi:{10.1371/journal.pbio.1001745}.
+
+\bibitem{wilson2016}
+Wilson G, Bryan J, Cranston K, Kitzes J, Nederbragt L, Teal TK.
+\newblock Good Enough Practices in Scientific Computing.
+\newblock arxivorg. 2016;abs/1609.00037.
+
+\bibitem{noble2009}
+Noble WS.
+\newblock {A Quick Guide to Organizing Computational Biology Projects}.
+\newblock PLoS Computational Biology. 2009;5(7).
+\newblock doi:{10.1371/journal.pcbi.1000424}.
+
+\bibitem{blischak2016}
+Blischak JD, Davenport ER, Wilson G.
+\newblock A Quick Introduction to Version Control with Git and GitHub.
+\newblock PLOS Computational Biology. 2016;12(1):1--18.
+\newblock doi:{10.1371/journal.pcbi.1004668}.
+
+\bibitem{karimzadeh2016}
+Karimzadeh M, Hoffman MM. Creating great documentation for bioinformatics
+  software; 2016.
+\newblock \url{http://hdl.handle.net/1807/73111}.
+
+\bibitem{Johnson1997}
+Johnson M.
+\newblock Building a Better ReadMe.
+\newblock Technical Communication. 1997;44(1):28--36.
+
+\bibitem{gnustandards}
+{\relax Free Software Foundation}. GNU Coding Standards; 2016.
+\newblock \url{https://www.gnu.org/prep/standards/standards.html}.
+
+\bibitem{posix2016}
+{\relax The IEEE}, {\relax The Open Group}. The Open Group Base Specifications
+  Issue 7 IEEE Std 1003.1-2008. 12. Utility Conventions; 2016.
+\newblock
+  \url{http://pubs.opengroup.org/onlinepubs/9699919799/basedefs/V1_chap12.html}.
+
+\bibitem{pertea2016}
+Pertea M, Kim D, Pertea GM, Leek JT, Salzberg SL.
+\newblock Transcript-level expression analysis of RNA-seq experiments with
+  HISAT, StringTie and Ballgown.
+\newblock Nature Protocols. 2016;11(9):1650--1667.
+
+\bibitem{steinmacher2015}
+Steinmacher I, Silva MAG, Gerosa MA, Redmiles DF.
+\newblock A systematic literature review on the barriers faced by newcomers to
+  open source software projects.
+\newblock Information and Software Technology. 2015;59:67 -- 85.
+\newblock doi:{http://dx.doi.org/10.1016/j.infsof.2014.11.001}.
+
+\end{thebibliography}
 

### 2017-02-13 robust-software.tex
Tidying up layout of source

@@
 
-Software produced for research,
-published and otherwise,
-suffers from a number of common problems
-that make it difficult or impossible to run outside the original institution,
-or even off the primary developer's computer.
-We present ten simple rules to make such software robust enough to be run by anyone,
-anywhere,
-and thereby delight your users and collaborators.
+Software produced for research, published and otherwise, suffers from
+a number of common problems that make it difficult or impossible to
+run outside the original institution, or even off the primary
+developer's computer.  We present ten simple rules to make such
+software robust enough to be run by anyone, anywhere, and thereby
+delight your users and collaborators.
 
@@
 
-Many researchers have found out the hard way that there's a world of difference
-between ``works for me on my machine'' and ``works for other people on theirs''.
-Many common challenges can be avoided by following a few simple rules; doing so
-not only improves reproducibility, but can accelerate research.
-
-% \linenumbers
+Many researchers have found out the hard way that there's a world of
+difference between ``works for me on my machine'' and ``works for
+other people on theirs''.  Many common challenges can be avoided by
+following a few simple rules; doing so not only improves
+reproducibility, but can accelerate research.
 
@@
 
-Scientific software is typically developed and used by a single person,
-usually a graduate student or postdoc~\cite{prins2015}.
-It may produce the intended results in their hands,
-but what happens when someone else wants to run it? Everyone with
-a few years of experience feels a bit nervous when told to use
-another person's code to analyze their data:
-it will often be undocumented,
-work in unexpected ways (if it works at all),
-rely on nonexistent paths or resources,
-be tuned for a single dataset,
-or simply be an older version than was used in published papers.
-The potential new user is then faced with two unpalatable options:
-hack the existing code to make it work, or start over.
+Scientific software is typically developed and used by a single
+person, usually a graduate student or postdoc~\cite{prins2015}.  It
+may produce the intended results in their hands, but what happens when
+someone else wants to run it? Everyone with a few years of experience
+feels a bit nervous when told to use another person's code to analyze
+their data: it will often be undocumented, work in unexpected ways (if
+it works at all), rely on nonexistent paths or resources, be tuned for
+a single dataset, or simply be an older version than was used in
+published papers.  The potential new user is then faced with two
+unpalatable options: hack the existing code to make it work, or start
+over.
 
-Being unable to
-replicate results is so common that one publication refers to it as ``a rite of
-passage''~\cite{baker2016}.
-The root cause of this problem is that most research software
-is essentially a prototype, and therefore is not
-\emph{robust}. The lack of
-robustness in published, distributed software
-leads to duplicated efforts with little practical benefit,
-which slows the pace of research~\cite{prabhu2011,lawlor2015}.
-Bioinformatics software repositories \cite{ison2016,brazas2012} catalogue dozens to
-hundreds of tools that perform similar tasks:
-for example,
-in 2016 the Bioinformatics Links Directory included 84 different multiple sequence aligners, 141 tools
-to analyze transcript expression, and 182 pathway and interaction resources.
-Some of these tools are legitimate efforts to improve the state-of-the-art, but
-often they are difficult to install and run~\cite{stajich2002,Seemann2013}, and are effectively abandoned
-after publication~\cite{nekrutenko2012}.
+Being unable to replicate results is so common that one publication
+refers to it as ``a rite of passage''~\cite{baker2016}.  The root
+cause of this problem is that most research software is essentially a
+prototype, and therefore is not \emph{robust}. The lack of robustness
+in published, distributed software leads to duplicated efforts with
+little practical benefit, which slows the pace of
+research~\cite{prabhu2011,lawlor2015}.  Bioinformatics software
+repositories \cite{ison2016,brazas2012} catalogue dozens to hundreds
+of tools that perform similar tasks: for example, in 2016 the
+Bioinformatics Links Directory included 84 different multiple sequence
+aligners, 141 tools to analyze transcript expression, and 182 pathway
+and interaction resources.  Some of these tools are legitimate efforts
+to improve the state-of-the-art, but often they are difficult to
+install and run~\cite{stajich2002,Seemann2013}, and are effectively
+abandoned after publication~\cite{nekrutenko2012}.
 
-This problem is not unique to bioinformatics, or even to computing~\cite{baker2016}.
-Best practices in software engineering specifically aim to increase software
-robustness. However, most bioinformaticians learn what they know about software development
-on the job or otherwise informally~\cite{prins2015,atwood2015}.
-Existing training programs and initiatives rarely have the time to cover software engineering
-in depth, especially since the field is so broad and developing so rapidly~\cite{atwood2015,lawlor2015}.
-In addition, making software robust is not directly rewarded
-in science, and funding is difficult to come by~\cite{prins2015}. Some proposed
+This problem is not unique to bioinformatics, or even to
+computing~\cite{baker2016}.  Best practices in software engineering
+specifically aim to increase software robustness. However, most
+bioinformaticians learn what they know about software development on
+the job or otherwise informally~\cite{prins2015,atwood2015}.  Existing
+training programs and initiatives rarely have the time to cover
+software engineering in depth, especially since the field is so broad
+and developing so rapidly~\cite{atwood2015,lawlor2015}.  In addition,
+making software robust is not directly rewarded in science, and
+funding is difficult to come by~\cite{prins2015}. Some proposed
 solutions to this problem include restructuring educational programs,
 hiring dedicated software engineers~\cite{lawlor2015,sanders2008},
-partnering with private sector or grassroots organizations~\cite{prins2015,ison2016},
-or using specific technical tools like containerization or cloud
-computing~\cite{afgan2016,howe2012}. Each of these requires time and, in some
-cases, institutional change.
+partnering with private sector or grassroots
+organizations~\cite{prins2015,ison2016}, or using specific technical
+tools like containerization or cloud
+computing~\cite{afgan2016,howe2012}. Each of these requires time and,
+in some cases, institutional change.
 
-The good news is,
-you don't need to be a professionally-trained programmer to write robust software.
-In fact,
-some of the best, most reliable pieces of software in many scientific
-communities are written by researchers~\cite{prabhu2011,sanders2008}
-who have adopted strong software
-engineering approaches, have high standards of reproducibility, use good testing
-practices, and foster 
-strong user
-bases through constantly evolving, clearly documented, useful, and useable software.
-In the bioinformatics community, Bioconductor and Galaxy follow this
-path~\cite{gentleman2004,afgan2016}.
-Not all scientific software needs to be robust~\cite{varoquaux2015}, but if you
-publish a paper about your software, it should, at minimum, satisfy these
-rules.
+The good news is, you don't need to be a professionally-trained
+programmer to write robust software.  In fact, some of the best, most
+reliable pieces of software in many scientific communities are written
+by researchers~\cite{prabhu2011,sanders2008} who have adopted strong
+software engineering approaches, have high standards of
+reproducibility, use good testing practices, and foster strong user
+bases through constantly evolving, clearly documented, useful, and
+useable software.  In the bioinformatics community, Bioconductor and
+Galaxy follow this path~\cite{gentleman2004,afgan2016}.  Not all
+scientific software needs to be robust~\cite{varoquaux2015}, but if
+you publish a paper about your software, it should, at minimum,
+satisfy these rules.
 
-So what \emph{is} ``robust'' software?
-We implied above that it is software that works for people other than the original author,
-on machines other than its creator's.
-More specifically, we mean that:
+So what \emph{is} ``robust'' software?  We implied above that it is
+software that works for people other than the original author, on
+machines other than its creator's.  More specifically, we mean that:
 
 \begin{itemize}
+
 \item
   it can be installed on more than one computer with relative ease,
+
 \item
   it works consistently as advertised, and
+
 \item
   it can be integrated with other tools.
+
 \end{itemize}
 
-Our rules are generic and can be applied to all languages, libraries, packages,
-documentation styles, and operating systems for both closed-source and open-source software. 
-They are also necessary steps toward making computational research replicable and reproducible:
-after all,
-if your tools and libraries cannot be run by others,
-they cannot be used to verify your results or as a stepping stone for future work~\cite{brown2013}.
+Our rules are generic and can be applied to all languages, libraries,
+packages, documentation styles, and operating systems for both
+closed-source and open-source software.  They are also necessary steps
+toward making computational research replicable and reproducible:
+after all, if your tools and libraries cannot be run by others, they
+cannot be used to verify your results or as a stepping stone for
+future work~\cite{brown2013}.
 
@@
 Version control is essential to sustainable software development
-\cite{wilson2014,wilson2016}.
-In particular,
-developers will struggle to understand what they have actually built,
-what it actually does,
-and what they have actually released
-without some mechanical way to keep track of changes.
-They should therefore
-\ruleminor{put everything created manually into version control as soon as it is created},
-including programs, original field observations, and the source files for papers.
-Files that can be regenerated at need,
-such as the binaries for compiled programs or intermediate files generated during data analysis,
-should not be versioned;
-instead,
-it is often more sensible to use an archiving system for them,
-and store the metadata describing their contents in version control instead~\cite{noble2009}.
+\cite{wilson2014,wilson2016}.  In particular, developers will struggle
+to understand what they have actually built, what it actually does,
+and what they have actually released without some mechanical way to
+keep track of changes.  They should therefore \ruleminor{put
+everything created manually into version control as soon as it is
+created}, including programs, original field observations, and the
+source files for papers.  Files that can be regenerated at need, such
+as the binaries for compiled programs or intermediate files generated
+during data analysis, should not be versioned; instead, it is often
+more sensible to use an archiving system for them, and store the
+metadata describing their contents in version control
+instead~\cite{noble2009}.
 
-If you are new to version control,
-it is simplest to treat it as ``a better \withurl{Dropbox}{http://dropbox.com}''
-(or, if you are of a certain age, a better FTP),
-and to use it simply to synchronize files between multiple developers and machines~\cite{blischak2016}.
-Once you are comfortable working that way,
-you should \ruleminor{use a feature branch workflow}:
-designate one parallel copy (or ``branch'') of the repository as the master,
-and create a new branch from it each time you want to fix a bug or add a new feature.
-This allows work on independent changes to proceed in isolation;
-once the work has been completed and tested,
-it can be merged into the master branch for release.
+If you are new to version control, it is simplest to treat it as ``a
+better \withurl{Dropbox}{http://dropbox.com}'' (or, if you are of a
+certain age, a better FTP), and to use it simply to synchronize files
+between multiple developers and machines~\cite{blischak2016}.  Once
+you are comfortable working that way, you should \ruleminor{use a
+feature branch workflow}: designate one parallel copy (or
+``branch'') of the repository as the master, and create a new branch
+from it each time you want to fix a bug or add a new feature.  This
+allows work on independent changes to proceed in isolation; once the
+work has been completed and tested, it can be merged into the master
+branch for release.
 
@@
 
-How to write high quality documentation has been described elsewhere~\cite{karimzadeh2016}
-and so here we only cover two minimal types: the README and usage.
-The README is usually available even before the software is installed,
-exists
-to get a new user started, and points them towards more
-help. Usage is a terse, informative command-line help message that
-guides the user in the correct use of the software.
+How to write high quality documentation has been described
+elsewhere~\cite{karimzadeh2016} and so here we only cover two minimal
+types: the README and usage.  The README is usually available even
+before the software is installed, exists to get a new user started,
+and points them towards more help. Usage is a terse, informative
+command-line help message that guides the user in the correct use of
+the software.
 
-Numerous guidelines exist on how to \ruleminor{write a good
-README file}~\cite{Johnson1997,gnustandards}.
-At a minimum, your README should:
+Numerous guidelines exist on how to \ruleminor{write a good README
+file}~\cite{Johnson1997,gnustandards}.  At a minimum, your README
+should:
 
@@
 
-\item \textit{Explain what the software does}.
-There's nothing more frustrating
-than downloading and installing something only to
-find out that it doesn't do what you thought it did.
+\item
+  \textit{Explain what the software does}.  There's nothing more
+  frustrating than downloading and installing something only to find
+  out that it doesn't do what you thought it did.
 
-\item \textit{List required dependencies}.
-We address dependencies in more detail in Rule~5.
+\item
+  \textit{List required dependencies}.  We address dependencies in
+  more detail in Rule~5.
 
-\item \textit{Provide compilation or installation instructions}.
+\item
+  \textit{Provide compilation or installation instructions}.
 
-\item \textit{List all input and output files},
-even those considered self-explanatory.
-Link to specifications for standard formats
-and list the required fields and acceptable values in other files.
-If there is no rigorous definition for a format, explain its parts
-as clearly as possible in plain English.
+\item
+  \textit{List all input and output files}, even those considered
+  self-explanatory.  Link to specifications for standard formats and
+  list the required fields and acceptable values in other files.  If
+  there is no rigorous definition for a format, explain its parts as
+  clearly as possible in plain English.
 
-\item \textit{List a few example commands} to get a user started quickly.
+\item
+  \textit{List a few example commands} to get a user started quickly.
 
-\item \textit{State attributions and licensing}. Attributions are how you credit
-your contributors; licenses dictate how others may use and
+\item
+  \textit{State attributions and licensing}. Attributions are how you
+  credit your contributors; licenses dictate how others may use and
   need to credit your work.
+
 \end{enumerate}
 
-The program should also
-\ruleminor{print usage information} when launching from the command line.
-Usage provides the first line of help for both new and experienced users.
-Terseness is important: usage that extends for multiple screens is
-difficult to read or refer to on the fly.
+The program should also \ruleminor{print usage information} when
+launching from the command line.  Usage provides the first line of
+help for both new and experienced users.  Terseness is important:
+usage that extends for multiple screens is difficult to read or refer
+to on the fly.
 
 Almost all command-line applications use a combination of
-POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for usage.
-More standard command-line behaviours are detailed in \cite{Seemann2013}.
-Your software's usage should:
+POSIX~\cite{posix2016} and GNU~\cite{gnustandards} standards for
+usage.  More standard command-line behaviours are detailed in
+\cite{Seemann2013}.  Your software's usage should:
 
@@
 
-\item \textit{Describe the syntax for running the program}, including the
-  name of the program, the relative location of optional
-  and required flags, other arguments, and values for execution.
+\item
+  \textit{Describe the syntax for running the program}, including the
+  name of the program, the relative location of optional and required
+  flags, other arguments, and values for execution.
 
-\item \textit{Give a short description} to remind users of the software's primary function.
+\item
+  \textit{Give a short description} to remind users of the software's
+  primary function.
 
-\item \textit{List the most commonly used arguments}, a description of each, and
-    the default values.
+\item
+  \textit{List the most commonly used arguments}, a description of
+  each, and the default values.
 
-\item \textit{State where to find more information}.
+\item
+  \textit{State where to find more information}.
 
@@
 
-Usage should be
-printed to standard output so that it can be
-combined with other bash utilities like \texttt{grep},
-and it should finish with an appropiate exit code.
+Usage should be printed to standard output so that it can be combined
+with other bash utilities like \texttt{grep}, and it should finish
+with an appropiate exit code.
 
-Documentation beyond the README and usage is up to the developer's discretion.
-We think it is very important for developers to document their work, but our
-experience is that people are unlikely do it during normal development
-However, it
-is worth noting that software that is widely used and contributed to has and
-enforces the need for good documentation~\cite{gentleman2004}.
+Documentation beyond the README and usage is up to the developer's
+discretion.  We think it is very important for developers to document
+their work, but our experience is that people are unlikely do it
+during normal development However, it is worth noting that software
+that is widely used and contributed to has and enforces the need for
+good documentation~\cite{gentleman2004}.
 
@@
 Being able to change parameters on the fly to determine if and how
-they change the results is important as your software gains more users,
-as it facilitates exploratory analysis and parameter sweeping.
-Programs should therefore
-\ruleminor{allow the most commonly changed parameters to be configured from the command line}.
+they change the results is important as your software gains more
+users, as it facilitates exploratory analysis and parameter sweeping.
+Programs should therefore \ruleminor{allow the most commonly changed
+parameters to be configured from the command line}.
 
-Users will want to change some values more often than others.
-Since parameters are software-specific, the appropriate 'tunable' ones cannot be detailed here,
-but a short list includes input and reference files and directories,
-output files and directories,
-filtering parameters,
-random number generation seeds,
-and
-alternatives such as compressing results,
-use a variant algorithm,
-or verbose output.
+Users will want to change some values more often than others.  Since
+parameters are software-specific, the appropriate 'tunable' ones
+cannot be detailed here, but a short list includes input and reference
+files and directories, output files and directories, filtering
+parameters, random number generation seeds, and alternatives such as
+compressing results, use a variant algorithm, or verbose output.
 
-\ruleminor{Check that all input values are in a reasonable range at startup}.
-Few things are as annoying as having a program announce after running for two
-hours that it isn't going to save its results because the requested directory
-doesn't exist.
+\ruleminor{Check that all input values are in a reasonable range at
+startup}.  Few things are as annoying as having a program announce
+after running for two hours that it isn't going to save its results
+because the requested directory doesn't exist.
 
-To make programs even easier to use,
-\ruleminor{choose reasonable defaults where they exist}
-and \ruleminor{set no defaults at all when there aren't any reasonable ones}.
-You can set reasonable default values
-as long as any command line arguments
-override those values.
+To make programs even easier to use, \ruleminor{choose reasonable
+defaults where they exist} and \ruleminor{set no defaults at all
+when there aren't any reasonable ones}.  You can set reasonable
+default values as long as any command line arguments override those
+values.
 
-Changeable values should \emph{never} be hard-coded:
-if users have to edit your software in order to run it,
-you have done something wrong.
-Changeable but infrequently-changed values should therefore be stored in configuration files.
-These can be in a standard location,
-e.g. \texttt{.packagerc} in the user's home directory,
-or provided on the command line as an additional argument.
-Configuration files are often created during installation
-to set up such things as server names,
-network drives,
-and other defaults for your lab or institution.
+Changeable values should \emph{never} be hard-coded: if users have to
+edit your software in order to run it, you have done something wrong.
+Changeable but infrequently-changed values should therefore be stored
+in configuration files.  These can be in a standard location,
+e.g. \texttt{.packagerc} in the user's home directory, or provided on
+the command line as an additional argument.  Configuration files are
+often created during installation to set up such things as server
+names, network drives, and other defaults for your lab or institution.
 
@@
 
-Software evolves over time, with developers adding or removing features as need
-dictates. Making official releases stamps a particular set of features with a
-project-specific identifier so that version can be retrieved for later use. For
-example, if a paper is published, the software should be released at the same
-time so that the results can be reproduced.
+Software evolves over time, with developers adding or removing
+features as need dictates. Making official releases stamps a
+particular set of features with a project-specific identifier so that
+version can be retrieved for later use. For example, if a paper is
+published, the software should be released at the same time so that
+the results can be reproduced.
 
 Most software has a version number composed of a decimal number that
-increments as new versions are released.
-There are many different ways
-to construct and interpret this number, but most importantly for us, a
-particular software version run with the same parameters should give
-identical results no matter when it's run. Results include both correct
-output as well as any errors.
-\ruleminor{Increment your version number every time you release your software to
-other people}.
+increments as new versions are released.  There are many different
+ways to construct and interpret this number, but most importantly for
+us, a particular software version run with the same parameters should
+give identical results no matter when it's run. Results include both
+correct output as well as any errors.  \ruleminor{Increment your
+version number every time you release your software to other
+people}.
 
-\withurl{Semantic versioning}{http://semver.org/} is one of the most common
-types of versioning for open-source software. Version numbers take the
-form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6.
-Changes in the major
-version number herald significant changes in the software that are not
-backwards compatible, such as changing or removing features or altering
-the primary functions of the software. Increasing the minor version
-represents incremental improvements in the software, like adding new
-features. Following the minor version number can be an arbitrary number
-of project-specific identifiers, including patches, builds and qualifiers.
-Common qualifiers include \texttt{alpha}, \texttt{beta}, and \texttt{SNAPSHOT},
-for applications that are
-not yet stable or released, and \texttt{-RC} for release candidates prior
+\withurl{Semantic versioning}{http://semver.org/} is one of the most
+common types of versioning for open-source software. Version numbers
+take the form of \emph{MAJOR.MINOR{[}.PATCH{]}}, e.g., 0.2.6.  Changes
+in the major version number herald significant changes in the software
+that are not backwards compatible, such as changing or removing
+features or altering the primary functions of the software. Increasing
+the minor version represents incremental improvements in the software,
+like adding new features. Following the minor version number can be an
+arbitrary number of project-specific identifiers, including patches,
+builds and qualifiers.  Common qualifiers include \texttt{alpha},
+\texttt{beta}, and \texttt{SNAPSHOT}, for applications that are not
+yet stable or released, and \texttt{-RC} for release candidates prior
 to an official release.
@@
 \ruleminor{The version of your software should be easily available by
-supplying \texttt{-\/-version} or \texttt{-v} on the command line}. This command should print
-the software name and version number, and it should
-also be \ruleminor{included in all of the program's output}, particularly debugging
-traces.  If someone needs help, it's important that they be able to tell
-whoever's helping them which version of the software they're using.
+supplying \texttt{-\/-version} or \texttt{-v} on the command
+line}. This command should print the software name and version
+number, and it should also be \ruleminor{included in all of the
+program's output}, particularly debugging traces.  If someone needs
+help, it's important that they be able to tell whoever's helping them
+which version of the software they're using.
 
-While new releases may make a program better in general,
-they can simultaneously create work for someone
-who integrated the old version into their own workflow a year or two ago,
-and won't see any benefits from upgrading.
-A program's authors should therefore \ruleminor{ensure that old released versions
-continue to be available.}
-A number of mechanisms exist for
-controlled release that range from as simple as adding an appropriate
-commit message or tag to version control~\cite{blischak2016}, to official releases alongside
-code on Bitbucket or GitHub, to depositing into a
-repository like apt, yum, homebrew, CPAN, etc. Choose the method that
-best suits the number and expertise of users you anticipate.
+While new releases may make a program better in general, they can
+simultaneously create work for someone who integrated the old version
+into their own workflow a year or two ago, and won't see any benefits
+from upgrading.  A program's authors should therefore
+\ruleminor{ensure that old released versions continue to be
+available.}  A number of mechanisms exist for controlled release
+that range from as simple as adding an appropriate commit message or
+tag to version control~\cite{blischak2016}, to official releases
+alongside code on Bitbucket or GitHub, to depositing into a repository
+like apt, yum, homebrew, CPAN, etc. Choose the method that best suits
+the number and expertise of users you anticipate.
 
@@
 
-In the spirit of code reuse and interoperability, developers often want
-to reuse software written by others.
-With a few lines, a call
-is made out to another library or program and the results are incorporated into the
-primary script. Using popular projects reduces the amount of code that
-needs to be maintained and leverages the work done by the other software.
+In the spirit of code reuse and interoperability, developers often
+want to reuse software written by others.  With a few lines, a call is
+made out to another library or program and the results are
+incorporated into the primary script. Using popular projects reduces
+the amount of code that needs to be maintained and leverages the work
+done by the other software.
 
-Unfortunately, reusing software (whether software libraries or separate executables)
-introduces dependencies, which can
-bring their own special pain. The interface between two software
-packages can be a source of considerable frustration: all too
-often, support requests descend into debugging errors produced by the
-other project 
-due to incompatible libraries, versions, or operating
-systems~\cite{brown2013}. Even introducing libraries in the same programming
-language can rely on software installed in the environment, and the problem
-becomes much more difficult when relying on executables, or even on web
-services.
+Unfortunately, reusing software (whether software libraries or
+separate executables) introduces dependencies, which can bring their
+own special pain. The interface between two software packages can be a
+source of considerable frustration: all too often, support requests
+descend into debugging errors produced by the other project due to
+incompatible libraries, versions, or operating
+systems~\cite{brown2013}. Even introducing libraries in the same
+programming language can rely on software installed in the
+environment, and the problem becomes much more difficult when relying
+on executables, or even on web services.
 
-Despite these problems, software developers in research should
-re-use existing software provided a few guidelines are adhered to.
+Despite these problems, software developers in research should re-use
+existing software provided a few guidelines are adhered to.
 
-First,
-\ruleminor{make sure that you really need the auxiliary program}. If you are
-executing GNU sort instead of figuring out how to sort lists in Python,
-it may not be worth the pain of integration. Reuse software that offers some
-measurable improvement to your project.
+First, \ruleminor{make sure that you really need the auxiliary
+program}. If you are executing GNU sort instead of figuring out how
+to sort lists in Python, it may not be worth the pain of
+integration. Reuse software that offers some measurable improvement to
+your project.
 
-Second, if launching an executable,
-\ruleminor{ensure the appropriate software and version is available}.
-Either allow the user to
-configure the exact path to the package, distribute the program with the
-dependent software, or download it during installation using your
-package manager. If the executable requires internet access, check for that
-early in execution.
+Second, if launching an executable, \ruleminor{ensure the appropriate
+software and version is available}.  Either allow the user to
+configure the exact path to the package, distribute the program with
+the dependent software, or download it during installation using your
+package manager. If the executable requires internet access, check for
+that early in execution.
 
-Third, \ruleminor{ensure that reused software is robust}. Relying on erratic third
-party libraries or software is a recipe for tears. Prefer software that follows
-good software development practices, is open for support questions, and is
-available from a stable location or repository using your package manager.
+Third, \ruleminor{ensure that reused software is robust}. Relying on
+erratic third party libraries or software is a recipe for
+tears. Prefer software that follows good software development
+practices, is open for support questions, and is available from a
+stable location or repository using your package manager.
 
-Exercise caution especially when transitioning across languages or using
-separate executables, as they tend to be especially sensitive to operating
-systems, environments, and locales.
+Exercise caution especially when transitioning across languages or
+using separate executables, as they tend to be especially sensitive to
+operating systems, environments, and locales.
 
@@
 To compile code, deploy applications, and automate other tasks,
-programmers routinely use build tools like Make, Rake, Maven, Ant or MS Build.
-These tools can also be used to manage runtime environments,
-i.e.,
-to check that the right versions of required packages are installed
-and install or upgrade them if they are not.
-As mentioned in Rule~5,
-a package manager can mitigate some of the difficulties in software reuse.
+programmers routinely use build tools like Make, Rake, Maven, Ant or
+MS Build.  These tools can also be used to manage runtime
+environments, i.e., to check that the right versions of required
+packages are installed and install or upgrade them if they are not.
+As mentioned in Rule~5, a package manager can mitigate some of the
+difficulties in software reuse.
 
-The same tools can and should be used to manage runtime environments on users' machines as well.
-Accordingly,
-developers should
+The same tools can and should be used to manage runtime environments
+on users' machines as well.  Accordingly, developers should
 \ruleminor{document all dependencies in a machine-readable form}.
-Package managers like apt and yum are available on most Unix-like systems, and
-application package managers exist for specific languages like Python (pip),
-Java (Maven/Gradle), and Ruby (RubyGems). These package managers can be used
-together with the build utility to ensure that dependencies are available at
-compile/run time.
+Package managers like apt and yum are available on most Unix-like
+systems, and application package managers exist for specific languages
+like Python (pip), Java (Maven/Gradle), and Ruby (RubyGems). These
+package managers can be used together with the build utility to ensure
+that dependencies are available at compile/run time.
 
@@
 
-This file can be read by the pip package manager, which can check that the
-required software is available and install it if it is not.
-Whatever is used,
-developers should \emph{always} install dependencies
-using their dependency description, especially on their personal machines, so that
-they're sure it works.
+This file can be read by the pip package manager, which can check that
+the required software is available and install it if it is not.
+Whatever is used, developers should \emph{always} install dependencies
+using their dependency description, especially on their personal
+machines, so that they're sure it works.
 
-Conversely, developers should
-\ruleminor{avoid depending on scripts and tools which are not available as packages}.
-In many cases, a program's author may not realize that some tool was built locally, and
-doesn't exist elsewhere. At present, the only sure way to discover such
-unknown dependencies is to install on a system administered by someone
-else and see what breaks. As use of virtualization containers becomes more
-widespread, software installation can also be tested on a virtual machine or
-container system like Docker.
+Conversely, developers should \ruleminor{avoid depending on scripts
+and tools which are not available as packages}.  In many cases, a
+program's author may not realize that some tool was built locally, and
+doesn't exist elsewhere. At present, the only sure way to discover
+such unknown dependencies is to install on a system administered by
+someone else and see what breaks. As use of virtualization containers
+becomes more widespread, software installation can also be tested on a
+virtual machine or container system like Docker.
 
@@
 
-Root (also known as ``superuser'' or ``admin'') is a special account on
-a computer that has (among other things) the power to modify or delete
-system files and user accounts. Conversely, files and directories owned
-by root usually cannot be modifed by normal users.
+Root (also known as ``superuser'' or ``admin'') is a special account
+on a computer that has (among other things) the power to modify or
+delete system files and user accounts. Conversely, files and
+directories owned by root usually cannot be modifed by normal users.
 
 Installing or running a program with root privileges is often
-convenient, since doing so automatically bypasses all those pesky safety
-checks that might otherwise get in the user's way. However, those checks
-are there for a reason: scientific software packages may not
-intentionally be malware, but one small bug or over-eager file-matching
-expression can certainly make them behave as if they were. Outside of
-very unusual circumstances,
-\ruleminor{packages should not require root privileges to set up or use}.
+convenient, since doing so automatically bypasses all those pesky
+safety checks that might otherwise get in the user's way. However,
+those checks are there for a reason: scientific software packages may
+not intentionally be malware, but one small bug or over-eager
+file-matching expression can certainly make them behave as if they
+were. Outside of very unusual circumstances, \ruleminor{packages
+should not require root privileges to set up or use}.
 
@@
 package before installing it system-wide on a cluster. Requiring root
-privileges will frustrate such efforts, and thereby reduce uptake of the
-package. Requiring, as Apache Tomcat does, that software be installed
-under its own user account---i.e.,
-that \texttt{packagename} be made a user, and all of the
-package's software be installed in that pseudo-user's space---is similarly limiting,
-and makes side-by-side installation of multiple versions of
-the package more difficult.
+privileges will frustrate such efforts, and thereby reduce uptake of
+the package. Requiring, as Apache Tomcat does, that software be
+installed under its own user account---i.e., that \texttt{packagename}
+be made a user, and all of the package's software be installed in that
+pseudo-user's space---is similarly limiting, and makes side-by-side
+installation of multiple versions of the package more difficult.
 
-Developers should therefore
-\ruleminor{allow packages to be installed in an arbitrary location},
-e.g., under a user's home directory in
-\texttt{\textasciitilde{}/packagename}, or in directories with standard
-names like \texttt{bin}, \texttt{lib}, and \texttt{man} under a chosen
-directory. If the first option is chosen, the user may need to modify
-her search path to include the package's executables and libraries, but
-this can (more or less) be automated, and is much less risky than
-setting things up as root.
+Developers should therefore \ruleminor{allow packages to be installed
+in an arbitrary location}, e.g., under a user's home directory in
+\texttt{\textasciitilde{}/packagename}, or in directories with
+standard names like \texttt{bin}, \texttt{lib}, and \texttt{man} under
+a chosen directory. If the first option is chosen, the user may need
+to modify her search path to include the package's executables and
+libraries, but this can (more or less) be automated, and is much less
+risky than setting things up as root.
 
-Testing the ability to install software has traditionally been regarded as difficult,
-since it necessarily alters the machine on which the test is conducted.
-Lightweight virtualization containers like Docker make this much easier as well,
-or simply \ruleminor{ask another person to try and build your software before releasing it}.
+Testing the ability to install software has traditionally been
+regarded as difficult, since it necessarily alters the machine on
+which the test is conducted.  Lightweight virtualization containers
+like Docker make this much easier as well, or simply \ruleminor{ask
+another person to try and build your software before releasing it}.
 
@@
 It's easy to write software that reads input from a file called
-\texttt{mydata.csv}, but also very limiting. If a colleague asks you to
-process her data, you must either overwrite your data file (which is
-risky) or edit your code to read \texttt{otherdata.csv} (which is also
-risky, because there's every likelihood you'll forget to change the
-filename back, or will change three uses of the filename but not a
+\texttt{mydata.csv}, but also very limiting. If a colleague asks you
+to process her data, you must either overwrite your data file (which
+is risky) or edit your code to read \texttt{otherdata.csv} (which is
+also risky, because there's every likelihood you'll forget to change
+the filename back, or will change three uses of the filename but not a
 fourth).
 
-Hard-coding file paths in a program also makes the software harder to run
-in other environments. If your package is installed on a cluster, for
-example, the user's data will almost certainly \emph{not} be in the same
-directory as the software, and the folder
+Hard-coding file paths in a program also makes the software harder to
+run in other environments. If your package is installed on a cluster,
+for example, the user's data will almost certainly \emph{not} be in
+the same directory as the software, and the folder
 \texttt{C:\textbackslash{}users\textbackslash{}yourname\textbackslash{}}
@@
 
-For these reasons, users should be able to
-\ruleminor{set the names and locations of input and output files as command-line parameters}.
+For these reasons, users should be able to \ruleminor{set the names
+and locations of input and output files as command-line parameters}.
 This rule applies to reference data sets as well as the user's own
@@
 a different set of genes as a training set, she should not have to
-edit the software to do so.
-A corollary to this rule is
-\ruleminor{do not require users to navigate to a particular directory to do their work},
-since ``where I have to be'' is just another hard-coded path.
+edit the software to do so.  A corollary to this rule is \ruleminor{do
+not require users to navigate to a particular directory to do their
+work}, since ``where I have to be'' is just another hard-coded path.
 
 In order to save typing, it is often convenient to allow users to
-specify an input or output \emph{directory}, and then require that there
-be files with particular names in that directory. This practice is an
-example of ``convention over configuration'', a principle is used by
-software frameworks such as WordPress and Ruby on Rails that often
-strikes a good balance between adaptability and consistency.
+specify an input or output \emph{directory}, and then require that
+there be files with particular names in that directory. This practice
+is an example of ``convention over configuration'', a principle is
+used by software frameworks such as WordPress and Ruby on Rails that
+often strikes a good balance between adaptability and consistency.
 
@@
 
-Every package should come with a set of tests for users to run
-after installation. Its purpose is not only to check that the software
-is working correctly (although that is extremely helpful), but also to
+Every package should come with a set of tests for users to run after
+installation. Its purpose is not only to check that the software is
+working correctly (although that is extremely helpful), but also to
 ensure that it works at all. This test script can also serve as a
@@
 
-In order to be useful, \ruleminor{make the tests easy to find and run}.
-Many build systems will also run unit tests if provided them at compile time.
-For users, or
-if the build system is not amenable to testing, 
-provide a working script
-in the project's root directory named \texttt{runtests.sh}
-or something equally obvious. 
-This lets new users build their analysis from a working script.
-For example, with its distribution, 
-HISAT2 includes a full set of very small files,
-and a 'Getting Started with HISAT2' section in its manual 
-that leads you through the entire data lifecycle~\cite{pertea2016}.
+In order to be useful, \ruleminor{make the tests easy to find and
+run}.  Many build systems will also run unit tests if provided them
+at compile time.  For users, or if the build system is not amenable to
+testing, provide a working script in the project's root directory
+named \texttt{runtests.sh} or something equally obvious.  This lets
+new users build their analysis from a working script.  For example,
+with its distribution, HISAT2 includes a full set of very small files,
+and a 'Getting Started with HISAT2' section in its manual that leads
+you through the entire data lifecycle~\cite{pertea2016}.
 
-Equally, \ruleminor{make the test script's output easy to interpret}. Screens
-full of correlation coefficients do not qualify: instead, the script's
-output should be simple to understand for non-experts,
-such as one line per test, with the test's name
-and its pass/fail status, followed by a single summary line saying how
-many tests were run and how many passed or failed. If many or all tests
+Equally, \ruleminor{make the test script's output easy to
+interpret}. Screens full of correlation coefficients do not qualify:
+instead, the script's output should be simple to understand for
+non-experts, such as one line per test, with the test's name and its
+pass/fail status, followed by a single summary line saying how many
+tests were run and how many passed or failed. If many or all tests
 fail because of missing dependencies, that fact should be displayed
@@
 Research has shown that the ease with which people can start making
-contributions is a strong predictor of whether they will or not~\cite{steinmacher2015}.
-By making it simpler for outsiders to contribute,
-a test suite of any kind also makes it more likely that they will, and software
-with collaborators stands a better chance of surviving in the busy field of
-scientific software.
+contributions is a strong predictor of whether they will or
+not~\cite{steinmacher2015}.  By making it simpler for outsiders to
+contribute, a test suite of any kind also makes it more likely that
+they will, and software with collaborators stands a better chance of
+surviving in the busy field of scientific software.
 
@@
 
-The usage message tells users what the program could do.
-It is equally important for the program to tell users what it actually did.
-Accordingly,
-when the program starts, it should \ruleminor{echo all parameters and software
-versions to standard out or a log file alongside the results} to increase
-the reproducibility of that step.
+The usage message tells users what the program could do.  It is
+equally important for the program to tell users what it actually did.
+Accordingly, when the program starts, it should \ruleminor{echo all
+parameters and software versions to standard out or a log file
+alongside the results} to increase the reproducibility of that step.
 
-Given a set of parameters and a dataset, \ruleminor{a particular version of a program
-should produce the same results every time it is run}
-to aid testing, debugging, and reproducibility.
-Even minor changes to code can cause minor changes in output because of floating-point issues,
-which means that getting exactly the same output for the same input and parameters
-probably won't work during development,
-but it should still be a goal for people who have deployed a specific version.
+Given a set of parameters and a dataset, \ruleminor{a particular
+version of a program should produce the same results every time it
+is run} to aid testing, debugging, and reproducibility.  Even minor
+changes to code can cause minor changes in output because of
+floating-point issues, which means that getting exactly the same
+output for the same input and parameters probably won't work during
+development, but it should still be a goal for people who have
+deployed a specific version.
 
-Many applications rely on randomized algorithms to
-improve performance or runtimes. As a consequence, results can change
-between runs, even when provided with the same data and parameters. By
-its nature, this randomness renders strict reproducibility and therefore
-debugging more difficult. If even the small test set (\#9) produces
-different results for each run, new users may not be able to tell whether the software is
-working properly. When comparing
-results between versions or after changing parameters, even small
-differences can confuse or muddy the comparison. And especially when
-producing results for publications, grants or diagnoses, any analysis
-should be absolutely reproducible.
+Many applications rely on randomized algorithms to improve performance
+or runtimes. As a consequence, results can change between runs, even
+when provided with the same data and parameters. By its nature, this
+randomness renders strict reproducibility and therefore debugging more
+difficult. If even the small test set (\#9) produces different results
+for each run, new users may not be able to tell whether the software
+is working properly. When comparing results between versions or after
+changing parameters, even small differences can confuse or muddy the
+comparison. And especially when producing results for publications,
+grants or diagnoses, any analysis should be absolutely reproducible.
 
 Given the size of biological data, it is unreasonable to suggest that
-random algorithms be removed. However, most programs use a pseudo-random
-number generator, which uses a starting seed and an equation to
-approximate random numbers. Setting the seed to a consistent value
-can remove randomness between runs. \ruleminor{Allow the user to optionally provide
-the random seed as an input parameter}, thus rendering the program deterministic
-for those cases where it matters. If the seed is set internally (e.g.,
-using clock time), echo it to the output for re-use later.
-If setting the seed is not possible, \ruleminor{make sure the acceptable tolerance is
-known and detailed in documentation and in the tests}.
+random algorithms be removed. However, most programs use a
+pseudo-random number generator, which uses a starting seed and an
+equation to approximate random numbers. Setting the seed to a
+consistent value can remove randomness between runs. \ruleminor{Allow
+the user to optionally provide the random seed as an input
+parameter}, thus rendering the program deterministic for those cases
+where it matters. If the seed is set internally (e.g., using clock
+time), echo it to the output for re-use later.  If setting the seed is
+not possible, \ruleminor{make sure the acceptable tolerance is known
+and detailed in documentation and in the tests}.
 

### 2017-02-24 robust-software.tex
Made fixups according to PLOS guidelines

@@
 \newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}
-\newcommand{\rulemajor}[1]{\section{#1}}
+\newcommand{\rulemajor}[1]{\section*{#1}}
 \newcommand{\ruleminor}[1]{\textbf{#1}}
@@
 If you are new to version control, it is simplest to treat it as ``a
-better \withurl{Dropbox}{http://dropbox.com}'' (or, if you are of a
+better Dropbox'' (or, if you are of a
 certain age, a better FTP), and to use it simply to synchronize files
@@
 
-\withurl{Semantic versioning}{http://semver.org/} is one of the most
+Semantic versioning~\cite{semver} is one of the most
 common types of versioning for open-source software. Version numbers
@@
 
+\section*{Supporting information}
+
+\emph{S1 Checklist. Robust software checklist.} A checklist summarizing these ten simple rules to apply to your own
+software.
+
 \begin{thebibliography}{10}

### 2017-02-24 Robust-Paper_PLOS_CopyEdit.tex
PLOS copyediting fixups. Added the second tex version with copyedit title and bibliography in main document

--- 2017-02-24 robust-software.tex
@@
 
-% enumerate package lets us use letters instead of numbers
-\usepackage{enumerate}
-
 % create "+" rule type for thick vertical lines
@@
 \noalign{\global\arrayrulewidth\savedwidth}}
-
-\usepackage{color}
 
@@
 \newcommand{\ruleminor}[1]{\textbf{#1}}
+%% END MACROS SECTION
+
+% enumerate package lets us use letters instead of numbers
+\usepackage{enumerate}
+\usepackage{color}
+
 \begin{document}
@@
 \\
-\textbf{1} Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
+\bigskip
+\textbf{1} Genome Sequence Informatics, Ontario Institute for Cancer Research, Toronto, Ontario, Canada
 \\
-\textbf{2} Software Carpentry Foundation / gvwilson@software-carpentry.org
+\textbf{2} Software Carpentry Foundation
 \\
@@
 \\
-* Corresponding author.
+* morgan.taschuk@oicr.on.ca
 \end{flushleft}
@@
 keep track of changes.  They should therefore \ruleminor{put
-everything created manually into version control as soon as it is
+everything into version control as soon as it is
 created}, including programs, original field observations, and the
@@
 
-\emph{S1 Checklist. Robust software checklist.} A checklist summarizing these ten simple rules to apply to your own
-software.
+\paragraph*{S1 Checklist.}
+\label{S1_Checklist}
+{\bf Robust software checklist.} A checklist summarizing these ten simple rules
+to apply to your own software.
 
@@
 
+\bibitem{semver}
+Preston-Werner T. Semantic Versioning 2.0.0; 2016.
+\newblock \url{http://semver.org/spec/v2.0.0.html}.
+
 \bibitem{pertea2016}

### 2017-02-24 robust-checks.tex
PLOS copyediting fixups. Added the second tex version with copyedit title and bibliography in main document

@@
 \documentclass[10pt,letterpaper]{article}
-\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
-
-% amsmath and amssymb packages, useful for mathematical formulas and symbols
-\usepackage{amsmath,amssymb}
-
-% Use adjustwidth environment to exceed column width (see example table in text)
-\usepackage{changepage}
+\usepackage[top=0.85in,left=1.0in,footskip=0.75in]{geometry}
 
@@
 
-% textcomp package and marvosym package for additional characters
-\usepackage{textcomp,marvosym}
-
-% cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
-
-% Use nameref to cite supporting information files (see Supporting Information section for more info)
-\usepackage{nameref,hyperref}
-
-% line numbers
-\usepackage[right]{lineno}
-
-% ligatures disabled
-\usepackage{microtype}
-\DisableLigatures[f]{encoding = *, family = * }
-
-% color can be used to apply background shading to table cells only
-\usepackage[table]{xcolor}
-
-% array package and thick rules for tables
-\usepackage{array}
-
-% enumerate package lets us use letters instead of numbers
-\usepackage{enumerate}
-
-% create "+" rule type for thick vertical lines
-\newcolumntype{+}{!{\vrule width 2pt}}
-
-% create \thickcline for thick horizontal lines of variable length
-\newlength\savedwidth
-\newcommand\thickcline[1]{%
-  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-  \cline{#1}%
-  \noalign{\vskip\arrayrulewidth}%
-  \noalign{\global\arrayrulewidth\savedwidth}%
-}
-
-% \thickhline command for thick horizontal lines that span the table
-\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
-\hline
-\noalign{\global\arrayrulewidth\savedwidth}}
-
 % Remove comment for double spacing
-%\usepackage{setspace}
-%\doublespacing
-
-% Text layout
-\raggedright
-\setlength{\parindent}{0.5cm}
-\textwidth 5.25in
-\textheight 8.75in
-
-% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
-% Captions will be left justified
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
-\renewcommand{\figurename}{Fig}
-
-% Use the PLoS provided BiBTeX style
-\bibliographystyle{plos2015}
-
-% Remove brackets from numbering in List of References
-\makeatletter
-\renewcommand{\@biblabel}[1]{\quad#1.}
-\makeatother
-
-% Leave date blank
-\date{}
-
-% Header and Footer with logo
-\usepackage{lastpage,fancyhdr,graphicx}
-\usepackage{epstopdf}
-\pagestyle{myheadings}
-\pagestyle{fancy}
-\fancyhf{}
-\setlength{\headheight}{27.023pt}
-\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
-\rfoot{\thepage/\pageref{LastPage}}
-\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
-\fancyheadoffset[L]{2.25in}
-\fancyfootoffset[L]{2.25in}
-\lfoot{\sf PLOS}
-
-%% Include all macros below
-\newcommand{\fixme}[2]{\textsc{\textbf{{#1}: {#2}}}}
-\newcommand{\recommend}[1]{\textit{#1}}
-\newcommand{\withurl}[2]{{#1}\footnote{\texttt{#2}}}
+\usepackage{setspace}
+\onehalfspacing
 
@@
 \begin{document}
-\vspace*{0.2in}
 
@@
 }
-\newline
-\\
-{Morgan~Taschuk}\textsuperscript{1,\ddag *},
-{Greg~Wilson}\textsuperscript{2,\ddag}
-\\
-\textbf{1} Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
-\\
-\textbf{2} Software Carpentry Foundation / gvwilson@software-carpentry.org
-\\
-\bigskip
-{\ddag} These authors contributed equally to this work.
-\\
-* Corresponding author.
 \end{flushleft}
@@
 & Use version control.
-&& Put everything created manually into version control as soon as it is created.
+&& Put everything into version control as soon as it is created.
+&& Use a feature branch workflow.
 
@@
 & Include a small test set that can be run to ensure the software is actually working.
-&& Make the test script easy to find and run.
+&& Make the tests easy to find and run.
 && Make the test script's output easy to interpret.

### 2017-03-03 robust-software.tex
Fixups from PR; added Software Carpentry URL to affiliation

@@
 \\
-\textbf{1} Ontario Institute for Cancer Research / morgan.taschuk@oicr.on.ca
+\bigskip
+\textbf{1} Genome Sequence Informatics, Ontario Institute for Cancer Research, Toronto, Ontario, Canada
 \\
-\textbf{2} Software Carpentry Foundation / gvwilson@software-carpentry.org
+\textbf{2} Software Carpentry Foundation, https://software-carpentry.org/
 \\
@@
 
-\emph{S1 Checklist. Robust software checklist.} A checklist summarizing these ten simple rules to apply to your own
-software.
+\paragraph*{S1 Checklist.}
+\label{S1_Checklist}
+{\bf Robust software checklist.} A checklist summarizing these ten simple rules
+to apply to your own software.
 
-\begin{thebibliography}{10}
-
-\bibitem{prins2015}
-Prins P, de~Ligt J, Tarasov A, Jansen RC, Cuppen E, Bourne PE.
-\newblock Toward effective software solutions for big biology.
-\newblock Nature Biotechnology. 2015;33(7):686--687.
-\newblock doi:{10.1038/nbt.3240}.
-
-\bibitem{baker2016}
-Baker M.
-\newblock 1,500 scientists lift the lid on reproducibility.
-\newblock Nature. 2016;533(7604):452--454.
-\newblock doi:{10.1038/533452a}.
-
-\bibitem{prabhu2011}
-Prabhu P, Jablin TB, Raman A, Zhang Y, Huang J, Kim H, et~al.
-\newblock A Survey of the Practice of Computational Science.
-\newblock In: State of the Practice Reports. SC '11. New York, NY, USA: ACM;
-  2011. p. 19:1--19:12.
-
-\bibitem{lawlor2015}
-Lawlor B, Walsh P.
-\newblock Engineering bioinformatics: building reliability, performance and
-  productivity into bioinformatics software.
-\newblock Bioengineered. 2015;6(4).
-\newblock doi:{10.1080/21655979.2015.1050162}.
-
-\bibitem{ison2016}
-Ison J, Rapacki K, Ménager H, Kalaš M, Rydza E, Chmura P, et~al.
-\newblock Tools and data services registry: a community effort to document
-  bioinformatics resources.
-\newblock Nucleic Acids Research. 2016;44(D1):D38--D47.
-\newblock doi:{10.1093/nar/gkv1116}.
-
-\bibitem{brazas2012}
-Brazas MD, Yim D, Yeung W, Ouellette BFF.
-\newblock A decade of web server updates at the bioinformatics links directory:
-  2003-–2012.
-\newblock Nucleic Acids Research. 2012;doi:{10.1093/nar/gks632}.
-
-\bibitem{stajich2002}
-Stajich JE, Block D, Boulez K, Brenner SE, Chervitz SA, Dagdigian C, et~al.
-\newblock The Bioperl Toolkit: Perl Modules for the Life Sciences.
-\newblock Genome Research. 2002;12(10):1611--1618.
-\newblock doi:{10.1101/gr.361602}.
-
-\bibitem{Seemann2013}
-Seemann T.
-\newblock Ten recommendations for creating usable bioinformatics command line
-  software.
-\newblock GigaScience. 2013;2(1):15.
-\newblock doi:{10.1186/2047-217X-2-15}.
-
-\bibitem{nekrutenko2012}
-Nekrutenko A, Taylor J.
-\newblock Next-generation sequencing data interpretation: enhancing
-  reproducibility and accessibility.
-\newblock Nature ReviewsGenetics. 2012;13(9):667--72.
-\newblock doi:{10.1038/nrg3305}.
-
-\bibitem{atwood2015}
-Atwood TK, Bongcam-Rudloff E, Brazas ME, Corpas M, Gaudet P, Lewitter F, et~al.
-\newblock GOBLET: The Global Organisation for Bioinformatics Learning,
-  Education and Training.
-\newblock PLoS Computational Biology. 2015;doi:{10.1371/journal.pcbi.1004143}.
-
-\bibitem{sanders2008}
-Sanders R, Kelly D.
-\newblock Dealing with Risk in Scientific Software Development.
-\newblock Software, IEEE. 2008;25(4):21--28.
-\newblock doi:{10.1109/ms.2008.84}.
-
-\bibitem{afgan2016}
-Afgan E, Baker D, van~den Beek M, Blankenberg D, Bouvier D, Čech M, et~al.
-\newblock The Galaxy platform for accessible, reproducible and collaborative
-  biomedical analyses: 2016 update.
-\newblock Nucleic Acids Research. 2016;44(W1):W3--W10.
-\newblock doi:{10.1093/nar/gkw343}.
-
-\bibitem{howe2012}
-Howe B.
-\newblock Virtual Appliances, Cloud Computing, and Reproducible Research.
-\newblock Computing in Science Engineering. 2012;14(4):36--41.
-\newblock doi:{10.1109/MCSE.2012.62}.
-
-\bibitem{gentleman2004}
-Gentleman RC, Carey VJ, Bates DM, Bolstad B, Dettling M, Dudoit S, et~al.
-\newblock Bioconductor: open software development for computational biology and
-  bioinformatics.
-\newblock Genome Biology. 2004;5(10):1--16.
-\newblock doi:{10.1186/gb-2004-5-10-r80}.
-
-\bibitem{varoquaux2015}
-Varoquaux G. Software for reproducible science: let’s not have a
-  misunderstanding; 2015.
-\newblock
-  \url{http://gael-varoquaux.info/programming/software-for-reproducible-science-lets-not-have-a-misunderstanding.html}.
-
-\bibitem{brown2013}
-Brown CT. Replication, reproduction, and remixing in research software; 2013.
-\newblock \url{http://ivory.idyll.org/blog/research-software-reuse.html}.
-
-\bibitem{wilson2014}
-Wilson G, Aruliah DA, Brown CT, Hong NPC, Davis M, Guy RT, et~al.
-\newblock Best Practices for Scientific Computing.
-\newblock PLoS Biology. 2014;12(1):e1001745.
-\newblock doi:{10.1371/journal.pbio.1001745}.
-
-\bibitem{wilson2016}
-Wilson G, Bryan J, Cranston K, Kitzes J, Nederbragt L, Teal TK.
-\newblock Good Enough Practices in Scientific Computing.
-\newblock arxivorg. 2016;abs/1609.00037.
-
-\bibitem{noble2009}
-Noble WS.
-\newblock {A Quick Guide to Organizing Computational Biology Projects}.
-\newblock PLoS Computational Biology. 2009;5(7).
-\newblock doi:{10.1371/journal.pcbi.1000424}.
-
-\bibitem{blischak2016}
-Blischak JD, Davenport ER, Wilson G.
-\newblock A Quick Introduction to Version Control with Git and GitHub.
-\newblock PLOS Computational Biology. 2016;12(1):1--18.
-\newblock doi:{10.1371/journal.pcbi.1004668}.
-
-\bibitem{karimzadeh2016}
-Karimzadeh M, Hoffman MM. Creating great documentation for bioinformatics
-  software; 2016.
-\newblock \url{http://hdl.handle.net/1807/73111}.
-
-\bibitem{Johnson1997}
-Johnson M.
-\newblock Building a Better ReadMe.
-\newblock Technical Communication. 1997;44(1):28--36.
-
-\bibitem{gnustandards}
-{\relax Free Software Foundation}. GNU Coding Standards; 2016.
-\newblock \url{https://www.gnu.org/prep/standards/standards.html}.
-
-\bibitem{posix2016}
-{\relax The IEEE}, {\relax The Open Group}. The Open Group Base Specifications
-  Issue 7 IEEE Std 1003.1-2008. 12. Utility Conventions; 2016.
-\newblock
-  \url{http://pubs.opengroup.org/onlinepubs/9699919799/basedefs/V1_chap12.html}.
-
-\bibitem{pertea2016}
-Pertea M, Kim D, Pertea GM, Leek JT, Salzberg SL.
-\newblock Transcript-level expression analysis of RNA-seq experiments with
-  HISAT, StringTie and Ballgown.
-\newblock Nature Protocols. 2016;11(9):1650--1667.
-
-\bibitem{steinmacher2015}
-Steinmacher I, Silva MAG, Gerosa MA, Redmiles DF.
-\newblock A systematic literature review on the barriers faced by newcomers to
-  open source software projects.
-\newblock Information and Software Technology. 2015;59:67 -- 85.
-\newblock doi:{http://dx.doi.org/10.1016/j.infsof.2014.11.001}.
-
-\end{thebibliography}
+\bibliography{robust-software}
 
